### 2018-08-24 doc/paper.tex
add a paper stub

\documentclass{article}
\title{A Concurrent Semantic for Shell-Like Processes}
\author{Jeanine Adkisson, Hidehiko Masuhara}
\usepackage{comment}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphics,graphicx}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{auto-pst-pdf}

\begin{document}
\maketitle

\end{document}

### 2018-08-24 lib/magritte/channel.rb
use Blocker structures instead of thread variables

@@ -1,4 +1,34 @@
 module Magritte
+  class Blocker
+    attr_reader :thread, :val
+
+    def interrupt!
+      @thread[:magritte_proc].interrupt!
+    end
+
+    def wakeup
+      raise "#{self.inspect} still running!" unless @thread.stop?
+
+      @thread.run
+    end
+  end
+
+  class Sender < Blocker
+    def initialize(thread, val)
+      @thread, @val = thread, val
+    end
+  end
+
+  class Receiver < Blocker
+    def initialize(thread)
+      @thread = thread
+    end
+
+    def <<(val)
+      @val = val
+    end
+  end
+
   class LogMutex
     def initialize(name)
       @name = name
@@ -67,7 +97,7 @@ module Magritte
 
     def remove_reader(p)
       @mutex.synchronize do
-        @block_set.delete(p.thread) if @block_type == :read
+        unregister_thread(p.thread) if @block_type == :read
 
         cleanup_from(p, @readers)
       end.call
@@ -75,7 +105,7 @@ module Magritte
 
     def remove_writer(p)
       @mutex.synchronize do
-        @block_set.delete(p.thread) if @block_type == :write
+        unregister_thread(p.thread) if @block_type == :write
 
         cleanup_from(p, @writers)
       end.call
@@ -135,6 +165,10 @@ module Magritte
 
     private
 
+    def unregister_thread(t)
+      @block_set.reject! { |b| b.thread == t }
+    end
+
     def cleanup_from(p, set)
       should_close = open? && (set.delete(p); set.empty?)
 
@@ -142,7 +176,7 @@ module Magritte
         @open = false
         to_clean = @block_set.dup
         @block_set.clear
-        proc { to_clean.each { |t| t[:magritte_proc].interrupt! } }
+        proc { to_clean.each(&:interrupt!) }
       else
         proc { } # nop
       end
@@ -165,8 +199,7 @@ module Magritte
     end
 
     def block_write(val)
-      Thread.current[:__magritte_write] = val
-      @block_set << Thread.current
+      @block_set << Sender.new(Thread.current, val)
       @mutex.sleep
 
       # action for after unlocking mutexes
@@ -174,21 +207,20 @@ module Magritte
     end
 
     def block_read
-      @block_set << Thread.current
+      receiver = Receiver.new(Thread.current)
+      @block_set << receiver
       @mutex.sleep
 
       # action for after unlocking mutexes
-      proc { Thread.current[:__magritte_read] }
+      proc { receiver.val }
[…]

### 2018-08-24 notes/simple.rb
remove the use of thread variables in simple.rb

@@ -1,14 +1,22 @@
 require 'thread'
 require 'set'
 
+$exception = StandardError.new
+
 # $... means global variable
 
 $mutex = Mutex.new
 $readers = Set.new
 $writers = Set.new
 $block_type = :none
-$block_set = []
 $open = true
+
+# an element of the block set will look like:
+# [thread, value]
+# where value is either the value to be written (if writers are blocked)
+# or a ref to receive the value (if readers are blocked)
+$block_set = []
+
 
 $output_mutex = Mutex.new
 $output = []
@@ -24,25 +32,25 @@ end
 def rm_reader(t)
   to_close = $mutex.synchronize do
     if $block_type == :read
-      $block_set.delete(t)
+      $block_set.reject! { |b| b[0] == t }
     end
 
     cleanup_from(t, $readers)
   end
 
-  to_close.each { |thread| thread.raise(Interrupt.new) }
+  to_close.each { |thread| thread.raise($exception) }
 end
 
 def rm_writer(t)
   to_close = $mutex.synchronize do
     if $block_type == :write
-      $block_set.delete(t)
+      $block_set.reject! { |b| b[0] == t }
     end
 
     cleanup_from(t, $writers)
   end
 
-  to_close.each { |thread| thread.raise(Interrupt.new) }
+  to_close.each { |thread| thread.raise($exception) }
 end
 
 def cleanup_from(t, set)
@@ -59,7 +67,7 @@ def cleanup_from(t, set)
     $open = false
     to_clean = $block_set.dup
     $block_set.clear
-    return to_clean
+    return to_clean.map { |b| b[0] }
   else
     return Set.new
   end
@@ -68,7 +76,7 @@ end
 def read(t)
   $mutex.synchronize do
     if not $open
-      t.raise(Interrupt.new)
+      t.raise($exception)
     end
 
     case $block_type
@@ -76,60 +84,57 @@ def read(t)
       # in this case there are no waiting write threads,
       # so this process must block.
       $block_type = :read
-      $block_set << t
+      ref = [t, nil]
+      $block_set << ref
 
       # this releases the mutex and sleeps in one atomic action.
       # on wakeup, it will acquire the mutex again (and immediately
       # release it at the end of the synchronize block).
       $mutex.sleep
+
+      return ref[1]
     when :write
       # in this case there are waiting write threads,
       # so we do *not* block, and instead wake up a
       # write thread.
-      write_thread = $block_set.shift
[…]

### 2018-08-24 spec/code_spec.rb
dynamically retry waiting for dangling threads to finish

@@ -3,14 +3,23 @@ describe Magritte::Code do
     orig_threads = Thread.list
     out = yield
 
-    # yield the current thread to allow other threads to be cleaned up
-    # since sometimes it takes a bit of time for thread.kill to actually
-    # kill the thread and there's no way to wait for it :\
-    sleep 0.1
+    begin
+      dangling = Thread.list - orig_threads
 
-    dangling = Thread.list - orig_threads
-    assert { dangling.empty? }
-    out
+      assert { dangling.empty? }
+      out
+    rescue Minitest::Assertion
+      retry_count ||= 0
+      retry_count += 1
+      raise if retry_count > 20
+
+      # yield the current thread to allow other threads to be cleaned up
+      # since sometimes it takes a bit of time for thread.raise to actually
+      # kill the thread and there's no way to wait for it :\
+      sleep 0.1
+
+      retry
+    end
   end
 
   let(:output) { with_no_dangling_threads { code.spawn_collect } }

### 2018-09-02 doc/paper.tex
add an abstract and attempt acm formatting

@@
-\documentclass{article}
-\title{A Concurrent Semantic for Shell-Like Processes}
-\author{Jeanine Adkisson, Hidehiko Masuhara}
+\documentclass[format=sigconf, review=true, screen=true]{acmart}
+
+% don't print the conference ref for now
+\settopmatter{printacmref=false}
+
 \usepackage{comment}
@@
 
+\title{A Concurrent Semantic for Shell-Like Processes}
+\author{Jeanine Adkisson}
+\affiliation{Tokyo Institute of Technology}
+\email{jneen@prg.is.titech.edu}
+
+\author{Hidehiko Masuhara}
+\affiliation{Tokyo Institute of Technology}
+\email{masuhara@prg.is.titech.edu}
+
 \begin{document}
+
+\begin{abstract}
+Despite the widespread use of GUI tools for configuration tasks, CLI tools, and more specifically shell-based interfaces remain a powerful standard for software distribution and integration.
+Several attempts have been made at solving the desktop-integration problem, most notably TCL(TODO: cite!) and Guile Scheme(TODO: cite!), but neither has been as succesful as they intended to be: the \emph{lingua franca} for a large portion of tools remains shell-oriented, relying on such concepts as standard I/O, string-based argument vectors, and the expectation that users will be able to freely pipe streams of bytes in a shell-like way.
+
+Unfortunately, shell languages have significant known interoperability issues, such as lack of any kind of usable encapsulation, and the lack of in-language data structures to represent any kind of value other than a string.
+We believe that a shell-compatible language with well-defined semantics that could also function as a general-purpose systems language could go a long way towards solving the desktop-integration problem.
+
+In this paper, we explore a potential concurrency model for such a language, inspired by such systems as Go, Clojure, and Erlang.
+A persistent problem with these battle-tested systems is that they don't account for closing channels, and require manual cleanup of processes.
+In a systems programming context, errors are commonplace and expected.
+We propose a channel-based communication system that integrates with a workflow compensation system and is capable of recovering from errors and, importantly, cleaning up processes when communication halts.
+\end{abstract}
 \maketitle

### 2018-09-02 notes/simple.rb
add random sleeps in simple.rb

@@ -159,6 +159,8 @@ def spawn_consumer!
     begin
       add_reader(Thread.current)
       10.times do
+        # yield the thread randomly to increase the chance of races
+        sleep(rand / 10)
         $output_mutex.synchronize { $output << read(Thread.current) }
       end
     ensure

### 2018-10-09 lib/magritte/ast.rb
Implement free variable scanner and factor out tree

@@ -1,223 +1,47 @@
 module Magritte
   module AST
-    class Attr
-      attr_reader :value
-      def initialize(value)
-        @value = value
-      end
 
-      def each(&block)
-        raise "abstract"
-      end
-
-      def map(&block)
-        raise "abstract"
-      end
-
-      # Object equality based on 
-      # https://stackoverflow.com/questions/1931604/whats-the-right-way-to-implement-equality-in-ruby
-      def ==(o)
-        o.class == self.class && o.value == @value
-      end
-
-      alias eql? ==
-
-      def hash
-        [self.class, self.value].hash
-      end
-    end
-
-    class DataAttr < Attr
-      def each(&block)
-        #pass
-      end
-
-      def map(&block)
-        self
-      end
-
-      def inspect
-        value.inspect
-      end
-    end
-
-    class RecAttr < Attr
-      def each(&block)
-        block.call(@value)
-      end
-
-      def map(&block)
-        RecAttr.new(block.call(@value))
-      end
-
-      def inspect
-        "*#{value.inspect}"
-      end
-    end
-
-    class ListRecAttr < Attr
-      def each(&block)
-        @value.each(&block)
-      end
-
-      def map(&block)
-        ListRecAttr.new(@value.map(&block))
-      end
-
-      def inspect
-        "**#{value.inspect}"
-      end
-    end
-
-    class Node
-      class << self
-        def attrs
-          @attrs ||= []
-        end
-
-        def types
-          @types ||= []
-        end
-
-        def short_name
-          @short_name ||= begin
-            self.class.name =~ /(?:.*::)?(.*)/
-            $1.gsub(/([[:lower:]])([[:upper:]])/) { "#{$1}_#{$2.downcase}" }
-              .downcase
-          end
-        end
-
-        def defattr(name, type)
-          attrs << name
-          types << type
-          index = attrs.size - 1
-          define_method name do
-            attrs[index].value
-          end
-        end
-
-        def defdata(name)
-          defattr(name, DataAttr)
-        end
-
-        def defrec(name)
-          defattr(name, RecAttr)
-        end
-
-        def deflistrec(name)
-          defattr(name, ListRecAttr)
-        end
-
-        def make(*attrs)
-          if attrs.size != types.size
-            raise ArgumentError.new("Expected #{types.size} arguments, got #{attrs.size}")
-          end
-
-          new(attrs.zip(types).map { |(value, type)|
[…]

### 2018-10-09 lib/magritte/free_vars.rb

@@ -1,0 +1,22 @@
+module Magritte
+  module FreeVars
+    def self.scan(node)
+      Scanner.new.collect(node)
+    end
+
+    class Scanner < Tree::Collector
+      def visit_lex_variable(node)
+        Set.new([node.name])
+      end
+
+      def visit_lambda(node)
+        unless node.patterns.all? { |n| n.is_a? Binder }
+          raise "Not implemented yet!"
+        end
+        used_vars = collect_from(node.bodies)
+        bound_vars = node.patterns.map(&:name)
+        used_vars - bound_vars
+      end
+    end
+  end
+end

### 2018-10-09 lib/magritte/tree.rb

@@ -1,0 +1,207 @@
+module Magritte
+  module Tree
+    class Attr
+      attr_reader :value
+      def initialize(value)
+        @value = value
+      end
+
+      def each(&block)
+        raise "abstract"
+      end
+
+      def map(&block)
+        raise "abstract"
+      end
+
+      # Object equality based on 
+      # https://stackoverflow.com/questions/1931604/whats-the-right-way-to-implement-equality-in-ruby
+      def ==(o)
+        o.class == self.class && o.value == @value
+      end
+
+      alias eql? ==
+
+      def hash
+        [self.class, self.value].hash
+      end
+    end
+
+    class DataAttr < Attr
+      def each(&block)
+        #pass
+      end
+
+      def map(&block)
+        self
+      end
+
+      def inspect
+        value.inspect
+      end
+    end
+
+    class RecAttr < Attr
+      def each(&block)
+        block.call(@value)
+      end
+
+      def map(&block)
+        RecAttr.new(block.call(@value))
+      end
+
+      def inspect
+        "*#{value.inspect}"
+      end
+    end
+
+    class ListRecAttr < Attr
+      def each(&block)
+        @value.each(&block)
+      end
+
+      def map(&block)
+        ListRecAttr.new(@value.map(&block))
+      end
+
+      def inspect
+        "**#{value.inspect}"
+      end
+    end
+
+    class Node
+      class << self
+        def attrs
+          @attrs ||= []
+        end
+
+        def types
+          @types ||= []
+        end
+
+        def short_name
+          @short_name ||= begin
+            self.name =~ /(?:.*::)?(.*)/
+            $1.gsub(/([[:lower:]])([[:upper:]])/) { "#{$1}_#{$2.downcase}" }
+              .downcase
+          end
+        end
+
+        def defattr(name, type)
+          attrs << name
+          types << type
+          index = attrs.size - 1
+          define_method name do
+            attrs[index].value
+          end
+        end
+
+        def defdata(name)
+          defattr(name, DataAttr)
+        end
+
+        def defrec(name)
+          defattr(name, RecAttr)
+        end
+
+        def deflistrec(name)
+          defattr(name, ListRecAttr)
+        end
+
+        def make(*attrs)
+          if attrs.size != types.size
+            raise ArgumentError.new("Expected #{types.size} arguments, got #{attrs.size}")
+          end
+
+          new(attrs.zip(types).map { |(value, type)|
+            type.new(value)
+          })
[…]

### 2018-10-09 spec/ast_spec.rb

@@ -83,17 +83,16 @@ describe Magritte::AST do
       assert { command.args == ["arg1", "arg2", "arg3"] }
     end
 
-    #describe "maps" do
-    #  let(:mapped) { command.map {  }}
-    #
-    #  it "maps" do
-    #    assert { mapped.is_a? Magritte::AST::Command }
-    #    assert { mapped.head.is_a? Magritte::AST::RecAttr }
-    #    assert { mapped.head.expr.name == "bar" }
-    #    assert { mapped.args.is_a? Magritte::AST::ListRecAttr }
-    #    assert { mapped.args.map { |arg| arg == "bar" }.all? }
-    #  end
-    #end
+    describe "maps" do
+      let(:mapped) { command.map { |val| (val.is_a? Magritte::AST::Variable) ? val.name : val }}
+
+      it "maps" do
+        assert { mapped.is_a? Magritte::AST::Command }
+        assert { mapped.head.is_a? String}
+        assert { mapped.head == "foo" }
+        assert { mapped.args == ["arg1", "arg2", "arg3"]}
+      end
+    end
 
     it "equals itself" do
       assert { command == command }
@@ -117,17 +116,15 @@ describe Magritte::AST do
       assert { block.lines == [command,command] }
     end
 
-    #describe "maps" do
-    #  let(:mapped) { command.map {  }}
-    #
-    #  it "maps" do
-    #    assert { mapped.is_a? Magritte::AST::Command }
-    #    assert { mapped.head.is_a? Magritte::AST::RecAttr }
-    #    assert { mapped.head.expr.name == "bar" }
-    #    assert { mapped.args.is_a? Magritte::AST::ListRecAttr }
-    #    assert { mapped.args.map { |arg| arg == "bar" }.all? }
-    #  end
-    #end
+    describe "maps" do
+      let(:mapped) { block.map { |val| val.head.name }}
+
+      it "maps" do
+        assert { mapped.is_a? Magritte::AST::Block }
+        assert { mapped.lines.map { |elem| elem.is_a? String }.all? }
+        assert { mapped.lines.map { |elem| elem == "foo" }.all? }
+      end
+    end
 
     it "equals itself" do
       assert { block == block }

### 2018-10-10 doc/paper.tex
paper updates

@@
+% notes
+
 \documentclass[format=sigconf, review=true, screen=true]{acmart}
@@
 \usepackage{auto-pst-pdf}
+
+\usepackage{listings}
 
@@
 
-In this paper, we explore a potential concurrency model for such a language, inspired by such systems as Go, Clojure, and Erlang.
-A persistent problem with these battle-tested systems is that they don't account for closing channels, and require manual cleanup of processes.
-In a systems programming context, errors are commonplace and expected.
+On the other hand, a persistent problem with more established systems like Erlang or Go is that they don't account for closing channels, and require manual cleanup of processes.
+In a systems programming context, errors and hangups are commonplace and expected, and it is critical to have some measure of automatic process cleanup for even naively-written code.
+
 We propose a channel-based communication system that integrates with a workflow compensation system and is capable of recovering from errors and, importantly, cleaning up processes when communication halts.
@@
 
+\section{Motivation}
+
+Consider the following pseudo-code for a shell-like language:
+
+\begin{verbatim}
+  read-lines tmp/large-file
+    | take 10
+    | each (?line => do-expensive-work $line)
+\end{verbatim}
+In this example, three processes are spawned: (1) a process with a file open that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
+A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.
+
+In a pipeline-based programming system such as a shell, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
+A user will also expect that, after the 10 lines are processed, the file is closed, and that all three processes exit.
+This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.
+In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
+Similarly, the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.
+
+PEP 533 Authors(TODO: cite properly) describe an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
+But channel-based systems generally allow for multiple concurrent readers and writers - so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
+We propose a system such that processes are interrupted as soon as it is no longer possible for them to resume.
+
+\section{Channel Closing / Interruption}
+
+When a process is blocked on a channel, it will be woken up as soon as another process communicates on the other end.
+Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations.
+In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.
+
+We extend the notion of a synchronous channel with \emph{registered readers} and \emph{registered writers}, and add four intrinsic functions: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, each receiving a channel and a process handle.
+These store and delete processes from the channel's internal sets of registered readers and registered writers.
+Additionally, \verb/remove_reader/ and \verb/remove_writer/ will cause the channel to close if their corresponding sets are empty after the operation.
+
+When a channel is closed, we first interrupt every process that is blocked on the channel.
+Thereafter, any attempted read or write operations will cause the initiating process to be interrupted.
+When a channel is interrupted, we must not only perform any cleanup actions such as closing files or writing to error logs, but also call \verb/remove_reader/ and \verb/remove_writer/ on any channels the process has registered - thereby potentially cascading the interruptions to other processes.
+
+\section{Magritte}
+
+We have implemented \verb/magritte/(TODO: cite), a Ruby DSL that allows communication between native Ruby threads, and automatically cleans up dangling threads with the above semantics.
+The API revolves around a \verb/Magritte::Code/ object, which can also be created with the builtin \verb/s/ primitive.
+
+Within a code object, the primitive functions \verb/put/ and \verb/get/ can be used to read from the input and write to the output.
+A code object can be run in-process (\verb/.call/), in the background (\verb/.go/).
+The default output is a special kind of write-only channel called a \verb/Collector/, which never blocks, but collects every written value into an array.
+
+Input and output channels can be set \emph{outside} of a channel using the \verb/.from(channel)/ and \verb/.to(channel)/ methods.
+However, they are by default inherited from the enclosing context, meaning that it is straightforward to spawn multiple concurrent processes that are attached to the same channels.
+
+Consider the example:
+
+\begin{verbatim}
+s { i = 0; loop { put i; i += 1 } }
+  .p { s { drain }.go; s { drain }.go }
+  .p { take 10 }.call
+\end{verbatim}
+In this example, we spawn four processes.
+The first line spawns an infinite loop that generates all positive integers.
+The second line spawns two processes, both reading from the same input and writing to the same output.
+Finally the third line spawns a process that reads and writes 10 values, then returns.
+The definition of \verb/drain/ is very simply:
+
+\begin{verbatim}
+def drain
+  loop { put(get) }
+end
+\end{verbatim}
+so that it completely drains the input channel into the output.
+But since there are two of these processes running concurrently, races are possible.
+But according to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed.
+Notably, the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning.
+But this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.
+
+% \verb{...} inline \verbatim
+% \verb+abc_def+
+
+% example of bash race conditions based on re-opening channels
+
+\section{Related Work}
+
+Python PEP 533; Go, Erlang; Guile Scheme, TCL;
+
 \end{document}

### 2018-10-15 doc/paper.tex
paper updates

@@
 
-\documentclass[format=sigconf, review=true, screen=true]{acmart}
+\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
 
@@
 This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.
+
 In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
@@
 
-PEP 533 Authors(TODO: cite properly) describe an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
-But channel-based systems generally allow for multiple concurrent readers and writers - so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
-We propose a system such that processes are interrupted as soon as it is no longer possible for them to resume.
+To meet these expectations, we need to design a semantic of process \emph{interruption}.
+
+\section{Interrupting on Operations vs. Eager Interruption}
+
+Consider the following example:
+
+\begin{verbatim}
+count-forever | ( take 5; do-other-operation ) | take 5
+\end{verbatim}
+
+The middle process will write five times, and then continue to do other processing in-thread that does not write to the standard output. 
+The final \verb/take 5/ operation will return after 5 inputs are read.
+It is important to decide, then, whether the middle process should be \emph{interrupted} in the middle of \verb/do-other-operation/, or should it be left alone?
+
+(TODO show race condition with eager-interruption)
+
+(TODO give example and show race condition with channel-reopening)
 
@@
 
-When a process is blocked on a channel, it will be woken up as soon as another process communicates on the other end.
+(TODO explain guiding principle: \emph{A process is interrupted exactly when it can no longer be woken up} and the addendum \emph{without spawning new processes})
+
+When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
 Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations.
@@
 
+(TODO therefore we add the constraint that processes must only interact with specifically registered channels)
+
+\section{Description of Semantics}
+
+(TODO: \verb/{add,remove}_{reader,writer}/ are hidden intrinsics, only called automatically at process spawn and exit (even for exits that are in error).
+and we only allow reads/writes for processes *attached* to channels, rather than freely through channel variables.
+explain this)
+
 We extend the notion of a synchronous channel with \emph{registered readers} and \emph{registered writers}, and add four intrinsic functions: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, each receiving a channel and a process handle.
 These store and delete processes from the channel's internal sets of registered readers and registered writers.
+
+In this way, we can know relatively certainly when a channel will never unblock - if either the read or the write end has no remaining processes holding it open, we immediately close it.
+While it is still possible for processes to attempt to attach new processes to a closed channel
+
 Additionally, \verb/remove_reader/ and \verb/remove_writer/ will cause the channel to close if their corresponding sets are empty after the operation.
@@
 The second line spawns two processes, both reading from the same input and writing to the same output.
-Finally the third line spawns a process that reads and writes 10 values, then returns.
-The definition of \verb/drain/ is very simply:
+Finally the third line spawns a process that reads and writes 10 values, then returns. The definition of \verb/drain/ is very simply:
 
@@
 But since there are two of these processes running concurrently, races are possible.
-But according to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed.
-Notably, the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning.
-But this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.
+
+According to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed.
+Notably, even though the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning, this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.
 
@@
 
+PEP 533 Authors(TODO: cite properly) describe an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
+But channel-based systems generally allow for multiple concurrent readers and writers - so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
+We propose a system such that processes are interrupted as soon as it is no longer possible for them to resume.
+
+The Bourne Again Shell (\"bash\")
+
 Python PEP 533; Go, Erlang; Guile Scheme, TCL;

### 2018-10-15 lib/magritte/free_vars.rb
Implement minor functions and tests for free var scaning
More specifically, implement call function for Value class
String and visit_command function for interpreter. Also implement
tests for free variable scanner to ensure that it behaves as
expected.

@@ -10,9 +10,10 @@ module Magritte
       end
 
       def visit_lambda(node)
-        unless node.patterns.all? { |n| n.is_a? Binder }
+        unless node.patterns.all? { |n| n.is_a? AST::Binder }
           raise "Not implemented yet!"
         end
+        #binding.pry
         used_vars = collect_from(node.bodies)
         bound_vars = node.patterns.map(&:name)
         used_vars - bound_vars

### 2018-10-15 lib/magritte/intepret.rb

@@ -13,6 +13,13 @@ module Magritte
         yield Proc.current.env.get(node.name)
       end
 
+      def visit_command(node)
+        args = []
+        command = visit(node.head)
+        args = node.args.each { |child| visit(child) { |x| args << x } }
+        command.call(args)).each { |x| yield x}
+      end
+
       def visit_vector(node)
         elems = []
         elems = node.elems.each { |child| visit(child) { |x| elems << x } }

### 2018-10-15 lib/magritte/value.rb

@@ -17,7 +17,7 @@ module Magritte
       def call(*args)
         # Semantics: Look up the thing in current env and call it
         # Proc.current.env
-        raise "TODO"
+        Proc.current.env.get(@value).call(*args)
       end
     end
 

### 2018-10-15 spec/ast_spec.rb

@@ -74,44 +74,7 @@ describe Magritte::AST do
     end
   end
 
-  describe "command nodes" do
-    it "has a head" do
-      assert { command.head == var }
-    end
-
-    it "has args" do
-      assert { command.args == ["arg1", "arg2", "arg3"] }
-    end
-
-    describe "maps" do
-      let(:mapped) { command.map { |val| (val.is_a? Magritte::AST::Variable) ? val.name : val }}
-
-      it "maps" do
-        assert { mapped.is_a? Magritte::AST::Command }
-        assert { mapped.head.is_a? String}
-        assert { mapped.head == "foo" }
-        assert { mapped.args == ["arg1", "arg2", "arg3"]}
-      end
-    end
-
-    it "equals itself" do
-      assert { command == command }
-    end
-
-    it "equals object with same state" do
-      assert { command == Magritte::AST::Command[var,["arg1","arg2","arg3"]] }
-    end
-
-    it "does not equal objects with different state" do
-      assert { command != Magritte::AST::Command[var,["arg1","arg2"]] }
-    end
-
-    it "does not equal different objects" do
-      assert { command != var }
-    end
-  end
-
-  describe "block nodes" do
+  describe "list recursive nodes" do
     it "has lines" do
       assert { block.lines == [command,command] }
     end

### 2018-10-15 spec/free_vars_spec.rb

@@ -7,8 +7,17 @@ describe Magritte::FreeVars do
     result[node]
   end
 
+  # Define helper functions for setting up tests
+  # Example: Magritte::AST::Variable["foo"] = _variable("foo")
+  Magritte::AST.constants.each do |c|
+    c = Magritte::AST.const_get(c)
+    define_method("_#{c.short_name}") do |*a|
+      c.make(*a)
+    end
+  end
+
   describe "a variable" do
-    let(:expr) { Magritte::AST::Variable["foo"] }
+    let(:expr) { _variable("foo") }
 
     it "has no free variables" do
       assert { free_vars.empty? }
@@ -16,10 +25,81 @@ describe Magritte::FreeVars do
   end
 
   describe "a lexical variable" do
-    let(:expr) {Magritte::AST::LexVariable["foo"] }
+    let(:expr) { _lex_variable("foo") }
 
     it "has a free variable" do
       assert { free_vars == Set.new(["foo"]) }
+    end
+  end
+
+  describe "a binder" do
+    let(:expr) { _binder("hoge") }
+
+    it "has no free variables" do
+      assert { free_vars.empty? }
+    end
+  end
+
+  describe "a lambda" do
+    let(:expr) { _lambda("x",[_binder("foo")],[_command(_variable("hoge"),[_lex_variable("bar")])]) }
+    let(:expr2) { _lambda("x",[_binder("foo")],[_command(_variable("hoge"),[_lex_variable("foo")])]) }
+
+    it "has a free variables" do
+      assert { free_vars == Set.new(["bar"]) }
+    end
+
+    it "has no free variables" do
+      assert { Magritte::FreeVars.scan(expr2)[expr2].empty? }
+    end
+  end
+
+  describe "a pipe" do
+    let(:expr) { _pipe(_binder("in"),_binder("out"),[]) }
+
+    it "has no free variables" do
+      assert { free_vars.empty? }
+    end
+  end
+
+  describe "a compensation" do
+    let(:expr) { _compensation(_lex_variable("hoge"),_lex_variable("bar"),"boo") }
+
+    it "has a free variable" do
+      assert { free_vars == Set.new(["hoge","bar"]) }
+    end
+  end
+
+  describe "a spawn" do
+    let(:expr) { _spawn(_block([_command(_binder("bar"),[_lex_variable("foo")])])) }
+
+    it "has a free variable" do
+      assert { free_vars == Set.new(["foo"]) }
+    end
+  end
+
[…]

### 2018-11-15 doc/paper.tex
stashed paper updates

@@
 
-\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
+% \documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
+\documentclass[sigplan,10pt,nosumlimits]{acmart}
 
 % don't print the conference ref for now
-\settopmatter{printacmref=false}
+% \settopmatter{printacmref=false}
 
@@
 
-\title{A Concurrent Semantic for Shell-Like Processes}
+\title{Concurrency for Pipe-based Programming}
 \author{Jeanine Adkisson}
-\affiliation{Tokyo Institute of Technology}
-\email{jneen@prg.is.titech.edu}
+\affiliation{\institution{Tokyo Institute of Technology}}
+\email{jneen@prg.is.titech.ac.jp}
 
 \author{Hidehiko Masuhara}
-\affiliation{Tokyo Institute of Technology}
-\email{masuhara@prg.is.titech.edu}
+\affiliation{\institution{Tokyo Institute of Technology}}
+\email{masuhara@prg.is.titech.ac.jp}
 
@@
 
+% thought experiment - parallel fetching
+% find an example of named pipes
+
 \begin{abstract}
-Despite the widespread use of GUI tools for configuration tasks, CLI tools, and more specifically shell-based interfaces remain a powerful standard for software distribution and integration.
-Several attempts have been made at solving the desktop-integration problem, most notably TCL(TODO: cite!) and Guile Scheme(TODO: cite!), but neither has been as succesful as they intended to be: the \emph{lingua franca} for a large portion of tools remains shell-oriented, relying on such concepts as standard I/O, string-based argument vectors, and the expectation that users will be able to freely pipe streams of bytes in a shell-like way.
+The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
+Beyond the most simple tasks, however, \texttt{bash} and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
 
-Unfortunately, shell languages have significant known interoperability issues, such as lack of any kind of usable encapsulation, and the lack of in-language data structures to represent any kind of value other than a string.
+We explore a pipe-based programming model that attempts to synthesize the good parts of UNIX shells, departing at key points, along with a supporting concurrency model and a semantic of interruption.
+
+% Despite the widespread use of GUI tools for configuration and orchestration tasks, CLI tools, and more specifically shell-based interfaces remain a powerful standard for software distribution and integration.
+Several attempts have been made at solving the desktop-integration problem: how should we allow programs written in different languages and on different platforms to be scripted and integrated in a coherent system?
+The most notable attempts at this problem have been TCL(TODO: cite!) and Guile Scheme(TODO: cite!), but neither has been as succesful as they intended to be: the common language for a large portion of tools remains shell-based, relying on such concepts as standard I/O, string-based argument vectors, and the expectation that users will be able to freely pipe streams of bytes in a pipe-based way.
+% 
+% Unfortunately, existing UNIX shell languages have significant known interoperability issues, such as lack of any kind of usable encapsulation, and the lack of in-language data structures to represent any kind of value other than a string.
 We believe that a shell-compatible language with well-defined semantics that could also function as a general-purpose systems language could go a long way towards solving the desktop-integration problem.
-
-On the other hand, a persistent problem with more established systems like Erlang or Go is that they don't account for closing channels, and require manual cleanup of processes.
+% 
+% On the other hand, a persistent problem with more established systems like Erlang or Go is that they don't account for closing channels, and require manual cleanup of processes.
 In a systems programming context, errors and hangups are commonplace and expected, and it is critical to have some measure of automatic process cleanup for even naively-written code.
-
-We propose a channel-based communication system that integrates with a workflow compensation system and is capable of recovering from errors and, importantly, cleaning up processes when communication halts.
+% 
+% We propose a channel-based communication system that integrates with a workflow compensation system and is capable of recovering from errors and, importantly, cleaning up processes when communication halts.
 \end{abstract}
@@
 
-Consider the following pseudo-code for a shell-like language:
+The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
+Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
+
+In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes in which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.
+
+Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are specified for performance reasons.
+However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, resulting in the ability to send data in batch without resorting to a queue that may interleave writes.
+
+A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used.
+Consider the following pseudo-code for a pipe-based language:
 
@@
 
-In a pipeline-based programming system such as a shell, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
-A user will also expect that, after the 10 lines are processed, the file is closed, and that all three processes exit.
+In a pipe-based language, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
+A user will also expect that, after the first 10 lines are processed, the file is closed, and that all three processes exit.
 This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.
@@
 
-To meet these expectations, we need to design a semantic of process \emph{interruption}.
+\section{Design Considerations}
+\paragraph{Compensation.} In order to observe interruption, we define a system of \emph{compensation actions}, which will be run in the case of an interruption.
+For example, the function \verb/file-lines/ above could be implemented as:
 
-\section{Interrupting on Operations vs. Eager Interruption}
+\begin{verbatim}
+(file-lines ?fname) = (
+  handle = (open-file $fname) %%! close-file $fname
+  until (=> eof? $handle) (=> read-until "\n" $handle)
+  checkpoint
+)
+\end{verbatim}
 
-Consider the following example:
+In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next \verb/checkpoint/ invocation, or when the process is interrupted.
+In this way, we can ensure that the file is closed when the process exits.
+Compensations are the only way in which interruptions can be directly observed by client code - without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program.
+Thus, the primary task of our system is \emph{to run a process's compensation actions at the appropriate time}.
+
+\paragraph{Interrupting on Operations vs. Eager Interruption.}
+
+In designing an interruption semantic, we must first decide whether processes are eagerly interrupted when their channels close, or only interrupted on channel operations.
+
+Consider the following example, with the assumption that \verb/do-other-operation/ does not write to the standard output:
 
@@
 
-(TODO give example and show race condition with channel-reopening)
-
-\section{Channel Closing / Interruption}
+\paragraph{Channel registration.}
 
@@
 
-\section{Description of Semantics}
+\paragraph{Closing Permanence.}
+(TODO: explain why we do not reopen channels)
+
+\section{Description of Semantics.}
 

### 2018-11-16 lib/magritte/builtins.rb
Start implementing standard library

@@ -8,7 +8,7 @@ module Magritte
       @builtins.each do |(name, func)|
         env.let(name, func)
       end
-      env
+      load_lib("Standard lib", STD_LIB, env)
     end
 
     @builtins = []
@@ -91,6 +91,22 @@ module Magritte
       File.foreach(fname.value) do |line|
         put(Value::String.new(line))
       end
+    end
+
+    # Initialize environment with functions that can be defined in the language itself
+    STD_LIB = """
+    (range ?n) = (count-forever | take %n)
+    """
+
+    def self.load_lib(lib_name, source, env)
+      # Transform the lib string into an ast
+      ast = Parser.parse(Skeleton.parse(Lexer.new(lib_name, source)))
+      c = Collector.new
+      env.own_outputs[0] = c
+      # Evaluate the ast, which will add the lib functions to the env
+      Proc.spawn(Code.new { Interpret.interpret(ast) }, env).start
+      c.wait_for_close
+      env
     end
   end
 end

### 2018-11-16 lib/magritte/lexer.rb

@@ -37,6 +37,10 @@ module Magritte
         :arrow,
       ])
 
+      FREE_NL = Set.new([
+        :lbrack,
+      ])
+
       def initialize(type, value, range)
         @type = type
         @value = value
@@ -56,7 +60,7 @@ module Magritte
       end
 
       def free_nl?
-        false #TODO
+        FREE_NL.include?(@type)
       end
 
       def is?(type)

### 2018-11-16 lib/magritte/matcher.rb

@@ -188,8 +188,8 @@ module Magritte
         Any[args]
       end
 
-      def all(*args)
-        All[args]
+      def all(arg)
+        All[arg]
       end
 
       def _

### 2018-11-16 lib/magritte/parser.rb

@@ -26,7 +26,7 @@ module Magritte
     end
 
     def parse_root(skel)
-      elems = skel.elems.map do |item|
+      elems = skel.sub_items.map do |item|
         parse_line(item)
       end
       AST::Block[elems]
@@ -41,7 +41,16 @@ module Magritte
         return AST::Pipe[parse_line(before), parse_command(after)]
       end
 
+      # Match any line that has an equal sign
       item.match(lsplit(~_, token(:equal), ~_)) do |lhs, rhs|
+        # Special syntax for lambda assignment
+        lhs.match(singleton(nested(:lparen, starts(~token(:bare), ~_)))) do |var, bindings|
+          unless bindings.elems.all? { |elem| elem.token?(:bind) }
+            error!(bindings,"Invalid pattern")
+          end
+          return AST::Assignment[[parse_term(var)], [parse_lambda("anon@#{item.range.repr}", bindings, rhs)]]
+        end
+        # Normal assignment
         unless lhs.elems.all? { |elem| elem.token?(:bare) || elem.token?(:var) || elem.token?(:lex_var) }
           error!(lhs, "Invalid lhs")
         end
@@ -55,6 +64,14 @@ module Magritte
 
     def parse_vector(vec)
       AST::Vector[parse_terms(vec)]
+    end
+
+    def parse_lambda(name, bindings, bodies)
+      unless bindings.elems.all? { |e| e.token?(:bind) }
+        error!(term, "TODO: Support patterns")
+      end
+      patterns = bindings.elems.map { |e| AST::Binder[e.value] }
+      return AST::Lambda[name, patterns, [parse_line(bodies)]]
     end
 
     def parse_command(command)
@@ -122,11 +139,7 @@ module Magritte
 
       term.match(nested(:lparen,~_)) do |item|
         item.match(lsplit(~_, token(:arrow), ~_)) do |before, after|
-          unless before.elems.all? { |e| e.token?(:bind) }
-            error!(term, "TODO: Support patterns")
-          end
-          patterns = before.elems.map { |e| AST::Binder[e.value] }
-          return AST::Lambda["anon@#{term.range.repr}", patterns, [parse_line(after)]]
+          return parse_lambda("anon@#{term.range.repr}", before, after)
         end
 
         return AST::Subst[item.sub_items.map { |e| parse_line(e) }]

### 2018-11-16 lib/magritte/skeleton.rb

@@ -83,6 +83,10 @@ module Magritte
     class Root < Base
       defdata :elems # Same question as for Item class
 
+      def sub_items
+        elems
+      end
+
       def repr
         "(#{elems.map(&:repr).join(" ")})"
       end

### 2018-11-16 spec/interpret_spec.rb

@@ -135,4 +135,33 @@ describe Magritte::Interpret do
       end
     end
   end
+
+  describe "special syntax" do
+    describe "lambda assignment" do
+      let(:input) {
+        """
+        (f ?x) = put $x
+        f 5
+        """
+      }
+
+      it do
+        assert { result == "5" }
+      end
+    end
+  end
+
+  describe "standard library" do
+    describe "range" do
+      let(:input) {
+        """
+        range 5
+        """
+      }
+
+      it do
+        assert { result == "0\n1\n2\n3\n4" }
+      end
+    end
+  end
 end

### 2018-11-16 spec/matcher_spec.rb

@@ -6,9 +6,6 @@ describe Magritte::Matcher do
   let(:matcher) { raise "Abstract" }
   let(:match_vars) { matcher.match_vars(tree) }
 
-  # Maybe add functionality to match many (not just singleton) matches
-  # E.g. a b c
-
   describe "a single token" do
     let(:input) {
       """
@@ -129,6 +126,24 @@ describe Magritte::Matcher do
         assert { match_vars.size == 2 }
         assert { match_vars[0].repr == "(.bare/p1 .pipe .bare/p2)" }
         assert { match_vars[1].repr == "(.bare/p3)" }
+      end
+    end
+  end
+
+  describe "lambda" do
+    let(:input) {
+      """
+      (f ?x) = add $x 1
+      """
+    }
+
+    describe "test lsplit" do
+      let(:matcher) { lsplit(~_, token(:equal), ~_) }
+
+      it "works" do
+        assert { match_vars.size == 2 }
+        assert { match_vars[0].repr == "([lparen|.bare/f .bind/x|rparen])" }
+        assert { match_vars[1].repr == "(.bare/add .var/x .num/1)" }
       end
     end
   end

### 2018-11-16 spec/skeleton_spec.rb

@@ -68,7 +68,7 @@ describe Magritte::Skeleton do
     }
 
     it "parses correctly" do
-      assert { tree.repr == "((.bare/s [lbrack|([lparen|.arrow .bare/c .gt .lex_var/ch|rparen]) ([lparen|.arrow .bare/d .lt .lex_var/ch|rparen])|rbrack]))" }
+      assert { tree.repr == "((.bare/s [lbrack|[lparen|.arrow .bare/c .gt .lex_var/ch|rparen] [lparen|.arrow .bare/d .lt .lex_var/ch|rparen]|rbrack]))" }
     end
   end
 

### 2018-11-16 lib/magritte.rb
Last throwaway

@@ -14,6 +14,7 @@ module Magritte
   end
 
   LIB_DIR = File.dirname(__FILE__)
+  ROOT_DIR = File.dirname(LIB_DIR)
   load "#{LIB_DIR}/magritte/log.rb"
 
   # AST

### 2018-11-16 lib/magritte/env.rb

@@ -72,7 +72,7 @@ module Magritte
     end
 
     def splice(new_parent)
-      Env.new(new_parent, @keys, @own_inputs, @own_outputs)
+      Env.new(new_parent, @keys.dup, @own_inputs.dup, @own_outputs.dup)
     end
 
     def slice(keys)

### 2018-11-16 spec/interpret_spec.rb

@@ -160,7 +160,45 @@ describe Magritte::Interpret do
       }
 
       it do
-        assert { result == "0\n1\n2\n3\n4" }
+        assert { results == ["0", "1", "2", "3", "4"] }
+      end
+    end
+
+    describe "repeat" do
+      let(:input) {
+        """
+        repeat 7 3
+        """
+      }
+
+      it do
+        assert { results == ["7", "7", "7"] }
+      end
+    end
+
+    describe "inc" do
+      let(:input) {
+        """
+        inc 1
+        inc 5
+        """
+      }
+
+      it do
+        assert { results == ["2","6"] }
+      end
+    end
+
+    describe "dec" do
+      let(:input) {
+        """
+        dec 1
+        dec 5
+        """
+      }
+
+      it do
+        assert { results == ["0","4"] }
       end
     end
   end

### 2018-11-19 lib/magritte/ast.rb
redo command parsing with a With node (breaks parser tests)

@@ -47,9 +47,13 @@ module Magritte
       defrec :target
     end
 
+    class With < Tree::Node
+      deflistrec :redirects
+      defrec :expr
+    end
+
     class Command < Tree::Node
       deflistrec :vec
-      deflistrec :redirects
 
       def initialize(*)
         super

### 2018-11-19 lib/magritte/parser.rb

@@ -76,23 +76,42 @@ module Magritte
 
     def parse_command(command)
       redirects = []
-      matched = true
-      while matched
-        matched = command.match(rsplit(~_, ~any(token(:gt), token(:lt)), ~_)) do |direct, before, target|
-          unless target.elems.size == 1
-            error!(Skeleton::Item[[direct]+target.elems], "Redirect target must be a single term")
+      vec = []
+      while true do
+        break if command.elems.empty?
+
+        next if command.match(starts(~any(token(:gt), token(:lt)), ~_)) do |dir, rest|
+          error!(command, 'redirect at end') if rest.elems.empty?
+          target, *rest = rest.elems
+          direction = dir.token?(:gt) ? :> : :<
+          if target.nested?(:lparen)
+            error!(target, 'TODO: redir to paren')
+          else
+            redirects << AST::Redirect[direction, parse_term(target)]
           end
-          command = before
-          redirects << AST::Redirect[direct.token?(:gt) ? :> : :<, parse_term(target.elems.first)]
+
+          command = Skeleton::Item[rest]
+        end
+
+        next if command.match(starts(~_, ~_)) do |head, rest|
+          vec << head
+          command = rest
         end
       end
 
-      command.match(singleton(nested(:lparen,~_))) do |i|
-        return parse_root(i)
+      bare_command = Skeleton::Item[vec]
+
+      bare_command.match(singleton(nested(:lparen,~_))) do |i|
+        return with(redirects, parse_root(i))
       end
 
-      vec = parse_terms(command)
-      AST::Command[vec, redirects]
+      vec = parse_terms(bare_command)
+      with(redirects, AST::Command[vec])
+    end
+
+    def with(redirects, expr)
+      return expr if redirects.empty?
+      AST::With[redirects, expr]
     end
 
     def parse_terms(terms)

### 2018-11-19 doc/paper.tex
update paper

@@
 
+% from matias
+\usepackage{balance}
+
 \usepackage{listings}
 
-\title{Concurrency for Pipe-based Programming}
+\title{Pipe-based Programming}
 \author{Jeanine Adkisson}
@@
 \begin{abstract}
+Shell languages such as bash are designed to integrate with an OS, which mainly involves managing processes with implicit input and output streams.
+They also attempt to do this in a compact way that could be reasonably typed on a command-line interface.
+
+However, existing shell languages are not sufficient to serve as general-purpose languages - values are not observable except in raw streams of bytes, and they lack modern language features such as lexical scope and higher-order functions.
+
+By way of a new programming language, \emph{Magritte}, we propose a general-purpose programming language with semantics similar to bash.
+In this paper, we discuss the early design of such a system, in which the primary unit of composition, like bash, is processes with asynchronous inputs and outputs, which can be read from or written to at any time, and which can be chained together via a pipe operator.
+We also explore concurrency semantics for such a language.
+
+We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
 The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
-Beyond the most simple tasks, however, \texttt{bash} and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
 
-We explore a pipe-based programming model that attempts to synthesize the good parts of UNIX shells, departing at key points, along with a supporting concurrency model and a semantic of interruption.
+% Beyond the most simple tasks, however, \texttt{bash} and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
+% %
+% We explore a pipe-based programming model that attempts to synthesize the good parts of UNIX shells, departing at key points, along with a supporting concurrency model and a semantic of interruption.
+%
+%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Functional and object-oriented programming suffer from a disconnect between the concept of \emph{functions} and \emph{programs}.
+Whereas functions and methods receive values as arguments and return single values as a result, programs can receive data not only as arguments but also asynchronously through an input, and write results asynchronously to an output.
+%
+% In functional and object-oriented languages, this can result in not only a significant amount of abstraction overhead for dealing with external functions, but the immediacy of function arguments and return values can make composing concurrent programs difficult.
+%
+% By way of a new programming language, \emph{Magritte}, we propose a programming model in which the primary unit of composition is not immediate function arguments and return values, but instead processes with asynchronous inputs and outputs, which can be read from or written to at any time, and which can be chained together via a pipe operator, much like programs in a UNIX environment.
+We also explore a concurrency semantic for such a language.
+%
+% We explore many implementation and programming challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms that would be difficult to write or compose in a functional or object-oriented language.
 
@@
 \end{document}
+
+%# fn pipe(lhs, rhs, inp, out) {
+%#   c = channel()
+%#   go fn () { lhs(inp, c); };
+%#   rhs(c, out)
+%# }
+%#
+%# collect = collector()
+%# pipe_collect(curry(range, 10), curry(fan, 2,
+%  fn(el, inp, out) { write(el + 1, out); }
+%), inp);
+%#

### 2018-11-19 lib/magritte/builtins.rb
make `make-channel` return a value channel
(and use a dash)

@@ -43,8 +43,8 @@ module Magritte
       put(get)
     end
 
-    builtin :make_channel, [] do
-      put Channel.new
+    builtin :'make-channel', [] do
+      put Value::Channel.new(Channel.new)
     end
 
     builtin :for, [:Vector] do |vec|

### 2018-11-19 lib/magritte/repl.rb
implement input streaming
and allow streamer reset! to work even if it hasn't been properly
closed, which is important in some error cases

@@ -3,11 +3,32 @@ module Magritte
   class REPL
     def initialize(argv)
       @line_num = 0
+      @input_num = 0
       @mutex = Mutex.new
       @streamer = Streamer.new do |val|
         @mutex.synchronize { puts val.repr }
       end
-      @env = Env.base.extend([], [@streamer])
+
+      @input = InputStreamer.new do
+        begin
+          source = @mutex.synchronize {
+            @input_num += 1
+            Readline.readline('<= ', false)
+          }
+
+          (puts; next) if source.nil?
+
+          source = "for [#{source}]"
+
+          ast = Parser.parse(Skeleton.parse(Lexer.new(input_name, source)))
+
+          Spawn.s_ { Interpret.interpret(ast) }.collect
+        rescue CompileError, RuntimeError
+          []
+        end
+      end
+
+      @env = Env.base.extend([@input], [@streamer])
     end
 
     def self.run(argv)
@@ -18,24 +39,31 @@ module Magritte
       "repl~#{@line_num}"
     end
 
+    def input_name
+      "#{source_name}<#{@input_num-1}"
+    end
+
     def evaluate(source)
       ast = Parser.parse(Skeleton.parse(Lexer.new(source_name, source)))
       Proc.spawn(Code.new { Interpret.interpret(ast) }, @env).start
       @streamer.wait_for_close
     rescue ::Interrupt => e
       #pass
-    ensure
-      @streamer.reopen!
+      false
+    rescue CompileError, RuntimeError => e
+      @mutex.synchronize { puts "error: #{e.class.name}\n#{e.to_s}" }
+      false
+    else
+      @streamer.reset!
+      @input.reset!
+      true
     end
 
     def process_line
       string = @mutex.synchronize { Readline.readline("; ", true) }
       raise IOError if string.nil?
+      return false if string =~ /\A\s*\z/m
       evaluate(string)
-      return true
-    rescue CompileError => e
-      @mutex.synchronize { puts "error: #{e.class.name}\n#{e.to_s}" }
-      return false
     end
 
     def run

### 2018-11-19 lib/magritte/streamer.rb

@@ -42,14 +42,57 @@ module Magritte
       end
     end
 
-    def reopen!
+    def reset!
       @mutex.synchronize do
-        raise "Oops" if @open
+        @open = true
         @close_waiters.clear
         @writers.clear
-        @open = true
       end
     end
+  end
+
+  class InputStreamer < Channel
+    def initialize(&b)
+      @block = b
+      @readers = Set.new
+      @mutex = Mutex.new
+      @open = true
+      @queue = []
+    end
+
+    def add_reader(*); end
+    def add_writer(*); end
+    def remove_reader(*); end
+    def remove_writer(*); end
+
+    def write(*); end
+
+    def read
+      @mutex.synchronize do
+        call_out while @open && @queue.empty?
+
+        next unless @open
+
+        return @queue.shift
+      end
+
+      Proc.current.interrupt!
+    end
+
+    def reset!
+      @mutex.synchronize do
+        @open = true
+        @queue.clear
+        @readers.clear
+      end
+    end
+
+  private
+    def call_out
+      input = @block.call or return (@open = false)
+      @queue.concat(input)
+    end
+
   end
 
   class Collector < Streamer

### 2018-11-19 lib/magritte/skeleton.rb
implement Skeleton#nested?

@@ -20,6 +20,10 @@ module Magritte
         false
       end
 
+      def nested?(type = nil)
+        false
+      end
+
       def range
         raise "Abstract"
       end
@@ -56,6 +60,11 @@ module Magritte
 
       def repr
         "[#{open.repr}|#{elems.map(&:repr).join(" ")}|#{close.repr}]"
+      end
+
+      def nested?(type = nil)
+        return true unless type
+        open.is?(type)
       end
 
       def range

### 2018-11-19 lib/magritte/parser.rb
better error handling for bang parsing

@@ -119,10 +119,10 @@ module Magritte
       elems = terms.elems.dup
       while x = elems.shift do
         if x.token?(:bang)
-          raise "Parse Error" if out == []
+          error!(terms, 'cannot start with a !') if out == []
           source = out.pop
           lookup = elems.shift
-          raise "Parse Errror" if lookup.nil?
+          error!(terms, 'cannot end with a !') if lookup.nil?
           out << AST::Access[source, parse_term(lookup)]
         else
           out << parse_term(x)

### 2018-11-19 spec/code_spec.rb
testing: add support for abstract(:symbl)

@@ -24,6 +24,7 @@ describe Magritte::Code do
 
   let(:output) { with_no_dangling_threads { code.spawn_collect } }
 
+  abstract(:code)
   def self.code(&b)
     let(:code) { Magritte::Code.new { Magritte::Code::DSL.instance_eval(&b) } }
   end

### 2018-11-19 spec/free_vars_spec.rb

@@ -1,5 +1,5 @@
 describe Magritte::FreeVars do
-  let(:expr) { raise "Abstract" }
+  abstract(:expr)
   let(:result) { Magritte::FreeVars.scan(expr) }
 
   def free_vars(node = nil)

### 2018-11-19 spec/interpret_spec.rb

@@ -1,5 +1,6 @@
 describe Magritte::Interpret do
-  let(:input) { "" }
+  abstract(:input)
+
   let(:lex) { Magritte::Lexer.new("test",input) }
   let(:skel) { Magritte::Skeleton::Parser.parse(lex) }
   let(:ast) { Magritte::Parser.parse_root(skel) }

### 2018-11-19 spec/lexer_spec.rb

@@ -1,5 +1,5 @@
 describe Magritte::Lexer do
-  let(:input) { "" }
+  abstract(:input)
   let(:lex) { Magritte::Lexer.new("test",input) }
   let(:tokens) { lex.to_a }
 

### 2018-11-19 spec/matcher_spec.rb

@@ -1,6 +1,6 @@
 describe Magritte::Matcher do
   include Magritte::Matcher::DSL
-  let(:input) { "" }
+  abstract(:input)
   let(:lex) { Magritte::Lexer.new("test",input) }
   let(:tree) { Magritte::Skeleton::Parser.parse(lex).elems.first }
   let(:matcher) { raise "Abstract" }

### 2018-11-19 spec/parser_spec.rb

@@ -1,5 +1,6 @@
 describe Magritte::Parser do
-  let(:input) { "" }
+  abstract(:input)
+
   let(:lex) { Magritte::Lexer.new("test",input) }
   let(:skel) { Magritte::Skeleton::Parser.parse(lex) }
   let(:ast) { Magritte::Parser.parse_root(skel) }

### 2018-11-19 spec/skeleton_spec.rb

@@ -1,5 +1,5 @@
 describe Magritte::Skeleton do
-  let(:input) { "" }
+  abstract(:input)
   let(:lex) { Magritte::Lexer.new("test",input) }
   let(:tree) { Magritte::Skeleton::Parser.parse(lex) }
 

### 2018-11-19 spec/support/abstract.rb

@@ -1,0 +1,7 @@
+module HasAbstract
+  def abstract(*a)
+    let(*a) { raise "abstract: #{a.map(&:inspect).join(', ')}" }
+  end
+end
+
+Minitest::Spec.send(:extend, HasAbstract)

### 2018-11-19 spec/free_vars_spec.rb
Fix broken tests

@@ -1,23 +1,21 @@
 describe Magritte::FreeVars do
-  abstract(:expr)
-  let(:result) { Magritte::FreeVars.scan(expr) }
+  abstract(:input)
+  let(:lex) { Magritte::Lexer.new("test",input) }
+  let(:skel) { Magritte::Skeleton::Parser.parse(lex) }
+  let(:ast) { Magritte::Parser.parse_root(skel) }
+  let(:result) { Magritte::FreeVars.scan(ast) }
 
   def free_vars(node = nil)
-    node ||= expr
+    node ||= ast
     result[node]
   end
 
-  # Define helper functions for setting up tests
-  # Example: Magritte::AST::Variable["foo"] = _variable("foo")
-  Magritte::AST.constants.each do |c|
-    c = Magritte::AST.const_get(c)
-    define_method("_#{c.short_name}") do |*a|
-      c.make(*a)
-    end
-  end
-
   describe "a variable" do
-    let(:expr) { _variable("foo") }
+    let(:input) {
+      """
+      $foo
+      """
+    }
 
     it "has no free variables" do
       assert { free_vars.empty? }
@@ -25,7 +23,11 @@ describe Magritte::FreeVars do
   end
 
   describe "a lexical variable" do
-    let(:expr) { _lex_variable("foo") }
+    let(:input) {
+      """
+      %foo
+      """
+    }
 
     it "has a free variable" do
       assert { free_vars == Set.new(["foo"]) }
@@ -33,7 +35,11 @@ describe Magritte::FreeVars do
   end
 
   describe "a binder" do
-    let(:expr) { _binder("hoge") }
+    let(:input) {
+      """
+      ?x
+      """
+    }
 
     it "has no free variables" do
       assert { free_vars.empty? }
@@ -41,36 +47,61 @@ describe Magritte::FreeVars do
   end
 
   describe "a lambda" do
-    let(:expr) { _lambda("x",[_binder("foo")],[_command([_variable("hoge"),_lex_variable("bar")],[])]) }
-    let(:expr2) { _lambda("x",[_binder("foo")],[_command([_variable("hoge"),_lex_variable("foo")],[])]) }
+    describe "has a free variable" do
+      let(:input) {
+        """
+        (x ?foo) = $hoge %bar
+        """
+      }
 
-    it "has a free variables" do
-      assert { free_vars == Set.new(["bar"]) }
+      it "has a free variables" do
+        assert { free_vars == Set.new(["bar"]) }
+      end
     end
[…]

### 2018-11-19 spec/parser_spec.rb

@@ -58,12 +58,12 @@ describe Magritte::Parser do
       """
     }
 
-    it "parses correctly" do
-      assert { ast.elems.size == 1 }
-      assert { ast.elems.first.is_a?(Magritte::AST::Command) }
-      assert { ast.elems.first.vec.inspect == "[#<Magritte::AST::String[\"multi-arg-command\"]>, #<Magritte::AST::String[\"arg1\"]>, #<Magritte::AST::String[\"arg2\"]>]" }
-      assert { ast.elems.first.redirects.inspect == "[#<Magritte::AST::Redirect[:<, *#<Magritte::AST::Variable[\"in\"]>]>, #<Magritte::AST::Redirect[:>, *#<Magritte::AST::Variable[\"out\"]>]>]" }
-    end
+    #it "parses correctly" do
+    #  assert { ast.elems.size == 1 }
+    #  assert { ast.elems.first.is_a?(Magritte::AST::Command) }
+    #  assert { ast.elems.first.vec.inspect == "[#<Magritte::AST::String[\"multi-arg-command\"]>, #<Magritte::AST::String[\"arg1\"]>, #<Magritte::AST::String[\"arg2\"]>]" }
+    #  assert { ast.elems.first.redirects.inspect == "[#<Magritte::AST::Redirect[:<, *#<Magritte::AST::Variable[\"in\"]>]>, #<Magritte::AST::Redirect[:>, *#<Magritte::AST::Variable[\"out\"]>]>]" }
+    #end
   end
 
   describe "lambda" do

### 2018-12-06 lib/magritte/parser.rb
Refactor parsing of assignment

@@ -43,38 +43,7 @@ module Magritte
 
       # Match any line that has an equal sign
       item.match(lsplit(~_, token(:equal), ~_)) do |lhs, rhs|
-        # Special syntax for lambda assignment
-        lhs.match(singleton(nested(:lparen, starts(~any(token(:bare),token(:var)), ~_)))) do |var, bindings|
-          # Determine whether we got a bare word or an access variable
-          is_bare = var.token?(:bare)
-          # Check that there're at least 2 elements in binding unless we have a bare word as a variable
-          # These two elements should be the bang token and bare token used in an access
-          unless is_bare || bindings.elems.length > 1
-            error!(var,"Invalid access variable")
-          end
-          # Pop the bang and bare token used for access and put it in the variable
-          if !is_bare
-            var = Skeleton::Item[[var, bindings.elems.shift, bindings.elems.shift]]
-          end
-          # Check all remainding bindings are actually bindings
-          unless bindings.elems.all? { |elem| elem.token?(:bind) }
-            error!(bindings,"Invalid pattern")
-          end
-          if is_bare
-            var = parse_term(var)
-            return AST::Assignment[[var], [parse_lambda(var.value, bindings, rhs)]]
-          else
-            var = parse_terms(var)
-            return AST::Assignment[var, [parse_lambda(var.first.lookup, bindings, rhs)]]
-          end
-        end
-
-        # Normal assignment
-        unless lhs.elems.all? { |elem| elem.token?(:bare) || elem.token?(:var) || elem.token?(:lex_var) || elem.token?(:bang) }
-          error!(lhs, "Invalid lhs")
-        end
-
-        return AST::Assignment[parse_terms(lhs), parse_terms(rhs)]
+        return parse_assignment(lhs, rhs)
       end
 
       # Parse double &
@@ -108,6 +77,40 @@ module Magritte
 
     def parse_vector(vec)
       AST::Vector[parse_terms(vec)]
+    end
+
+    def parse_assignment(lhs, rhs)
+      # Check if lhs have parenthesis
+      # In this case we're doing a special lambda assigment
+      lhs.match(singleton(nested(:lparen, starts(~_, ~_)))) do |var, bindings|
+        lambda_name = ""
+
+        # If we have an access variable we need to move the bang + one more token
+        # to the var
+        if !var.token?(:bare) && bindings.elems.length > 1 && bindings.elems.first.token?(:bang)
[…]

### 2018-12-06 spec/interpret_spec.rb
Add additional assignment tests

@@ -123,6 +123,34 @@ describe Magritte::Interpret do
         end
       end
 
+      describe "dynamic assignment" do
+        let(:input) {
+          """
+          x = 2
+          $x = -1
+          put $x
+          """
+        }
+
+        it do
+          assert { result == "-1" }
+        end
+      end
+
+      describe "lexical assignment" do
+        let(:input) {
+          """
+          x = 2
+          %x = -1
+          put $x
+          """
+        }
+
+        it do
+          assert { result == "-1" }
+        end
+      end
+
       describe "access assignment" do
         let(:input) {
           """
@@ -187,6 +215,36 @@ describe Magritte::Interpret do
       let(:input) {
         """
         (f ?x) = put $x
+        f 5
+        put $f
+        """
+      }
+
+      it do
+        assert { results == ["5","<func:f>"] }
+      end
+    end
+
+    describe "lambda assignment with dynamic var" do
+      let(:input) {
+        """
+        f = 1
+        ($f ?x) = put $x
+        f 5
+        put $f
+        """
+      }
+
+      it do
+        assert { results == ["5","<func:f>"] }
+      end
+    end
+
+    describe "lambda assignment with lexical var" do
+      let(:input) {
+        """
+        f = 1
+        (%f ?x) = put $x
         f 5
         put $f
         """

### 2018-12-06 spec/interpret_spec.rb
Update environment test to get better coverage

@@ -362,7 +362,7 @@ describe Magritte::Interpret do
       let(:input) {
         """
         e = {x = 1; y = 0}
-        put $e!x $e!y
+        put %e!x %e!y
         """
       }
 

### 2018-12-06 lib/magritte/skeleton.rb
Remove irrelevant comment

@@ -132,7 +132,7 @@ module Magritte
       end
 
       def parse(lexer)
-        loop do # Is there any reason why we do a "loop do" instead of "lexer.each do |token|"
+        loop do
           token = lexer.next
           if token.eof? and @open.nil?
             return Root[items]

### 2018-12-07 lib/magritte/parser.rb
Correctly parse anon lambdas with multi-line bodies

@@ -204,8 +204,21 @@ module Magritte
       end
 
       term.match(nested(:lparen,~_)) do |item|
+        # Anon lambda spanning one line
         item.match(lsplit(~_, token(:arrow), ~_)) do |before, after|
           return parse_lambda("anon@#{term.range.repr}", before, after)
+        end
+
+        # Anon lambda spanning multiple lines
+        item.elems.first.match(lsplit(~_, token(:arrow), ~_)) do |bindings, body|
+          name = "anon@#{term.range.repr}"
+          bodies = item.elems.drop(1).unshift(body)
+
+          unless bindings.elems.all? { |e| e.token?(:bind) }
+            error!(term, "TODO: Support patterns")
+          end
+          patterns = bindings.elems.map { |e| AST::Binder[e.value] }
+          return AST::Lambda[name, patterns, [AST::Block[bodies.map { |body| parse_line(body) }]]]
         end
 
         return AST::Subst[item.sub_items.map { |e| parse_line(e) }]

### 2018-12-07 spec/interpret_spec.rb

@@ -303,6 +303,18 @@ describe Magritte::Interpret do
         assert { results == ["0", "1", "2"] }
       end
     end
+
+    describe "lambda body with anon lambda" do
+      let(:input) {
+        """
+        put 1 2 3 | each (?a => put 10; put $a)
+        """
+      }
+
+      it do
+        assert { results == %w(10 1 10 2 10 3) }
+      end
+    end
   end
 
   describe "standard library" do

### 2018-12-07 lib/magritte/free_vars.rb
Add scoping for subst

@@ -17,6 +17,10 @@ module Magritte
         shadow(node.bodies, bound_vars, to_bind)
       end
 
+      def visit_subst(node, bound_vars)
+        visit_block(node, bound_vars)
+      end
+
       def visit_block(node, bound_vars)
         out = Set.new
         so_far = Set.new

### 2018-12-07 lib/magritte.rb
Refactor procs with an explicit frame stack
Register frames instead of processes.

@@ -26,6 +26,7 @@ module Magritte
   load "#{LIB_DIR}/magritte/parser.rb"
 
   # Runtime
+  load "#{LIB_DIR}/magritte/frame.rb"
   load "#{LIB_DIR}/magritte/status.rb"
   load "#{LIB_DIR}/magritte/std.rb"
   load "#{LIB_DIR}/magritte/code.rb"

### 2018-12-07 lib/magritte/frame.rb

@@ -1,0 +1,48 @@
+module Magritte
+  class Frame
+    attr_reader :env
+    def initialize(env, p)
+      @env = env
+      @proc = p
+      @compensations = []
+    end
+
+    def compensate(status)
+      @compensations.each(&:run)
+      unregister_channels
+    end
+
+    def checkpoint
+      @compensations.each(&:run_checkpoint)
+      unregister_channels
+    end
+
+    def add_compensation(comp)
+      @compensations << comp
+    end
+
+    def repr
+      "f"
+    end
+
+    def inspect
+      "#<Frame #{repr}>"
+    end
+
+    def thread
+      @proc.thread
+    end
+
+    def open_channels
+      @env.each_input { |c| c.add_reader(self) }
+      @env.each_output { |c| c.add_writer(self) }
+    end
+
+  private
+    def unregister_channels
+      PRINTER.p :unregister_channels => [@env.own_inputs, @env.own_outputs]
+      @env.each_input { |c| c.remove_reader(self) }
+      @env.each_output { |c| c.remove_writer(self) }
+    end
+  end
+end

### 2018-12-07 lib/magritte/proc.rb

@@ -33,17 +33,22 @@ module Magritte
           # wait for Proc#start
           start_mutex.lock
           Thread.current[:status] = Status[:incomplete]
-          Thread.current[:status] = Proc.current.with_compensations { code.run }
+          Thread.current[:status] = begin
+            code.run
+          rescue Interrupt => e
+            Proc.current.compensate(e)
+          end
         rescue Exception => e
           PRINTER.p :exception
           PRINTER.p e
           PRINTER.puts e.backtrace
-          Proc.current.compensate_all
-          Thread.current[:status] = Status[:crash, :bug, reason: Reason::Bug.new(e)]
+          status = Status[:crash, :bug, reason: Reason::Bug.new(e)]
+          Proc.current.compensate_all(status)
+          Thread.current[:status] = status
           raise
         ensure
           @alive = false
-          Proc.current.cleanup!
+          Proc.current.checkpoint_all
           PRINTER.puts('exiting')
         end
       end
@@ -79,20 +84,29 @@ module Magritte
       alive? && @thread.join
     end
 
-    attr_reader :thread, :env
+    attr_reader :thread
     def initialize(thread, code, env)
       @alive = false
       @thread = thread
       @code = code
       @env = env
-      @compensation_stack = []
+      @stack = []
 
       @interrupt_mutex = LogMutex.new :interrupt
     end
 
+    def env
+      frame.env
+    end
+
+    def frame
+      @stack.last
+    end
+
     def start
       @alive = true
-      open_channels
+      @stack << Frame.new(@env, self)
+      frame.open_channels
       @thread[:magritte_start_mutex].unlock
       self
     end
@@ -122,61 +136,34 @@ module Magritte
       @thread.run
     end
 
-    def cleanup!
-      PRINTER.p :cleanup => self
-      @env.each_input { |c| c.remove_reader(self) }
-      @env.each_output { |c| c.remove_writer(self) }
-    end
-
-    def open_channels
-      @env.each_input { |c| c.add_reader(self) }
-      @env.each_output { |c| c.add_writer(self) }
-    end
-
     def stdout
-      @env.stdout || Channel::Null.new
+      env.stdout || Channel::Null.new
     end
 
     def stdin
-      @env.stdin || Channel::Null.new
+      env.stdin || Channel::Null.new
     end
 
     def add_compensation(comp)
-      @compensation_stack[-1] << comp
+      frame.add_compensation(comp)
     end
 
[…]

### 2018-12-07 doc/paper.tex
paper updates

@@
 
-% \documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
-\documentclass[sigplan,10pt,nosumlimits]{acmart}
+\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
+% \documentclass[sigplan,authorversion,draft]{acmart}
 
@@
 \usepackage{listings}
-
-\title{Pipe-based Programming}
+\lstset{
+language=bash,
+% backgroundcolor=\color[rgb]{0.95, 0.95, 0.95},
+tabsize=2,
+rulecolor=,
+basicstyle=\ttfamily,
+upquote=true,
+% aboveskip={1.5\baselineskip},
+columns=fullflexible,
+% columns=fixed,
+showstringspaces=false,
+extendedchars=true,
+breaklines=true,
+prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
+frame=L,
+showtabs=false,
+showspaces=false,
+showstringspaces=false,
+% identifierstyle=\ttfamily,
+keywordstyle=\color[rgb]{1.0,0,0}\bfseries,
+keywordstyle=[1]\color[rgb]{0,0,0.75}\bfseries,
+keywordstyle=[2]\color[rgb]{0.5,0.0,0.0},
+keywordstyle=[3]\color[rgb]{0.127,0.427,0.514},
+keywordstyle=[4]\color[rgb]{0.4,0.4,0.4},
+commentstyle=\color[rgb]{0.5,0.5,0.5}\itshape,
+stringstyle=\color[rgb]{0.639,0.082,0.082},
+morekeywords={read-lines, take, each},
+deletekeywords={do, read},
+%numbers=left,%
+numbersep=5pt,%
+numberstyle=\tiny\color{gray},%
+emphstyle=\bfseries,%
+breaklines=true,
+breakatwhitespace=true,%
+escapechar=`,
+%numbers=left,
+}
+\lstset{escapeinside={<@}{@>}}
+\definecolor{ttblue}{rgb}{0,0,0.75}
+\title{A Shell-like Model for General Purpose Programming}
 \author{Jeanine Adkisson}
@@
 \email{jneen@prg.is.titech.ac.jp}
+
+\author{Johannes Westlund}
+\affiliation{\institution{TODO}}
+\email{TODO}
+\begin{document}
 
@@
 
-\begin{document}
-
 % thought experiment - parallel fetching
 % find an example of named pipes
+
+% TODO LIST
+% * Find an example of named pipes reopening
+% * Find an example of a non-linear pipe arrangement (prob. from flow-based programming)
+% * Plan the Description of Magritte section
+% * Add Call Semantics to Design Considerations
+% * Institute and email addr from Johannes
+% * Explain why we do not do eager interruption (limit sources of interruptions)
+% * Finish Non-Interruption section
+% * Completely redo the Magritte section
+% * Bibliography
+% * Pretty Syntax
 
@@
 
-We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
-The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
+% We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
+% The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
 
@@
 
+Our overarching goal is to create a language and a programming system that retains the pipe-based programming and interaction model of bash, but allows for large programs in a way that existing shell languages do not.
+
+\section{Design Considerations}
+In this section we discuss several feature requirements and design considerations for a pipe-based language to be viable for large programs.
+
+\paragraph{Not Everything Is a String.}
+The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as values.
+
+Several projects, including bash itself, have attempted to add lists and hash-maps as standard objects.
+Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions normally.
+
+The \verb/es/ shell (TODO: cite) attempts to extend the standard shell model with lambda functions with lexical scoping. However,
+
+\paragraph{Transactional Channels.}
 Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are specified for performance reasons.
@@
 
+Consider the fairly common architecture of producer/consumer: A \emph{producer} process producese values that are processed in parallel by multiple \emph{consumer} processes, and then the values are collected in a single output.
+This might be expressed in bash as:
+
+\begin{lstlisting}
+# pass through the stream, labeling every line
+label() {
+  while read line; do echo "$1$line"; done
+}
+
+# process the stream with different labels
+split() {
+  label a &
+  label b &
+}
+
+# process the numbers 1-100, limiting the
+# output to the first 3 lines
+seq 100 | split | head -3
+\end{lstlisting}
+
+Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved, resulting in outputs such as:
+
+\begin{verbatim}
+a1    b12   a
+a2    a     b12
+a3    a4    a34
+\end{verbatim}
+In a language intended for large programs, therefore, writes to a channel must be \emph{transactional} - a single write must be sent wholesale to a single reader without any interleaving, no matter how big the write.
+
+\paragraph{Interruption.}
 A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used.
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   read-lines tmp/large-file
     | take 10
-    | each (?line => do-expensive-work $line)
-\end{verbatim}
+    | each (?line => do-expensive-work <@\$@>line)
+\end{lstlisting}
+
 In this example, three processes are spawned: (1) a process with a file open that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
@@
 
-\section{Design Considerations}
-\paragraph{Compensation.} In order to observe interruption, we define a system of \emph{compensation actions}, which will be run in the case of an interruption.
+\paragraph{Compensation.} In order to observe interruption, user code must be able to react to and handle interruption.
+The most direct way to observe interruption is through a system of \emph{compensation actions}, which are expressions which will be evaluated in the case of an interruption.
+
 For example, the function \verb/file-lines/ above could be implemented as:
@@
   until (=> eof? $handle) (=> read-until "\n" $handle)
-  checkpoint
 )
@@
 
-In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next \verb/checkpoint/ invocation, or when the process is interrupted.
+In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next checkpoint (the end of the function), or when the process is interrupted.
 In this way, we can ensure that the file is closed when the process exits.
 Compensations are the only way in which interruptions can be directly observed by client code - without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program.
-Thus, the primary task of our system is \emph{to run a process's compensation actions at the appropriate time}.
+Thus, we can define the primary task of our system as \emph{running a process's compensation actions at the appropriate time}.
 
@@
 
-Consider the following example, with the assumption that \verb/do-other-operation/ does not write to the standard output:
+Consider the following example, with the assumption that \\ \verb/do-other-operation/ does not write to the standard output:
 
@@
 
-\paragraph{Channel registration.}
+\paragraph{Non-closing of Channels.}
 
@@
 
+\paragraph{Masking Interruptions.} (TODO: explain \verb/produce/ and \verb/consume/)
+
 \paragraph{Closing Permanence.}
@@
 
-\section{Description of Semantics.}
+\section{Description of Magritte.}
 

### 2018-12-07 lib/magritte/builtins.rb
add size, str, and contains builtins

@@ -183,6 +183,36 @@ module Magritte
       end
     end
 
+    builtin :size, [:any] do |e|
+      s = case e
+      when Value::String
+        e.value.size
+      when Value::Vector
+        e.elems.size
+      when Value::Environment
+        e.env.size
+      else
+        Proc.current.crash!("size: don't know how to size #{e.repr}")
+      end
+
+      put Value::Number.new(s)
+    end
+
+    builtin :str, [], :any do |*a|
+      put Value::String.new(a.map(&:to_s).join)
+    end
+
+    builtin :contains, [:any, :any] do |e, container|
+      case container
+      when Value::Vector
+        bool(container.elems.any? { |el| el == e })
+      when Value::String
+        bool(container.value[e.to_s])
+      else
+        bool(false)
+      end
+    end
+
     # Initialize environment with functions that can be defined in the language itself
     def self.load_lib(lib_name, source, env)
       # Transform the lib string into an ast

### 2018-12-07 lib/magritte/repl.rb
behave better on ^C in the repl

@@ -45,16 +45,15 @@ module Magritte
 
     def evaluate(source)
       ast = Parser.parse(Skeleton.parse(Lexer.new(source_name, source)))
-      status = Proc.spawn(Code.new { Interpret.interpret(ast) }, @env).wait
+      p = Proc.spawn(Code.new { Interpret.interpret(ast) }, @env)
+      status = p.wait
       @streamer.wait_for_close
     rescue CompileError => e
       Status[:fail, reason: Reason::Compile.new(e)]
-    rescue ::Interrupt => e
-      Status[:crash, reason: Reason::Crash.new("interrupted")]
     else
-      @streamer.reset!
-      @input.reset!
       status
+    ensure
+      p && p.crash!("ended")
     end
 
     def process_line
@@ -64,6 +63,12 @@ module Magritte
       status = evaluate(string)
       @mutex.synchronize { puts "% #{status.repr}" }
       !status.fail?
+    rescue ::Interrupt => e
+      puts "^C"
+      false
+    ensure
+      @streamer.reset!
+      @input.reset!
     end
 
     def run

### 2018-12-07 lib/magritte/value.rb
add to_s, defaulting to repr

@@ -10,8 +10,16 @@ module Magritte
     class String < Base
       attr_reader :value
 
+      def to_s
+        @value
+      end
+
       def initialize(value)
         @value = value
+      end
+
+      def ==(other)
+        other.is_a?(self.class) && other.value = @value
       end
 
       def call(args)
@@ -38,6 +46,10 @@ module Magritte
 
       def repr
         value.to_s.gsub(/\.0$/, '')
+      end
+
+      def to_s
+        repr
       end
     end
 

### 2018-12-07 lib/magritte/value.rb
behave better on missing functions in String#call

@@ -26,6 +26,8 @@ module Magritte
         # Semantics: Look up the thing in current env and call it
         # Proc.current.env
         Proc.current.env.get(@value).call(args)
+      rescue Env::MissingVariable => e
+        Proc.current.crash!(e.to_s)
       end
 
       def repr

### 2018-12-07 lib/magritte/value.rb
rm a binding.pry

@@ -117,7 +117,6 @@ module Magritte
         end
 
         if args.size > bindnames.size
-          binding.pry
           Proc.current.crash!("#{repr}: too many arguments: #{args.size} > #{bindnames.size}")
         end
         args.zip(bindnames) do |arg, bind|

### 2018-12-07 spec/interpret_spec.rb
add tests for interruption on redirected channels

@@ -513,5 +513,44 @@ describe Magritte::Interpret do
         assert { result == "crashed" }
       end
     end
+
+    describe "interrupts on redirections" do
+      let(:input) {
+        """
+        c = (make-channel)
+        & put 1 2 3 > $c
+        drain < $c
+        (dr) = (put (get); dr)
+        dr < $c
+
+        # we never get here, because we get
+        # interrupted by the above
+        put 4
+        """
+      }
+
+      it do
+        assert { results == %w(1 2 3) }
+      end
+    end
+
+    describe "get masking" do
+      let(:input) {
+        """
+        c = (make-channel)
+        & put 1 > $c
+        get < $c
+        get < $c
+
+        # we never get here, because we get interrupted
+        # by the above
+        put 4
+        """
+      }
+
+      it do
+        assert { results == %w(1) }
+      end
+    end
   end
 end

### 2018-12-10 doc/paper.tex
paper updates

@@
 
+% TODO: discuss this
+%# fn pipe(lhs, rhs, inp, out) {
+%#   c = channel()
+%#   go fn () { lhs(inp, c); };
+%#   rhs(c, out)
+%# }
+%#
+%# collect = collector()
+%# pipe_collect(curry(range, 10), curry(fan, 2,
+%  fn(el, inp, out) { write(el + 1, out); }
+%), inp);
+%#
+
 Our overarching goal is to create a language and a programming system that retains the pipe-based programming and interaction model of bash, but allows for large programs in a way that existing shell languages do not.
@@
 \paragraph{Not Everything Is a String.}
-The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as values.
+The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.
 
-Several projects, including bash itself, have attempted to add lists and hash-maps as standard objects.
-Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions normally.
+Several projects\cite{haahr}\cite{duff}, including bash itself, have attempted to add lists and hash-maps as standard objects.
+Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes normally.
+Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not available.
 
-The \verb/es/ shell (TODO: cite) attempts to extend the standard shell model with lambda functions with lexical scoping. However,
+Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
 
@@
 
+Luckily, given the decision to allow rich values to be passed through channels, a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data - or for that matter a pointer.
+
 \paragraph{Interruption.}
-A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used.
-Consider the following pseudo-code for a pipe-based language:
+A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 (file-lines ?fname) = (
@@
 )
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-(TODO show race condition with eager-interruption)
+An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted - those points which run a \verb/put/ or a \verb/get/.
+Whereas with eager-interruption semantics any point in the code may be interrupted, introducing the need for users to either mark critical uninterruptible sections or be very careful about setting correct compensations in the case of mutative algorithms.
+
+On the other hand, lazy interruption has the disadvantage that a producing process must do enough work to produce one more value than will be consumed.
+If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in doing a large amount of unnecessary work.
+
+Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for extra values, and we discuss some features that may mitigate this problem in \emph{Future Work}.
 
@@
 
-(TODO explain guiding principle: \emph{A process is interrupted exactly when it can no longer be woken up} and the addendum \emph{without spawning new processes})
+Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel.
+It would not be appropriate in this case to interrupt other processes attached to the channel, as they are still able to communicate.
 
+We define our guiding principle for the appropriate time to interrupt a process as: \emph{A process is interrupted exactly when it can no longer be woken up}.
 When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
-Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations.
+Therefore the appropriate time for it to be interrupted is if it is blocked on a channel that will never receive any more operations.
 In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.
 
-(TODO therefore we add the constraint that processes must only interact with specifically registered channels)
-
-\paragraph{Masking Interruptions.} (TODO: explain \verb/produce/ and \verb/consume/)
+Luckily, in pipe-based languages, the arrangement of pipes is usually specified at process spawn time.
+This enables us to maintain a process register inside of each channel, so that we can decide when all readers or all writers have detached from a channel, at which time the channel closes and interrupts all blocked processes.
 
 \paragraph{Closing Permanence.}
-(TODO: explain why we do not reopen channels)
+Though we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel.
+In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use.
+In UNIX, for example, the named-pipes API allows for persistent channels (TODO: cite!) that, being merely files, can be reopened at any time.
+
+However, this can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:
+
+\begin{lstlisting}
+c = (make-channel)
+& count-forever > $c
+& take 10 < $c
+put 10 > $c
+\end{lstlisting}
+
+This example, while contrived, represents an unavoidable race with first-class channels - whether the process running \verb/take 10/ closes the channel before or
+after the \verb/put 10/ invocation opens the channel for writing.
+If we allow channel reopening, we will either block forever, or insert the number 10 into the stream, depending on which happens first.
+However if we do not allow channel reopening, we can say that, if a process initiates a read or write on a closed channel, it is immediately interrupted.
+In this case, both sides of the former race have the same termination behavior - the process is closed when \verb/take 10/ returns.
+
+\paragraph{Masking Interruptions.}
+
+A user will often wish to continue after a particular output channel has closed, and in this case will need a facility to mask exceptions for particular channels.
+Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
+
+\begin{lstlisting}
+(sample) = (
+  hold = (get)
+  i = 1
+  each (?v =>
+    $i = (inc $i)
+    # mutate the $hold variable with probability 1/$i
+    prob $i 1 && ($hold = %v)
+  )
+
+  put $hold
+)
+\end{lstlisting}
+
+In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability.
+After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.
+
+However, with the semantics described above, the \verb/each/ function will loop until it is given an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line.
+In order to avoid this situation, the \verb/each/ function must somehow indicate that it intends to consume the entire stream, and handle the interrupt signal.
+
+While we plan to explore more general mechanisms for error handling, for cases such as this it suffices to provide two builtins, \verb/produce/ and \verb/consume/, each of which takes a single function to loop forever until the standard output or standard input respectively is consumed, whereby control flow continues after the invocation.
+Using these functions, we might define \verb/each/ as:
+
+\begin{lstlisting}
+(each ?fn) = (consume (=> %fn (get)))
+\end{lstlisting}
+
+With this definition, the call to \verb/consume/ will mask the interruption, and control flow within the \verb/sample/ function will proceed as normal.
 
@@
 
+The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash.
+However its list type is not nestable - nested lists are automatically flattened at creation time.
+Additionally, these lists cannot be passed through channels.
+
+The \verb/es/ shell \cite{haahr} attempts to extend the standard shell model with lambda functions with lexical scoping.
+These shell functions are marshalled into strings when passed through pipes (whether to shell functions or external programs), but are still written byte-by-byte and thus are subject to interleaving in concurrent cases.
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{paper}
+
 \end{document}
-
-%# fn pipe(lhs, rhs, inp, out) {
-%#   c = channel()
-%#   go fn () { lhs(inp, c); };
-%#   rhs(c, out)
-%# }
-%#
-%# collect = collector()
-%# pipe_collect(curry(range, 10), curry(fan, 2,
-%  fn(el, inp, out) { write(el + 1, out); }
-%), inp);
-%#

### 2018-12-10 lib/magritte/code.rb
Replace with_env with enter_frame

@@ -115,7 +115,7 @@ module Magritte
     def call
       if Proc.current?
         env = @env.extend(in_ch, out_ch)
-        Proc.with_env(env, &@block)
+        Proc.enter_frame(env, &@block)
       else
         spawn.wait
       end

### 2018-12-10 lib/magritte/interpret.rb

@@ -122,7 +122,7 @@ module Magritte
 
       def visit_environment(node)
         env = Proc.current.env.extend
-        Proc.with_env(env) do
+        Proc.enter_frame(env) do
           visit_exec(node.body)
         end
         env.unhinge!
@@ -141,7 +141,7 @@ module Magritte
             outputs << c.channel
           end
         end
-        Proc.with_env(Proc.current.env.extend(inputs, outputs)) do
+        Proc.enter_frame(Proc.current.env.extend(inputs, outputs)) do
           visit_exec(node.expr)
         end
       end

### 2018-12-10 lib/magritte/proc.rb

@@ -20,8 +20,8 @@ module Magritte
       current.send(:with_channels, in_ch, out_ch, &b)
     end
 
-    def self.with_env(env, &b)
-      current.send(:with_env, env, &b)
+    def self.enter_frame(*args, &b)
+      current.send(:enter_frame, *args, &b)
     end
 
     def self.spawn(code, env)
@@ -167,11 +167,7 @@ module Magritte
     end
 
   protected
-    def with_channels(new_in, new_out, &b)
-      with_env(@env.extend(new_in, new_out), &b)
-    end
-
-    def with_env(new_env, &b)
+    def enter_frame(*args, &b)
       stack_size = @stack.size
 
       @interrupt_mutex.synchronize do
@@ -182,7 +178,7 @@ module Magritte
 
       frame.open_channels
 
-      yield
+      out = yield
     rescue Interrupt => e
       compensate(e)
       PRINTER.p :interrupt => @stack
@@ -191,6 +187,7 @@ module Magritte
       binding.pry if stack_size+1 != @stack.size
       checkpoint
       PRINTER.p :checkpoint => @stack
+      out
     end
   end
 end

### 2018-12-10 lib/magritte/value.rb

@@ -122,7 +122,7 @@ module Magritte
         args.zip(bindnames) do |arg, bind|
           env.let(bind, arg)
         end
-        Proc.with_env(env) { Interpret.interpret(@expr[0]) }
+        Proc.enter_frame(env) { Interpret.interpret(@expr[0]) }
       end
 
       def repr

### 2018-12-10 lib/magritte/frame.rb
Rearrange constructor parameters for Frame class
Also add lex_env parameter to Frame instantiation but currently not used

@@ -1,8 +1,9 @@
 module Magritte
   class Frame
-    attr_reader :env
-    def initialize(env, p)
+    attr_reader :env, :lex_env
+    def initialize(p, env, lex_env = nil)
       @env = env
+      #@lex_env = lex_env || p.lex_env
       @proc = p
       @compensations = []
     end

### 2018-12-10 lib/magritte/proc.rb

@@ -105,7 +105,7 @@ module Magritte
 
     def start
       @alive = true
-      @stack << Frame.new(@env, self)
+      @stack << Frame.new(self, @env)
       frame.open_channels
       @thread[:magritte_start_mutex].unlock
       self
@@ -171,7 +171,7 @@ module Magritte
       stack_size = @stack.size
 
       @interrupt_mutex.synchronize do
-        frame = Frame.new(new_env, self)
+        frame = Frame.new(self, *args)
         PRINTER.p :stack => @stack
         @stack << frame
       end

### 2018-12-10 lib/magritte/frame.rb
Remove logging when channels are unregistered

@@ -41,7 +41,6 @@ module Magritte
 
   private
     def unregister_channels
-      PRINTER.p :unregister_channels => [@env.own_inputs, @env.own_outputs]
       @env.each_input { |c| c.remove_reader(self) }
       @env.each_output { |c| c.remove_writer(self) }
     end

### 2018-12-10 lib/magritte/proc.rb
Return status in spawned thread when interrupted

@@ -37,6 +37,7 @@ module Magritte
             code.run
           rescue Interrupt => e
             Proc.current.compensate(e)
+            e.status
           end
         rescue Exception => e
           PRINTER.p :exception

### 2018-12-11 lib/magritte.rb
Add pattern.rb file containing Evaluator class

@@ -33,6 +33,7 @@ module Magritte
   load "#{LIB_DIR}/magritte/channel.rb"
   load "#{LIB_DIR}/magritte/streamer.rb"
   load "#{LIB_DIR}/magritte/proc.rb"
+  load "#{LIB_DIR}/magritte/pattern.rb"
   load "#{LIB_DIR}/magritte/value.rb"
   load "#{LIB_DIR}/magritte/env.rb"
   load "#{LIB_DIR}/magritte/builtins.rb"

### 2018-12-11 lib/magritte/pattern.rb

@@ -1,0 +1,54 @@
+module Magritte
+  module Pattern
+
+    def self.evaluate(node, val, env)
+      Evaluator.new(env.extend).evaluate(node, val)
+    end
+
+    class Evaluator < Tree::Visitor
+
+      class PatternFail < StandardError
+      end
+
+      def initialize(env)
+        @env = env
+      end
+
+      def evaluate(node, val)
+        visit(node, val)
+      rescue PatternFail
+        nil
+      else
+        @env
+      end
+
+      def visit_default(node, val)
+        raise "TODO"
+      end
+
+      def visit_binder(node, val)
+        @env.let(node.name, val)
+      end
+
+      def visit_vector_pattern(node, val)
+        fail! unless val.is_a?(Value::Vector)
+
+        patterns = node.patterns
+        values = val.elems
+        # Check that the size is right
+        fail! unless node.rest.nil? ? patterns.size == values.size : patterns.size <= values.size
+        patterns.each_with_index do |pat, i|
+          visit(pat, values[i])
+        end
+
+        if node.rest
+          @env.let(node.rest.name, Value::Vector.new(values[patterns.size..-1]))
+        end
+      end
+
+      def fail!
+        raise PatternFail.new
+      end
+    end
+  end
+end

### 2018-12-11 lib/magritte/tree.rb
Define defopt method to represent optional data

@@ -125,6 +125,10 @@ module Magritte
           defattr(name, ListRecAttr)
         end
 
+        def defopt(name)
+          defattr(name, OptAttr)
+        end
+
         def make(*attrs)
           if attrs.size != types.size
             raise ArgumentError.new("Expected #{types.size} arguments, got #{attrs.size}")

### 2018-12-11 spec/parser_spec.rb
Remove too big tests from parser spec

@@ -33,57 +33,6 @@ describe Magritte::Parser do
     end
   end
 
-  describe "variables" do
-    let(:input) {
-      """
-      $x = 1
-      %n = 2
-      """
-    }
-
-    #it "parses correctly" do
-    #  assert { ast.elems.size == 2 }
-    #  assert { ast.elems.all?(Magritte::AST::Assignment) }
-    #  #assert { ast.elems[0].lhs.inspect == "#<Magritte::AST::Variable[\"x\"]>" }
-    #  assert { ast.elems[0].rhs.inspect == "#<Magritte::AST::Number[\"1\"]>" }
-    #  assert { ast.elems[1].lhs.inspect == "#<Magritte::AST::LexVariable[\"n\"]>" }
-    #  assert { ast.elems[1].rhs.inspect == "#<Magritte::AST::Number[\"2\"]>" }
-    #end
-  end
-
-  describe "command" do
-    let(:input) {
-      """
-      multi-arg-command arg1 arg2 >$out <$in
-      """
-    }
-
-    #it "parses correctly" do
-    #  assert { ast.elems.size == 1 }
-    #  assert { ast.elems.first.is_a?(Magritte::AST::Command) }
-    #  assert { ast.elems.first.vec.inspect == "[#<Magritte::AST::String[\"multi-arg-command\"]>, #<Magritte::AST::String[\"arg1\"]>, #<Magritte::AST::String[\"arg2\"]>]" }
-    #  assert { ast.elems.first.redirects.inspect == "[#<Magritte::AST::Redirect[:<, *#<Magritte::AST::Variable[\"in\"]>]>, #<Magritte::AST::Redirect[:>, *#<Magritte::AST::Variable[\"out\"]>]>]" }
-    #end
-  end
-
-  describe "lambda" do
-    let(:input) {
-      """
-      f = (?x ?y => add x y)
-      """
-    }
-
-    #it "parses correctly" do
-    #  assert { ast.elems.size == 1 }
-    #  assert { ast.elems[0].is_a?(Magritte::AST::Assignment) }
-    #  #assert { ast.elems[0].lhs.inspect == "#<Magritte::AST::String[\"f\"]>" }
-    #  assert { ast.elems[0].rhs.is_a?(Magritte::AST::Lambda) }
-    #  assert { ast.elems[0].rhs.name == "anon@test@1:11~1:29" }
[…]

### 2018-12-11 lib/magritte/ast.rb
Start on pattern matching and add VectorPatterns
Add a VectorPattern AST node to hold lambda function patterns.
This allows for some types of pattern matching but also introduce
modifications to several methods. Most notably how
free variables are detected and how lambda functions are called.

Parsing of anon lambda bodies streaching multiple lines also
work as intended. However, it was not possible to separate
it into it's own commit.

@@ -21,10 +21,20 @@ module Magritte
       defdata :value
     end
 
+    class VectorPattern < Tree::Node
+      deflistrec :patterns
+      defopt :rest
+    end
+
     class Lambda < Tree::Node
       defdata :name
       deflistrec :patterns
       deflistrec :bodies
+
+      def initialize(*)
+        super
+        raise "Pattern and body mismatch" unless patterns.size == bodies.size
+      end
     end
 
     class Pipe < Tree::Node

### 2018-12-11 lib/magritte/free_vars.rb

@@ -4,17 +4,25 @@ module Magritte
       Scanner.new.collect(node, Set.new)
     end
 
+    class BinderScanner < Tree::Collector
+      def visit_binder(node)
+        Set.new([node.name])
+      end
+    end
+
     class Scanner < Tree::Collector
       def visit_lex_variable(node, bound_vars)
         Set.new([node.name])
       end
 
       def visit_lambda(node, bound_vars)
-        unless node.patterns.all? { |n| n.is_a? AST::Binder }
-          raise "Not implemented yet!"
+        out = Set.new
+        node.patterns.zip(node.bodies) do |pat, body|
+          binders = BinderScanner.new.collect_one(pat)
+          out.merge(shadow(body, bound_vars, binders))
         end
-        to_bind = node.patterns.map(&:name)
-        shadow(node.bodies, bound_vars, to_bind)
+        #p :lambda_free => [node.name, out]
+        out
       end
 
       def visit_subst(node, bound_vars)

### 2018-12-11 lib/magritte/interpret.rb

@@ -87,7 +87,7 @@ module Magritte
 
       def visit_lambda(node)
         free_vars = @free_vars[node]
-        yield Value::Function.new(node.name, Proc.current.env.slice(free_vars), node.patterns.map(&:name), node.bodies)
+        yield Value::Function.new(node.name, Proc.current.env.slice(free_vars), node.patterns, node.bodies)
       end
 
       def visit_assignment(node)
@@ -149,7 +149,7 @@ module Magritte
       def visit_compensation(node)
         visit_exec(node.expr)
         name = "compensation@"
-        func = Value::Function.new(name, Env.empty, [], [node.compensation])
+        func = Value::Function.new(name, Env.empty, [AST::VectorPattern[[], nil]], [node.compensation])
         Proc.current.add_compensation(Value::Compensation.new(func, node.unconditional))
       end
 

### 2018-12-11 lib/magritte/parser.rb

@@ -118,7 +118,7 @@ module Magritte
         error!(term, "TODO: Support patterns")
       end
       patterns = bindings.elems.map { |e| AST::Binder[e.value] }
-      return AST::Lambda[name, patterns, [parse_root(bodies)]]
+      return AST::Lambda[name, [AST::VectorPattern[patterns, nil]], [parse_root(bodies)]]
     end
 
     def parse_command(command)
@@ -210,15 +210,26 @@ module Magritte
         end
 
         # Anon lambda spanning multiple lines
-        item.elems.first.match(lsplit(~_, token(:arrow), ~_)) do |bindings, body|
+        item.elems.first.match(lsplit(_, token(:arrow), _)) do
+          bindings = []
+          bodies = []
           name = "anon@#{term.range.repr}"
-          bodies = item.elems.drop(1).unshift(body)
+          item.elems.each do |elem|
+            tmp = elem
+            elem.match(lsplit(~_, token(:arrow), ~_)) do |patterns, body|
+              bindings << patterns
+              bodies << []
+              tmp = body
+            end
+            bodies.last << tmp
+          end
 
-          unless bindings.elems.all? { |e| e.token?(:bind) }
+          unless bindings.all? { |i| i.elems.all? { |e| e.token?(:bind) }}
             error!(term, "TODO: Support patterns")
           end
-          patterns = bindings.elems.map { |e| AST::Binder[e.value] }
-          return AST::Lambda[name, patterns, [AST::Block[bodies.map { |body| parse_line(body) }]]]
+
+          patterns = bindings.map { |i| AST::VectorPattern[i.elems.map { |e| AST::Binder[e.value] }, nil]}
+          return AST::Lambda[name, patterns, bodies.map { |body| AST::Block[body.map { |line| parse_line(line) }]}]
         end
 
         return AST::Subst[item.sub_items.map { |e| parse_line(e) }]

### 2018-12-11 lib/magritte/tree.rb

@@ -215,6 +215,10 @@ module Magritte
         collection
       end
 
+      def collect_one(node, *args)
+        collect(node, *args)[node]
+      end
+
     protected
       def collect_from(enum, *args)
         out = Set.new

### 2018-12-11 lib/magritte/value.rb

@@ -100,29 +100,33 @@ module Magritte
     class Function < Base
       attr_reader :name
       attr_reader :env
-      attr_reader :bindnames
+      attr_reader :patterns
       attr_reader :expr
 
-      def initialize(name, env, bindnames, expr)
+      def initialize(name, env, patterns, expr)
         @name = name
         @env = env
-        @bindnames = bindnames
+        @patterns = patterns
         @expr = expr
+
+        if @patterns.size != @expr.size
+          raise "malformed function (#{@patterns.size} patterns, #{@expr.size} bodies"
+        end
       end
 
       def call(args)
         env = @env.splice(Proc.current.env)
-        if args.size < bindnames.size
-          Proc.current.crash!("#{repr}: not enough arguments: #{args.size} < #{bindnames.size}")
-        end
+        bound, match_index = match(args, env)
+        Proc.enter_frame(bound) { Interpret.interpret(@expr[match_index]) }
+      end
 
-        if args.size > bindnames.size
-          Proc.current.crash!("#{repr}: too many arguments: #{args.size} > #{bindnames.size}")
+      def match(args, env)
+        args = Vector.new(args)
+        @patterns.each_with_index do |pat, i|
+          bound = Pattern.evaluate(pat, args, env)
+          return [bound, i] if bound
         end
-        args.zip(bindnames) do |arg, bind|
-          env.let(bind, arg)
-        end
-        Proc.enter_frame(env) { Interpret.interpret(@expr[0]) }
+        Proc.current.crash!("Pattern match failed on #{self.repr} for #{args.repr}")
       end
 
       def repr

### 2018-12-11 spec/interpret_spec.rb

@@ -86,14 +86,66 @@ describe Magritte::Interpret do
     end
 
     describe "lambdas" do
-      let(:input) {
-        """
-        (?x => put $x) 1
-        """
-      }
+      describe "call" do
+        let(:input) {
+          """
+          (?x => put $x) 1
+          """
+        }
 
-      it do
-        assert { result == "1" }
+        it do
+          assert { result == "1" }
+        end
+      end
+
+      describe "zero argument lambdas" do
+        let(:input) {
+          """
+          exec (=> put 1)
+          """
+        }
+
+        it do
+          assert { result == "1" }
+        end
+      end
+
+      describe "argument length patterns" do
+        describe "reducing" do
+          let(:input) {
+            """
+            f = (
+              ?x ?y => put [two %x %y]
+              ?x => put [one %x]
+            )
+
+            f 3
+            f 5 6
+            """
+          }
+
+          it do
+            assert { results == ["[one 3]", "[two 5 6]"] }
+          end
+        end
+
+        describe "increasing" do
+          let(:input) {
+            """
+            f = (
+              ?x => put [one %x]
+              ?x ?y => put [two %x %y]
+            )
+
+            f 3
+            f 5 6
+            """
+          }
+
+          it do
+            assert { results == ["[one 3]", "[two 5 6]"] }
+          end
+        end
       end
     end
 

### 2018-12-12 doc/paper.tex
paper updates

@@
 % notes
-
-\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
+% \documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
+\documentclass[english,preprint,JIP]{ipsj}
 % \documentclass[sigplan,authorversion,draft]{acmart}
 
-% don't print the conference ref for now
-% \settopmatter{printacmref=false}
+% \usepackage[utf8]{inputenc}
 
+\usepackage{textcomp}
 \usepackage{comment}
@@
 \lstset{
-language=bash,
+language=ruby,
 % backgroundcolor=\color[rgb]{0.95, 0.95, 0.95},
@@
 prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
-frame=L,
+frame=shadowbox,
 showtabs=false,
@@
 stringstyle=\color[rgb]{0.639,0.082,0.082},
-morekeywords={read-lines, take, each},
+morekeywords={read-lines, take, each, produce, consume, put, get, make-channel},
 deletekeywords={do, read},
@@
 \definecolor{ttblue}{rgb}{0,0,0.75}
+
+\usepackage[varg]{txfonts}%%!!
+\makeatletter%
+\input{ot1txtt.fd}
+\makeatother%
+
+\begin{document}
 \title{A Shell-like Model for General Purpose Programming}
-\author{Jeanine Adkisson}
-\affiliation{\institution{Tokyo Institute of Technology}}
-\email{jneen@prg.is.titech.ac.jp}
 
-\author{Johannes Westlund}
-\affiliation{\institution{TODO}}
-\email{TODO}
-\begin{document}
+\affiliate{TITech}{東京工業大学情報理工学院数理・計算科学系 \\
+Department of Mathematical and Computing Science, Tokyo Institute of Technology,
+Meguro, Tokyo 152-8552}
+\affiliate{KTH}{Kungliga Tekniska H\"ogskolan\\
+KTH Royal Institute of Technology, Stockholm, Sweden, 10044}
 
-\author{Hidehiko Masuhara}
-\affiliation{\institution{Tokyo Institute of Technology}}
-\email{masuhara@prg.is.titech.ac.jp}
+\author{Jeanine Miller Adkisson}{TITech}[jneen@prg.is.titech.ac.jp]
+\author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]
+\author{Johannes Westlund}{TITech,KTH}[jwestlun@kth.se]
 
-% thought experiment - parallel fetching
+% thought experiment---parallel fetching
 % find an example of named pipes
@@
 
-However, existing shell languages are not sufficient to serve as general-purpose languages - values are not observable except in raw streams of bytes, and they lack modern language features such as lexical scope and higher-order functions.
+However, existing shell languages are not sufficient to serve as general-purpose languages---values are not observable except in raw streams of bytes, and they lack modern language features such as lexical scope and higher-order functions.
 
@@
 % We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
-% The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
+% The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes.
 
@@
 \end{abstract}
+
 \maketitle
@@
 
-The UNIX shell programming model has been shown to be a successful model for composing programs - spawning unrelated programs in parallel to communicate over operating-system pipes.
+The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes.
 Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
@@
 In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes in which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.
-
-% TODO: discuss this
-%# fn pipe(lhs, rhs, inp, out) {
-%#   c = channel()
-%#   go fn () { lhs(inp, c); };
-%#   rhs(c, out)
-%# }
-%#
-%# collect = collector()
-%# pipe_collect(curry(range, 10), curry(fan, 2,
-%  fn(el, inp, out) { write(el + 1, out); }
-%), inp);
-%#
 
@@
 
-Consider the fairly common architecture of producer/consumer: A \emph{producer} process producese values that are processed in parallel by multiple \emph{consumer} processes, and then the values are collected in a single output.
+Consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and then the values are collected in a single output.
 This might be expressed in bash as:
-
 \begin{lstlisting}
@@
 \end{verbatim}
-In a language intended for large programs, therefore, writes to a channel must be \emph{transactional} - a single write must be sent wholesale to a single reader without any interleaving, no matter how big the write.
+In a language intended for large programs, therefore, writes to a channel must be \emph{transactional}---a single write must be sent wholesale to a single reader without any interleaving, no matter how big the write.
 
-Luckily, given the decision to allow rich values to be passed through channels, a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data - or for that matter a pointer.
+Luckily, given the decision to allow rich values to be passed through channels, a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data---or for that matter a pointer.
+
+\paragraph{Return Semantics.}
+Pipeline elements such as these do not generally return values---instead, they output data to their output stream.
+However, for serious programs it is desirable to capture some kind of return value from functions.
+ES Shell \cite{haahr} accomplishes this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.
+
+If we allow values to be written to output streams, however, this approach is redundant---a function author must decide whether to return a value or write it to standard output.
+Thus it is desirable to have an interpolation syntax which \emph{collects} the output of a command synchronously.
+
+A priori, such a collection mechanism would have to return an array, as any function call may output zero or more values.
+However, that approach would lead to the callers of functions needing to manually unwrap arrays on every function call.
+In order to avoid this, we \emph{expand} the return values into the command vector---increasing the argument number by the number of values output from the function.
+
+In Magritte, normal parentheses in argument position will collect and expand output values in exactly this manner:
+\begin{lstlisting}
+# a function defintion: output three values
+(count-three) = (put 1; put 2; put 3)
+
+# a function call, using the above function
+other-fn 0 (count-three) 4
+
+# expands to
+other-fn 0 1 2 3 4
+\end{lstlisting}
+
+In this way, we can use output values in exactly the same way as return values seamlessly.
+These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that because we write transactional values to the output, we have the ability to properly separate values without relying on whitespace.
+
+\paragraph{Variable Scope.}
+Like the Es shell\cite{haahr}, we also wish to provide lambda functions with lexically scoped variables.
+However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on runtime information.
+Additionally, since Es has no syntactic difference between let-binding and mutation, even code with no free variables may change in meaning signifcantly depending on what variables are in scope in its context.
+
+To remedy this issue, we introduce a separate syntax for lexically scoped variables, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
+Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable (e.g.
+\verb/x = 1/), so that we can maintain compatibility with standard environment files.
+Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
+\verb/?x/) to allow for a planned extension for pattern matching.
+For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
+\verb/%x = 1/ or \verb/$x = 1/.
+In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
+
+\paragraph{Objects.}
+For large programs, an object system is desirable as a way to provide user defined types and abstractions.
+The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer.
+If we allow environments as first-class objects, they can be used in a straightforward manner as prototype-based objects.
+
+A future planned extension would allow an environment to have multiple prototype pointers, with lookup defined in a depth-first manner.
 
@@
 A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
-
 \begin{lstlisting}
@@
 For example, the function \verb/file-lines/ above could be implemented as:
-
 \begin{lstlisting}
@@
 In this way, we can ensure that the file is closed when the process exits.
-Compensations are the only way in which interruptions can be directly observed by client code - without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program.
+Compensations are the only way in which interruptions can be directly observed by client code---without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program.
 Thus, we can define the primary task of our system as \emph{running a process's compensation actions at the appropriate time}.
@@
 
-An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted - those points which run a \verb/put/ or a \verb/get/.
+An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted---those points which run a \verb/put/ or a \verb/get/.
 Whereas with eager-interruption semantics any point in the code may be interrupted, introducing the need for users to either mark critical uninterruptible sections or be very careful about setting correct compensations in the case of mutative algorithms.
@@
 However, this can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:
-
 \begin{lstlisting}
@@
 
-This example, while contrived, represents an unavoidable race with first-class channels - whether the process running \verb/take 10/ closes the channel before or
+This example, while contrived, represents an unavoidable race with first-class channels---whether the process running \verb/take 10/ closes the channel before or
 after the \verb/put 10/ invocation opens the channel for writing.
@@
 However if we do not allow channel reopening, we can say that, if a process initiates a read or write on a closed channel, it is immediately interrupted.
-In this case, both sides of the former race have the same termination behavior - the process is closed when \verb/take 10/ returns.
+In this case, both sides of the former race have the same termination behavior---the process is closed when \verb/take 10/ returns.
 
@@
 Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
-
 \begin{lstlisting}
@@
 
-However, with the semantics described above, the \verb/each/ function will loop until it is given an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line.
+However, with the semantics described above, the \verb/each/ function will loop until it receives an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line.
 In order to avoid this situation, the \verb/each/ function must somehow indicate that it intends to consume the entire stream, and handle the interrupt signal.
 
-While we plan to explore more general mechanisms for error handling, for cases such as this it suffices to provide two builtins, \verb/produce/ and \verb/consume/, each of which takes a single function to loop forever until the standard output or standard input respectively is consumed, whereby control flow continues after the invocation.
+While we plan to explore more general mechanisms for error handling, for cases such a this it suffices to provide two builtins, \verb/produce/ and \verb/consume/, each of which takes a single function to loop forever until the standard output or standard input respectively is consumed, whereby control flow continues after the invocation.
 Using these functions, we might define \verb/each/ as:
-
 \begin{lstlisting}
@@
 
-\section{Description of Magritte.}
+\section{Implementation of Magritte}
 
-(TODO: \verb/{add,remove}_{reader,writer}/ are hidden intrinsics, only called automatically at process spawn and exit (even for exits that are in error).
-and we only allow reads/writes for processes *attached* to channels, rather than freely through channel variables.
-explain this)
+The current version of Magritte is implemented as a straightforward interpreter written in Ruby, using Ruby's builtin threads, exceptions, and mutexes to implement processes and channels.
 
-We extend the notion of a synchronous channel with \emph{registered readers} and \emph{registered writers}, and add four intrinsic functions: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, each receiving a channel and a process handle.
-These store and delete processes from the channel's internal sets of registered readers and registered writers.
+\paragraph{Channel Registry.}
 
-In this way, we can know relatively certainly when a channel will never unblock - if either the read or the write end has no remaining processes holding it open, we immediately close it.
-While it is still possible for processes to attempt to attach new processes to a closed channel
+Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described above.
+When \verb/remove_reader/ or \verb/remove_writer/ result in an empty registered set, they additionally close the channel and raise an internal \verb/Interrupt/ exception in every blocked thread.
 
-Additionally, \verb/remove_reader/ and \verb/remove_writer/ will cause the channel to close if their corresponding sets are empty after the operation.
+Once the channel is closed, every call to \verb/read/ and \verb/write/ will cause the calling process to be interrupted, as described above in \emph{Closing Permanence}.
 
-When a channel is closed, we first interrupt every process that is blocked on the channel.
-Thereafter, any attempted read or write operations will cause the initiating process to be interrupted.
-When a channel is interrupted, we must not only perform any cleanup actions such as closing files or writing to error logs, but also call \verb/remove_reader/ and \verb/remove_writer/ on any channels the process has registered - thereby potentially cascading the interruptions to other processes.
+In order to reduce unnecessary use of Ruby threads, we also find it is simpler to, instead of registering \emph{processes} to the channels, to register \emph{frames}.
+In this way, we can register all inputs and outputs on frame entry, and use standard Ruby exception handling to ensure we properly unregister all inputs and outputs on frame exit.
+This means that different frames in the same process can be redirected to different channels, which makes the \verb/>/ and \verb/</ redirection syntax straightforward to implement---we simply push a new frame with different channels attached.
 
-\section{Magritte}
+Cascading of interruptions then happens naturally---when a channel closes, a process is interrupted, causing it to unwind its stack and deregister channels, thereby potentially causing other channels to close.
 
-We have implemented \verb/magritte/(TODO: cite), a Ruby DSL that allows communication between native Ruby threads, and automatically cleans up dangling threads with the above semantics.
-The API revolves around a \verb/Magritte::Code/ object, which can also be created with the builtin \verb/s/ primitive.
+\paragraph{Spawning Order Dependency.}
 
-Within a code object, the primitive functions \verb/put/ and \verb/get/ can be used to read from the input and write to the output.
-A code object can be run in-process (\verb/.call/), in the background (\verb/.go/).
-The default output is a special kind of write-only channel called a \verb/Collector/, which never blocks, but collects every written value into an array.
+It is necessary to take some care with the implementation of the \verb/spawn/ primitive, that we wait until the spawned process has finished registering its channels before the spawning process continues.
+Consider the example, which outputs 10 numbers, possibly out of order:
+\begin{lstlisting}
+(drain) = (each (?x => put %x))
+count-forever | (& drain; & drain) | take 10
+\end{lstlisting}
+The middle process is responsible for spawning two processes that funnel data from their input to their output.
+However, since they are both spawned in the background, the spawning process will immediately return.
+In general this is not a problem, since the \verb/drain/ processes should be keeping the two pipes open.
+However, if we do not take care to wait until they have finished registering their channels, there is a risk that the spawning process will return first and close the two pipes.
 
-Input and output channels can be set \emph{outside} of a channel using the \verb/.from(channel)/ and \verb/.to(channel)/ methods.
-However, they are by default inherited from the enclosing context, meaning that it is straightforward to spawn multiple concurrent processes that are attached to the same channels.
+For this reason, we have special case handling for the root frame---the channels are registered in the \emph{spawning} process, before the spawned process is allowed to start.
 
-Consider the example:
+\paragraph{Collectors.}
+In order to implement the call semantics described above, we also implement a write-only channel called a \emph{collector}.
+Collectors cannot be created directly by users, but only appear in the interpreter when we evaluate a \verb/Subst/ node, which represents parentheses in argument position.
 
-\begin{verbatim}
-s { i = 0; loop { put i; i += 1 } }
-  .p { s { drain }.go; s { drain }.go }
-  .p { take 10 }.call
-\end{verbatim}
-In this example, we spawn four processes.
-The first line spawns an infinite loop that generates all positive integers.
-The second line spawns two processes, both reading from the same input and writing to the same output.
-Finally the third line spawns a process that reads and writes 10 values, then returns. The definition of \verb/drain/ is very simply:
+Naively, collectors would ignore registration commands and simply append written elements to an array. However, it is still necessary to track registered writers. Consider the example:
+\begin{lstlisting}
+(range ?n) = (count-forever | take %n)
+my-list = [(range 10 | (& drain; & drain))]
+\end{lstlisting}
 
-\begin{verbatim}
-def drain
-  loop { put(get) }
-end
-\end{verbatim}
-so that it completely drains the input channel into the output.
-But since there are two of these processes running concurrently, races are possible.
+In this example, there is an open question of how long we should wait until reading the collection and continuing.
+If we naively wait until the base command is finished, we will continue early and miss values that may be written later.
+Thus substitution waits until all writers to the collector have deregistered.
 
-According to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed.
-Notably, even though the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning, this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.
-
-% \verb{...} inline \verbatim
-% \verb+abc_def+
-
-% example of bash race conditions based on re-opening channels
+To implement this, we add a \verb/wait_for_close/ intrinsic to collectors, which returns immediately if closed, and otherwise adds the current thread to a waiting set and sleeps.
+Upon closing the channel in \verb/remove_writer/, all elements of this waiting set are awoken, and flow continues.
+Thus the algorithm for substiution is:
+\begin{lstlisting}
+* Create a new collector $c
+* Evaluate the contents with standard output into the collector
+* Run $c.wait_for_close
+\end{lstlisting}
 
@@
 
+\paragraph{Iterators.}
+(TODO: iterators in other languages?)
 PEP 533 Authors(TODO: cite properly) describe an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
-But channel-based systems generally allow for multiple concurrent readers and writers - so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
-We propose a system such that processes are interrupted as soon as it is no longer possible for them to resume.
+But channel-based systems generally allow for multiple concurrent readers and writers---so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
 
-The Bourne Again Shell (\"bash\")
+\paragraph{Go Language.}
+
+The Go programming language contains a robust channel implementation, which can be used to implement a wide variety of concurrent algorithms.
+In an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
+\begin{lstlisting}[language=C, morekeywords={fn,channel,nil,go}]
+fn pipe(lhs, rhs, inp, out) {
+  c = channel()
+  go fn () { lhs(inp, c); };
+  rhs(c, out)
+}
+
+collect = collector()
+pipe(curry(range, 10), curry(fan, 2, fn(el, inp, out) { write(el + 1, out); }), nil, collect);
+
+return collect.values()
+\end{lstlisting}
+
+However, such a system would still require manual use of input and output channels when a simple pipeline would suffice---and the user must still translate between returning vs.
+outputting functions and expressions.
+Additionally, Go does not contain any kind of channel closing semantics or automatic process cleanup.
+Finally, such a system would be far too heavyweight for a REPL context, and would not make for a viable shell.
 
@@
 
+\paragraph{RC Shell.}
 The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash.
-However its list type is not nestable - nested lists are automatically flattened at creation time.
+However its list type is not nestable---nested lists are automatically flattened at creation time.
 Additionally, these lists cannot be passed through channels.
 
+\paragraph{Es Shell.}
 The \verb/es/ shell \cite{haahr} attempts to extend the standard shell model with lambda functions with lexical scoping.
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{ipsjsort-e}
 \bibliography{paper}

### 2018-12-12 spec/interpret_spec.rb
failing test for block scope

@@ -247,6 +247,21 @@ describe Magritte::Interpret do
           assert { result == "103" }
         end
       end
+
+      describe "block scoping" do
+        let(:input) {
+          """
+          x = 1
+          put (x = 2; put $x)
+          (x = 3; put $x)
+          put $x
+          """
+        }
+
+        it do
+          assert { results == %w(2 3 1) }
+        end
+      end
     end
 
     describe "root-level blocks" do

### 2018-12-13 doc/paper.tex
paper updates

@@
 % notes
-% \documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}
-\documentclass[english,preprint,JIP]{ipsj}
-% \documentclass[sigplan,authorversion,draft]{acmart}
+\documentclass[english,preprint,JIP,uplatex]{ipsj}
 
@@
 
-\affiliate{TITech}{東京工業大学情報理工学院数理・計算科学系 \\
+\affiliate{TITech}{%東京工業大学情報理工学院数理・計算科学系 \\
 Department of Mathematical and Computing Science, Tokyo Institute of Technology,
@@
 
-\author{Jeanine Miller Adkisson}{TITech}[jneen@prg.is.titech.ac.jp]
+\author{Jeanine Miller Adkisson}{TITech}[jneen@jneen.net]
 \author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]
@@
 % TODO LIST
-% * Find an example of named pipes reopening
+% x Find an example of named pipes reopening
 % * Find an example of a non-linear pipe arrangement (prob. from flow-based programming)
-% * Plan the Description of Magritte section
-% * Add Call Semantics to Design Considerations
-% * Institute and email addr from Johannes
-% * Explain why we do not do eager interruption (limit sources of interruptions)
-% * Finish Non-Interruption section
-% * Completely redo the Magritte section
 % * Bibliography
@@
 We also explore concurrency semantics for such a language.
-
-% We explore the implementation and design challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms.
-% The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes.
-
-% Beyond the most simple tasks, however, \texttt{bash} and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
-% %
-% We explore a pipe-based programming model that attempts to synthesize the good parts of UNIX shells, departing at key points, along with a supporting concurrency model and a semantic of interruption.
-%
-%
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Functional and object-oriented programming suffer from a disconnect between the concept of \emph{functions} and \emph{programs}.
-Whereas functions and methods receive values as arguments and return single values as a result, programs can receive data not only as arguments but also asynchronously through an input, and write results asynchronously to an output.
-%
-% In functional and object-oriented languages, this can result in not only a significant amount of abstraction overhead for dealing with external functions, but the immediacy of function arguments and return values can make composing concurrent programs difficult.
-%
-% By way of a new programming language, \emph{Magritte}, we propose a programming model in which the primary unit of composition is not immediate function arguments and return values, but instead processes with asynchronous inputs and outputs, which can be read from or written to at any time, and which can be chained together via a pipe operator, much like programs in a UNIX environment.
-We also explore a concurrency semantic for such a language.
-%
-% We explore many implementation and programming challenges that come with such a model, and also show that this model is able to easily create and compose concurrent algorithms that would be difficult to write or compose in a functional or object-oriented language.
-
-% Despite the widespread use of GUI tools for configuration and orchestration tasks, CLI tools, and more specifically shell-based interfaces remain a powerful standard for software distribution and integration.
-Several attempts have been made at solving the desktop-integration problem: how should we allow programs written in different languages and on different platforms to be scripted and integrated in a coherent system?
-The most notable attempts at this problem have been TCL(TODO: cite!) and Guile Scheme(TODO: cite!), but neither has been as succesful as they intended to be: the common language for a large portion of tools remains shell-based, relying on such concepts as standard I/O, string-based argument vectors, and the expectation that users will be able to freely pipe streams of bytes in a pipe-based way.
-%
-% Unfortunately, existing UNIX shell languages have significant known interoperability issues, such as lack of any kind of usable encapsulation, and the lack of in-language data structures to represent any kind of value other than a string.
-We believe that a shell-compatible language with well-defined semantics that could also function as a general-purpose systems language could go a long way towards solving the desktop-integration problem.
-%
-% On the other hand, a persistent problem with more established systems like Erlang or Go is that they don't account for closing channels, and require manual cleanup of processes.
-In a systems programming context, errors and hangups are commonplace and expected, and it is critical to have some measure of automatic process cleanup for even naively-written code.
-%
-% We propose a channel-based communication system that integrates with a workflow compensation system and is capable of recovering from errors and, importantly, cleaning up processes when communication halts.
 \end{abstract}
@@
 
-\section{Motivation}
+\section{Motivation}\label{motivation}
 
@@
 
-\section{Design Considerations}
+\subsubsection*{Outline}\noindent
+In this work, we present a new shell language \emph{Magritte}.
+Section 2 describes design requirements for a language in this space.
+Section 3 and 4 describe Magritte's design and implementation.
+Section 5 describes our future goals with this project, and in Section 6 we review various related works and alternative approaches.
+
+\section{Design Considerations}\label{design-considerations}
 In this section we discuss several feature requirements and design considerations for a pipe-based language to be viable for large programs.
 
-\paragraph{Not Everything Is a String.}
-The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.
+\subsection{Programming With Values}
+\subsubsection{Not Everything Is a String}\noindent
+The first core issue that prevents most shells from being viable for large programs is that only strings can be first-class values.
 
-Several projects\cite{haahr}\cite{duff}, including bash itself, have attempted to add lists and hash-maps as standard objects.
-Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes normally.
+Several projects\cite{haahr,duff}, including bash itself, have attempted to add lists and hash maps as standard objects.
+Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
 Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not available.
 
-Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
+Thus we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
 
-\paragraph{Transactional Channels.}
-Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are specified for performance reasons.
-However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, resulting in the ability to send data in batch without resorting to a queue that may interleave writes.
+\subsubsection{Consistent Channels}\noindent
+Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are used for performance reasons.
+However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, which allows for sending data in batch without resorting to a queue that may interleave writes.
 
-Consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and then the values are collected in a single output.
+Consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and the values are collected in a single output.
 This might be expressed in bash as:
-\begin{lstlisting}
-# pass through the stream, labeling every line
-label() {
-  while read line; do echo "$1$line"; done
-}
+\noindent\begin{lstlisting}
+# consume the stream, labeling every line
+label() { while read x; do echo "$1$x"; done ;}
 
-# process the stream with different labels
-split() {
-  label a &
-  label b &
-}
+# process the stream in the background
+# with two threads
+split() { label a & label b & ;}
 
-# process the numbers 1-100, limiting the
-# output to the first 3 lines
+# process the numbers 1-100,
+# limiting output to the first 3 lines
 seq 100 | split | head -3
@@
 
+\noindent
+With this code, a user might expect the output to be three lines, each consisting of a letter \verb/a/ or \verb/b/, and a number, for example:
+
+\noindent\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+a1
+a2
+a3
+\end{lstlisting}
+\end{minipage}\hfill
+\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+b1
+a3
+b2
+\end{lstlisting}
+\end{minipage}\hfill
+\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+a2
+b1
+a3
+\end{lstlisting}
+\end{minipage}
+
+\noindent
 Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved, resulting in outputs such as:
 
-\begin{verbatim}
-a1    b12   a
-a2    a     b12
-a3    a4    a34
-\end{verbatim}
-In a language intended for large programs, therefore, writes to a channel must be \emph{transactional}---a single write must be sent wholesale to a single reader without any interleaving, no matter how big the write.
+\noindent\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+a2
+b1
+b
+\end{lstlisting}
+\end{minipage}\hfill
+\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+b12
+a
+a4
+\end{lstlisting}
+\end{minipage}\hfill
+\begin{minipage}{.14\textwidth}
+\begin{lstlisting}{Name}
+a
+b12
+a34
+\end{lstlisting}
+\end{minipage}
 
-Luckily, given the decision to allow rich values to be passed through channels, a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data---or for that matter a pointer.
+\noindent
+This behaves in accordance with the Linux User's Manual \cite{linux-pipe},
+\begin{lstlisting}
+The communication channel provided by a pipe is
+a <@\emph{byte stream}@>: there is no concept of message
+boundaries.
+\end{lstlisting}
 
-\paragraph{Return Semantics.}
+\noindent
+In order to provide a channel implementation that is usable for large programs, we must ensure that it is \emph{consistent}---that is, that every read corresponds to exactly one write.
+
+\subsubsection{Return Semantics}\label{return-semantics}\noindent
 Pipeline elements such as these do not generally return values---instead, they output data to their output stream.
 However, for serious programs it is desirable to capture some kind of return value from functions.
-ES Shell \cite{haahr} accomplishes this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.
+Es Shell\cite{haahr} accomplishes this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.
 
@@
 A priori, such a collection mechanism would have to return an array, as any function call may output zero or more values.
-However, that approach would lead to the callers of functions needing to manually unwrap arrays on every function call.
-In order to avoid this, we \emph{expand} the return values into the command vector---increasing the argument number by the number of values output from the function.
+However, that approach would require the callers of functions to manually unwrap arrays on every function call.
 
-In Magritte, normal parentheses in argument position will collect and expand output values in exactly this manner:
-\begin{lstlisting}
-# a function defintion: output three values
-(count-three) = (put 1; put 2; put 3)
+\subsection{Language Features}
+\subsubsection{Variable Scope}\noindent
+Like the Es shell, we also wish to provide lambda functions with lexically scoped variables.
+However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables are bound in the function's definition environment.
+Therefore a simple typo of a variable name will result not in an error, but in the variable silently becoming dynamically bound.
 
-# a function call, using the above function
-other-fn 0 (count-three) 4
-
-# expands to
-other-fn 0 1 2 3 4
-\end{lstlisting}
-
-In this way, we can use output values in exactly the same way as return values seamlessly.
-These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that because we write transactional values to the output, we have the ability to properly separate values without relying on whitespace.
-
-\paragraph{Variable Scope.}
-Like the Es shell\cite{haahr}, we also wish to provide lambda functions with lexically scoped variables.
-However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on runtime information.
-Additionally, since Es has no syntactic difference between let-binding and mutation, even code with no free variables may change in meaning signifcantly depending on what variables are in scope in its context.
-
-To remedy this issue, we introduce a separate syntax for lexically scoped variables, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
-Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable (e.g.
-\verb/x = 1/), so that we can maintain compatibility with standard environment files.
-Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
-\verb/?x/) to allow for a planned extension for pattern matching.
-For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
-\verb/%x = 1/ or \verb/$x = 1/.
-In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
-
-\paragraph{Objects.}
-For large programs, an object system is desirable as a way to provide user defined types and abstractions.
+\subsubsection{Objects}\noindent
+For large programs, an object system is desirable as a way to provide user-defined types and abstractions.
 The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer.
@@
 
-A future planned extension would allow an environment to have multiple prototype pointers, with lookup defined in a depth-first manner.
+\subsubsection{Trees and Pattern Matching}\noindent
+For programs that deal with syntax or other kinds of trees, it is convenient to have a compact tree representation, which can be made possible with nestable vectors.
+It is also desirable to have some mechanism for pattern matching against these kinds of values.
+These kinds of structures are impossible in systems such as RC Shell\cite{duff} and Es Shell, which implicitly flatten all lists.
 
-\paragraph{Interruption.}
+\subsection{Automatic Process Cleanup}
+\subsubsection{Interruption}\noindent
 A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
@@
 
-In this example, three processes are spawned: (1) a process with a file open that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
+\noindent
+In this example, three processes are spawned: (1) a process with an open file that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
 A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.
@@
 
-\paragraph{Compensation.} In order to observe interruption, user code must be able to react to and handle interruption.
+\subsubsection{Compensation}\noindent
+In order to observe interruption, user code must be able to react to and handle interruption.
 The most direct way to observe interruption is through a system of \emph{compensation actions}, which are expressions which will be evaluated in the case of an interruption.
 
-For example, the function \verb/file-lines/ above could be implemented as:
-\begin{lstlisting}
-(file-lines ?fname) = (
-  handle = (open-file $fname) %%! close-file $fname
-  until (=> eof? $handle) (=> read-until "\n" $handle)
-)
-\end{lstlisting}
+Compensations are the only way in which interruptions can be directly observed by user code---without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or to observe the memory footprint of the program.
+Thus, we can define the primary task of the interruption system as \emph{running a process's compensation actions at the appropriate time}.
 
-In this code, the \verb/%%!/ symbol registers a compensation action that will be run either at the next checkpoint (the end of the function), or when the process is interrupted.
-In this way, we can ensure that the file is closed when the process exits.
-Compensations are the only way in which interruptions can be directly observed by client code---without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or observe the memory footprint of the program.
-Thus, we can define the primary task of our system as \emph{running a process's compensation actions at the appropriate time}.
-
-\paragraph{Interrupting on Operations vs. Eager Interruption.}
-
+\subsubsection{Lazy Interruption vs. Eager Interruption}\label{lazy-interruption}\noindent
 In designing an interruption semantic, we must first decide whether processes are eagerly interrupted when their channels close, or only interrupted on channel operations.
@@
 Consider the following example, with the assumption that \\ \verb/do-other-operation/ does not write to the standard output:
+\begin{lstlisting}
+( put 1 2 3; do-other-operation ) | take 3
+\end{lstlisting}
 
-\begin{verbatim}
-count-forever | ( take 5; do-other-operation ) | take 5
-\end{verbatim}
-
-The middle process will write five times, and then continue to do other processing in-thread that does not write to the standard output.
-The final \verb/take 5/ operation will return after 5 inputs are read.
-It is important to decide, then, whether the middle process should be \emph{interrupted} in the middle of \verb/do-other-operation/, or should it be left alone?
+\noindent
+The first process will write three times, and then continue to do other processing in-thread that does not write to the standard output.
+The final \verb/take 3/ operation will return after 3 inputs are read.
+It is important to decide, then, whether the middle process should be \emph{interrupted} in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or should it be left alone until it attempts to write a value (\emph{lazy interruption})?
 
@@
 
-Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for extra values, and we discuss some features that may mitigate this problem in \emph{Future Work}.
-
-\paragraph{Non-closing of Channels.}
-
+\subsubsection{Non-closing of Channels}\noindent
 Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel.
@@
 
-\paragraph{Closing Permanence.}
+\subsubsection{Closing Permanence}\label{closing-permanence}\noindent
 Though we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel.
 In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use.
-In UNIX, for example, the named-pipes API allows for persistent channels (TODO: cite!) that, being merely files, can be reopened at any time.
+In UNIX, for example, the named-pipes API allows for persistent channels \cite{linux-pipe} that, being merely files, can be reopened at any time.
 
@@
 
+\noindent
 This example, while contrived, represents an unavoidable race with first-class channels---whether the process running \verb/take 10/ closes the channel before or
@@
 
-\paragraph{Masking Interruptions.}
-
+\subsubsection{Masking Interruptions}\label{masking}\noindent
 A user will often wish to continue after a particular output channel has closed, and in this case will need a facility to mask exceptions for particular channels.
@@
 
+\noindent
 In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability.
@@
 
-While we plan to explore more general mechanisms for error handling, for cases such a this it suffices to provide two builtins, \verb/produce/ and \verb/consume/, each of which takes a single function to loop forever until the standard output or standard input respectively is consumed, whereby control flow continues after the invocation.
+\section{Description of Magritte}\label{description}
+
+We propose a language that meets the above requirements.
+
+\subsubsection{Nestable Vectors}
+We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/. Combined with bareword strings, we can represent trees in a straightforward manner:
+\begin{lstlisting}
+a-tree = [node [node [leaf 1] [leaf 2]] [leaf 3]]
+\end{lstlisting}
+\noindent
+We discuss a pattern matching extension in \emph{Future Work}.
+A typical strategy for traversing this kind of structure might be a function that put all leaf node values to its output in a predefined order, to be consumed by another process.
+
+The builtin function \verb/for/ takes a vector as an argument and outputs each element in order, so that a normal vector can be traversed by the pattern:
+\begin{lstlisting}
+for [1 2 3 4 5] | each (?el => ...)
+\end{lstlisting}
+
+\subsubsection{Environments as Objects}
+We support environment capture with the straightforward \verb/{ }/ syntax.
+The semantics of curly braces are to run the code as normal, then detach the running environment from its parent and return it.
+Therefore all assignment syntaxes are available, including function definition.
+
+Environment lookup uses the \verb/!/ symbol, as in \verb/$env!key/.
+We use this symbol because it is already reserved by bash, unlike both \verb\/\ or \verb/./ which need to be available in bareword syntax to integrate with external programs.
+Environments can also be mutated using the same access syntax.
+
+\subsubsection{Synchronous Channels}\noindent
+Channels in Magritte are \emph{synchronous}---readers and writers cannot continue until a communication is completed successfully.
+Given the decision to allow rich values to be passed through channels, we have decided that a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data---or for that matter a process handle or object reference.
+Additionally, synchronous channels have simpler semantics both for implementation and for users, and we leave open the possibility of user-implemented queues, such as:
+\begin{lstlisting}
+producer | buffer 10 | consumer
+\end{lstlisting}
+
+\subsubsection{Reading, Writing, and Spawning}\noindent
+Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively.
+Spawning uses a syntax similar to bash, with a \verb/&/ at the beginning of a command indicating that it should be spawned in a new thread.
+
+\subsubsection{Collection and Substitution}\noindent
+In order to use output values seamlessly as return values without unwrapping arrays at every function call as described in Section \ref{return-semantics}, Magritte uses normal parentheses to collect and \emph{expand} the resulting values into the current command vector---increasing the argument number by the number of values output from the function.
+For example:
+\begin{lstlisting}
+# A function defintion: output three values
+(count-three) = (put 1; put 2; put 3)
+
+# A function call, using the above function.
+# Parentheses collect the three writes and expand
+# 1 2 3 in-place
+other-fn 0 (count-three) 4
+
+# equivalent to
+other-fn 0 1 2 3 4
+\end{lstlisting}
+
+\noindent
+These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that because we write transactional values to the output, we have the ability to properly separate values without relying on whitespace.
+
+This mechanism is also available in vector literals, allowing us to collect outputs as a vector:
+\begin{lstlisting}
+outputs = [(some-command)]
+\end{lstlisting}
+\noindent
+And the \verb/for/ function mentioned above can be used to splat vector arguments into function calls:
+\begin{lstlisting}
+some-fn $arg1 $arg2 (for [$arg3 $arg4])
+\end{lstlisting}
+
+\subsubsection{Lexical vs. Dynamic Variables}\noindent
+In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
+Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g.
+\verb/x = 1/), so that we can maintain compatibility with standard environment files.
+Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
+\verb/?x/) to allow for a planned extension for pattern matching.
+
+For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
+\verb/%x = 1/ or \verb/$x = 1/.
+In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
+
+\subsubsection{Compensation and Unconditional Compensation}\noindent
+We implement compensation directly using the \verb/%%/ operator to indicate a compensation action, which is cleared at the end of the current stack frame.
+\begin{lstlisting}
+(my-function) = (
+  # ...
+  action %% cleanup-action
+  # ...
+)
+\end{lstlisting}
+
+\noindent Additionally, we define \emph{unconditional compensations} using the \verb/%%!/ operator, which run both in the case of an interruption and in the case of a normal return.
+In this way, they are analogous to \verb/finally/ or \verb/ensure/ sections of standard exception handling.
+
+For example, the function \verb/read-lines/ above could be implemented as:
+\begin{lstlisting}
+(read-lines ?fname) = (
+  f = (open-file $fname) %%! close-file $fname
+  until (=> eof? $f) (=> read-until "\n" $f)
+)
+\end{lstlisting}
+
+\noindent
+In this way, we can ensure that the file is closed when the process exits, whether by a normal return or by interruption.
+
+\subsubsection{Lazy Interruption}\noindent
+Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for the extra-values problem discussed in Section \ref{lazy-interruption}, and we discuss some features that may mitigate this problem in \emph{Future Work}.
+
+\subsubsection{Interrupt Handling}\noindent
+While we plan to explore more general mechanisms for exception handling, we find that it suffices for most applications to provide two builtins, \verb/produce/ and \verb/consume/, to indicate the intent to consume the entirety of the output or input streams, respectively.
+Each of these builtins takes a single zero-argument function which will loop forever until the standard output or standard input respectively is closed, whereby control flow continues after the invocation.
 Using these functions, we might define \verb/each/ as:
@@
 
-With this definition, the call to \verb/consume/ will mask the interruption, and control flow within the \verb/sample/ function will proceed as normal.
+\noindent
+With this definition, the call to \verb/consume/ will mask the interruption from the standard input closing, and control flow after any \verb/each/ invocation will continue as normal.
+This is enough to resolve the issue discussed in Section \ref{masking}.
 
-\section{Implementation of Magritte}
+\section{Implementation of Magritte}\label{implementation}
 
@@
 
-\paragraph{Channel Registry.}
-
+\subsubsection{Channel Registry}\noindent
 Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described above.
@@
 
-Once the channel is closed, every call to \verb/read/ and \verb/write/ will cause the calling process to be interrupted, as described above in \emph{Closing Permanence}.
+Once the channel is closed, every call to \verb/read/ and \verb/write/ will cause the calling process to be interrupted, as described above in Section \ref{closing-permanence}.
 
@@
 
-\paragraph{Spawning Order Dependency.}
-
-It is necessary to take some care with the implementation of the \verb/spawn/ primitive, that we wait until the spawned process has finished registering its channels before the spawning process continues.
+\subsubsection{Spawning Order Dependency}\noindent
+It is necessary to take some care with the implementation of the \verb/spawn/ primitive (\verb/&/), that we wait until the spawned process has finished registering its channels before the spawning process continues.
 Consider the example, which outputs 10 numbers, possibly out of order:
@@
 \end{lstlisting}
+\noindent
 The middle process is responsible for spawning two processes that funnel data from their input to their output.
@@
 
-\paragraph{Collectors.}
-In order to implement the call semantics described above, we also implement a write-only channel called a \emph{collector}.
+\subsubsection{Collectors}\noindent
+In order to implement the return semantics described above, we also implement a write-only channel called a \emph{collector}.
 Collectors cannot be created directly by users, but only appear in the interpreter when we evaluate a \verb/Subst/ node, which represents parentheses in argument position.
@@
 
+\noindent
 In this example, there is an open question of how long we should wait until reading the collection and continuing.
@@
 To implement this, we add a \verb/wait_for_close/ intrinsic to collectors, which returns immediately if closed, and otherwise adds the current thread to a waiting set and sleeps.
-Upon closing the channel in \verb/remove_writer/, all elements of this waiting set are awoken, and flow continues.
+Upon closing the channel, all elements of this waiting set are awoken, and flow continues.
 Thus the algorithm for substiution is:
@@
 * Create a new collector $c
-* Evaluate the contents with standard output into the collector
+* Evaluate the contents with standard output set to the collector
 * Run $c.wait_for_close
@@
 
-\section{Related Work}
+\section{Future Work}\label{future-work}
 
-\paragraph{Iterators.}
-(TODO: iterators in other languages?)
-PEP 533 Authors(TODO: cite properly) describe an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
+\subsubsection{Open Checking}\noindent
+Because our system implements \emph{lazy interruption}, we must somehow mitigate the problem of producers creating one extra element than is needed, in the case that elements are expensive to produce.
+One possibility is to introduce the concept of \emph{open checking}, so that producing processes can periodically run a command that will interrupt in the case that a channel is closed, without actually writing any data to the channel.
+
+\subsubsection{Pattern Matching}\noindent
+Though the syntax design is not yet complete, we plan to allow pattern matching in all lambda bindings.
+In order to potentially support glob and regular expression matching, we must be careful which symbols we reserve in this context.
+
+\subsubsection{Prototypical Inheritance}\noindent
+One missing feature to support prototypical inheritance is the dynamic assignment of a \verb/self/ variable.
+We plan to implement a system similar to JavaScript, in that this variable is assigned dynamically based on the calling syntax, or overrideable with special-purpose functions.
+
+Additionally, we intend to support multiple inheritance by allowing for an environment to point to a list of parents, and resolve names in a depth-first manner.
+
+\subsubsection{UNIX Integration}\noindent
+In order to be viable as a shell, we must integrate seamlessly with a UNIX environment.
+This involves marshalling values to strings and byte streams in order to communicate with external processes, and allowing external processes to interact with Magritte values.
+We plan to use a suite of parsing and unparsing functions for linewise and table data, along with a custom JSON streaming format to allow integration with most languages.
+We also plan to use a socket and a static executable to allow external programs to call back into Magritte lambda functions.
+
+\subsubsection{Performance}\noindent
+The current implementation, being a Ruby interpreter, is slow. We plan to bootstrap using a JIT-compiled virtual machine implemented in RPython.
+
+\section{Related Work}\label{related-work}
+
+\subsubsection{Iterators}
+\noindent
+Many general purpose languages such as Python implement an iterator API, for code that is intended to produce values in a streamlike fashion.
+For example, in a Python Enhancement Proposal (PEP), Smith\cite{pep533} describes an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
+
 But channel-based systems generally allow for multiple concurrent readers and writers---so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
 
-\paragraph{Go Language.}
+\subsubsection{General-Purpose Languages}
 
-The Go programming language contains a robust channel implementation, which can be used to implement a wide variety of concurrent algorithms.
-In an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
+\noindent
+Many general purpose languages, including \verb/Go/ and \verb/Clojure/ contain robust channel implementations, which could be used to implement a wide variety of concurrent algorithms.
+For example, in an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
 \begin{lstlisting}[language=C, morekeywords={fn,channel,nil,go}]
@@
 \end{lstlisting}
-
+\noindent
 However, such a system would still require manual use of input and output channels when a simple pipeline would suffice---and the user must still translate between returning vs.
@@
 
-Python PEP 533; Go, Erlang; Guile Scheme, TCL;
+\subsubsection{General-Purpose Language Extensions}
+\noindent
+Projects such as \verb/xonsh/\cite{xonsh} and \verb/scheme-shell/(TODO: cite) attempt to extend the syntax of an existing general purpose language to include shell capabilities.
+Unfortunately, this approach creates a schism between shell concepts and host language concepts---in particular, host language objects cannot be passed through pipes.
 
-\paragraph{RC Shell.}
+\subsubsection{Object Shells}
+\noindent
+Projects such as \verb/powershell/ and \verb/mash/ allow objects (.NET objects and Scala objects, respectively) to be passed through pipelines.
+However, this approach still ties the language to one specific platform and makes it difficult to integrate with programs written for other platforms.
+
+\subsubsection{Desktop Scripting Languages}
+Projects such as \verb/Guile Scheme/\cite{guile} and \verb/TCL/(TODO: cite) attempt to create a separate API for integrating desktop programs, eliminating the need for shell entirely.
+However since this approach would require an unreasonably large integration effort on the part of application developers, their adoption has been limited in practice.
+Our goal is to create a system that can interact with programs in their current forms.
+
+\subsubsection{RC Shell}
+\noindent
 The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash.
@@
 
-\paragraph{Es Shell.}
+\subsubsection{Es Shell}
+\noindent
 The \verb/es/ shell \cite{haahr} attempts to extend the standard shell model with lambda functions with lexical scoping.
 These shell functions are marshalled into strings when passed through pipes (whether to shell functions or external programs), but are still written byte-by-byte and thus are subject to interleaving in concurrent cases.
+
+\subsubsection{PUSH Shell}
+\noindent
+The PUSH shell\cite{push} extends traditional shells with \emph{fan-out} and \emph{fan-in} operators which separate data into discrete records.
+This enables it to be used safely to orchestrate distributed computing tasks.
+However, the underlying pipe implementation remains byte-based, so that features like Magritte's collection and expansion remain impossible in general.
+
+\section{Conclusion}\label{conclusion}
+
+We believe that Magritte is a promising platform for shell scripting, and that it is also viable for large programs.
+Our channel implementation allows for robust composition of concurrent algorithms, and can also serve as the fundamental unit of composition for a general-purpose language.
+
+\begin{acknowledgment}
+Johannes Westlund's contribution to this research was made possible by funding from The Sweden-Japan Foundation and Stockholms Grosshandelssocietet.
+\end{acknowledgment}
 

### 2018-12-14 doc/paper.tex
paper updates

@@
 % notes
-\documentclass[english,preprint,JIP,uplatex]{ipsj}
+\documentclass[english,PRO]{ipsj}
+\usepackage{PROpresentation}
+\PROheadtitle{y-n-(x): Manuscript for presentation at IPSJ-SIGPRO, d m Y.}
 
@@
 \author{Jeanine Miller Adkisson}{TITech}[jneen@jneen.net]
+\author{Johannes Westlund}{TITech,KTH}[jwestlun@kth.se]
 \author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]
-\author{Johannes Westlund}{TITech,KTH}[jwestlun@kth.se]
 
@@
 % * Find an example of a non-linear pipe arrangement (prob. from flow-based programming)
-% * Bibliography
 % * Pretty Syntax
@@
 
+\noindent
 The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes.
-Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no support for transactional writes.
+Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no boundaries between values.
 
-In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes in which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.
+In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.
 
@@
 
-\subsubsection*{Outline}\noindent
+\subsection*{Outline}\noindent
 In this work, we present a new shell language \emph{Magritte}.
@@
 
-\section{Design Considerations}\label{design-considerations}
+\section{Design Considerations}\label{design-considerations}\noindent
 In this section we discuss several feature requirements and design considerations for a pipe-based language to be viable for large programs.
@@
 \subsubsection{Not Everything Is a String}\noindent
-The first core issue that prevents most shells from being viable for large programs is that only strings can be first-class values.
+The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.
 
 Several projects\cite{haahr,duff}, including bash itself, have attempted to add lists and hash maps as standard objects.
-Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
-Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not available.
+Unfortunately, while they allow these values to be stored in variables, most systems do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
+Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not possible.
 
-Thus we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
+Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
 
@@
 Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are used for performance reasons.
-However, in the languages we consider, we assume we can transactionally write a single complex value to a channel, which allows for sending data in batch without resorting to a queue that may interleave writes.
+However, it is important if a pipe is to be used for composition that reads correspond exactly to writes, and that data does not become corrupted by interleaving.
 
-Consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and the values are collected in a single output.
+For example, consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and the values are collected in a single output.
 This might be expressed in bash as:
-\noindent\begin{lstlisting}
-# consume the stream, labeling every line
+\noindent\begin{lstlisting}[morekeywords={do,done}]
+# <@\underline{\color{gray} consume}@> the stream, labeling every line
 label() { while read x; do echo "$1$x"; done ;}
@@
 
-# process the numbers 1-100,
+# <@\underline{\color{gray} produce}@> and process the numbers 1-100,
 # limiting output to the first 3 lines
@@
 \noindent
-This behaves in accordance with the Linux User's Manual \cite{linux-pipe},
+This behavior is in accordance with the Linux User's Manual \cite{linux-pipe}:
 \begin{lstlisting}
@@
 
-If we allow values to be written to output streams, however, this approach is redundant---a function author must decide whether to return a value or write it to standard output.
+If we allow values to be written to output streams, however, this approach is redundant---a function author must decide whether to return a value or write it to the standard output.
 Thus it is desirable to have an interpolation syntax which \emph{collects} the output of a command synchronously.
 
-A priori, such a collection mechanism would have to return an array, as any function call may output zero or more values.
-However, that approach would require the callers of functions to manually unwrap arrays on every function call.
+A priori, such a collection mechanism would have to return an list, as any function call may output zero or more values.
+However, that approach would require the callers of functions to manually unwrap lists on every function call.
 
@@
 Like the Es shell, we also wish to provide lambda functions with lexically scoped variables.
-However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables are bound in the function's definition environment.
-Therefore a simple typo of a variable name will result not in an error, but in the variable silently becoming dynamically bound.
+However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables happen to be bound in the function's definition environment.
+Therefore a simple mistype of a variable name will result not in an error, but in the variable silently becoming dynamically bound.
 
@@
 The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer.
-If we allow environments as first-class objects, they can be used in a straightforward manner as prototype-based objects.
+Therefore, if we allow environments as first-class values, they can be used in a straightforward manner as prototype-based objects.
 
@@
 It is also desirable to have some mechanism for pattern matching against these kinds of values.
-These kinds of structures are impossible in systems such as RC Shell\cite{duff} and Es Shell, which implicitly flatten all lists.
+These kinds of structures are impossible using the list features in systems such as RC Shell\cite{duff} and Es Shell, which implicitly flatten all lists.
 
@@
 \begin{lstlisting}
-  read-lines tmp/large-file
-    | take 10
-    | each (?line => do-expensive-work <@\$@>line)
+  read-lines tmp/large-file                        <@\color{gray}\hfill $(1)$@>
+    | take 10                                      <@\color{gray}\hfill $(2)$@>
+    | each (?line => do-expensive-work <@\$@>line) <@\color{gray}\hfill $(3)$@>
 \end{lstlisting}
@@
 In a pipe-based language, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
-A user will also expect that, after the first 10 lines are processed, the file is closed, and that all three processes exit.
+A user will also expect that, after the first 10 lines are processed and the \verb/take/ function returns, the file will be closed, and all three processes exit.
 This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.
@@
 The final \verb/take 3/ operation will return after 3 inputs are read.
-It is important to decide, then, whether the middle process should be \emph{interrupted} in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or should it be left alone until it attempts to write a value (\emph{lazy interruption})?
+It is important to decide, then, whether the first process should be \emph{interrupted} in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or whether it should be left alone until it attempts to write a value (\emph{lazy interruption}).
 
@@
 
-\subsubsection{Non-closing of Channels}\noindent
+\subsubsection{Non-closing of Channels}\label{non-closing}\noindent
 Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel.
@@
 When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
-Therefore the appropriate time for it to be interrupted is if it is blocked on a channel that will never receive any more operations.
+Therefore the appropriate time for it to be interrupted is when it is blocked on a channel that will never receive any more operations.
 In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.
@@
 \subsubsection{Closing Permanence}\label{closing-permanence}\noindent
-Though we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel.
+Even when we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel.
 In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use.
@@
     $i = (inc $i)
-    # mutate the $hold variable with probability 1/$i
-    prob $i 1 && ($hold = %v)
+    # mutate the $hold variable
+    # with probability 1/$i
+    prob (reciprocal $i) 1 && ($hold = %v)
   )
@@
 
-\section{Description of Magritte}\label{description}
-
+\section{Description of Magritte}\label{description}\noindent
 We propose a language that meets the above requirements.
 
-\subsubsection{Nestable Vectors}
+\subsection{Values and Variables}
+
+\subsubsection{Lexical vs. Dynamic Variables}\noindent
+In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
+Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g.
+\verb/x = 1/), so that we can maintain compatibility with standard environment files.
+Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
+\verb/?x/) to allow for a planned extension for pattern matching.
+
+For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
+\verb/%x = 1/ or \verb/$x = 1/.
+In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
+
+\subsubsection{Lambda Functions}\noindent
+Lambda functions are specified with parenthesized expressions containing the arrow symbol \verb/=>/, which separates the bindings from the body.
+Multiple matching clauses are possible, and will be extended in the future with more general pattern matching.
+Clauses are separated at the beginning of lines that contain \verb/=>/, which avoids the need for any indentation-based parsing.
+For example:
+\begin{lstlisting}
+my-function = (
+  ?x =>
+    put one-argument %x
+  ?y ?z =>
+    put two-arguments
+    put %y %z
+  ... => ...
+)
+\end{lstlisting}
+
+\noindent
+Named functions can also be defined using a parenthesized expression as the left hand side of an assignment:
+\begin{lstlisting}
+(my-function ?x ?y) = ...
+
+# equivalent to
+my-function = (?x ?y => ...)
+\end{lstlisting}
+
+\noindent
+Functions are scanned statically for free lexical variables, which are captured at function definition time, and merged into the environment at call time.
+
+\subsubsection{Nestable Vectors}\noindent
 We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/. Combined with bareword strings, we can represent trees in a straightforward manner:
@@
 
-\subsubsection{Environments as Objects}
+\subsubsection{Environments as Objects}\noindent
 We support environment capture with the straightforward \verb/{ }/ syntax.
@@
 Therefore all assignment syntaxes are available, including function definition.
+For example:
+\begin{lstlisting}
+(make-account ?balance) = {
+  balance = $balance
+  (deposit ?amt) =
+    (%balance = (add %amt %balance))
+  (withdraw ?amt) = (
+    (%balance = (sub %amt %balance))
+}
+\end{lstlisting}
 
+\noindent
 Environment lookup uses the \verb/!/ symbol, as in \verb/$env!key/.
@@
 Environments can also be mutated using the same access syntax.
-
-\subsubsection{Synchronous Channels}\noindent
-Channels in Magritte are \emph{synchronous}---readers and writers cannot continue until a communication is completed successfully.
-Given the decision to allow rich values to be passed through channels, we have decided that a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data---or for that matter a process handle or object reference.
-Additionally, synchronous channels have simpler semantics both for implementation and for users, and we leave open the possibility of user-implemented queues, such as:
+For example:
 \begin{lstlisting}
-producer | buffer 10 | consumer
+account = (make-account 10)
+$account!withdraw 4
+put $account!balance # => 6
 \end{lstlisting}
 
-\subsubsection{Reading, Writing, and Spawning}\noindent
-Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively.
-Spawning uses a syntax similar to bash, with a \verb/&/ at the beginning of a command indicating that it should be spawned in a new thread.
+\subsubsection{Reading and Writing}\noindent
+Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively. These block until they complete a communication.
 
@@
 
-# A function call, using the above function.
 # Parentheses collect the three writes and expand
@@
 \noindent
-And the \verb/for/ function mentioned above can be used to splat vector arguments into function calls:
+The \verb/list/ built-in function, which simply returns its argument vector, is also available for this purpose.
+The \verb/for/ function mentioned above can be used to splat vector arguments into function calls:
 \begin{lstlisting}
@@
 
-\subsubsection{Lexical vs. Dynamic Variables}\noindent
-In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
-Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g.
-\verb/x = 1/), so that we can maintain compatibility with standard environment files.
-Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
-\verb/?x/) to allow for a planned extension for pattern matching.
+\subsubsection{Blocks}\noindent
+A parenthesized grouping that is \emph{not} in argument position is a \emph{block}.
+Blocks do not collect or modify the environment's channels in any way, but instead simply run the code contained within them, and output values normally.
+These are mostly used to group commands within a pipeline, and in the body of function definitions:
+\begin{lstlisting}
+produce | (process; process; process)
+\end{lstlisting}
+\noindent
+In the case that a user might want to use a substitution at the root level---i.e.
+to generate a function and immediately call it, we provide the \verb/exec/ builtin function which executes its arguments as a normal command.
 
-For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
-\verb/%x = 1/ or \verb/$x = 1/.
-In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
+\subsection{Channels and Processes}
+\subsubsection{Synchronous Channels}\noindent
+Channels in Magritte are \emph{synchronous}---readers and writers cannot continue until a communication is completed successfully.
+Given the decision to allow rich values to be passed through channels, we have decided that a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data---or for that matter a process handle or object reference.
+Additionally, synchronous channels have simpler semantics both for implementation and for users, and we leave open the possibility of user-implemented queues, such as:
+\begin{lstlisting}
+producer | buffer 10 | consumer
+\end{lstlisting}
+
+\subsubsection{Spawning and Redirecting}\noindent
+Spawning uses a syntax similar to bash, with an ampersand (\verb/&/) at the beginning of a command indicating that it should be spawned in a new thread.
+
+Standard input and output may be redirected into and from channels using the \verb/>/ and \verb/</ symbols, or chained together with the pipe (\verb/|/) symbol, much like \verb/bash/.
+Commands in pipelines will all be spawned in their own threads, with the exception of the final command, which will be run in the current process.
+
+\subsection{Interruption and Compensation}
+\subsubsection{Lazy Interruption}\noindent
+Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for the extra-values problem discussed in Section \ref{lazy-interruption}, and we discuss some features that may mitigate this problem in \emph{Future Work}.
 
 \subsubsection{Compensation and Unconditional Compensation}\noindent
-We implement compensation directly using the \verb/%%/ operator to indicate a compensation action, which is cleared at the end of the current stack frame.
+We implement compensation directly using the \verb/%%/ operator to indicate a compensation action, which is cleared at the end of the current function body:
 \begin{lstlisting}
@@
 \noindent
-In this way, we can ensure that the file is closed when the process exits, whether by a normal return or by interruption.
-
-\subsubsection{Lazy Interruption}\noindent
-Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for the extra-values problem discussed in Section \ref{lazy-interruption}, and we discuss some features that may mitigate this problem in \emph{Future Work}.
+In this way, we can ensure that the file is closed when the function exits, whether by a normal return or by interruption.
 
 \subsubsection{Interrupt Handling}\noindent
-While we plan to explore more general mechanisms for exception handling, we find that it suffices for most applications to provide two builtins, \verb/produce/ and \verb/consume/, to indicate the intent to consume the entirety of the output or input streams, respectively.
-Each of these builtins takes a single zero-argument function which will loop forever until the standard output or standard input respectively is closed, whereby control flow continues after the invocation.
+While we plan to explore more general mechanisms for exception handling, we find that it suffices for most applications to provide two builtin functions, \verb/produce/ and \verb/consume/, to indicate the intent to fill or consume the entirety of the output or input streams, respectively.
+Each of these functions takes a single zero-argument function which will loop forever until the standard output or standard input respectively is closed, whereby control flow continues after the invocation.
 Using these functions, we might define \verb/each/ as:
@@
 
+\noindent
 The current version of Magritte is implemented as a straightforward interpreter written in Ruby, using Ruby's builtin threads, exceptions, and mutexes to implement processes and channels.
 
-\subsubsection{Channel Registry}\noindent
-Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described above.
+\subsection{Channel Registry}\noindent
+Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described in Section \ref{non-closing}.
 When \verb/remove_reader/ or \verb/remove_writer/ result in an empty registered set, they additionally close the channel and raise an internal \verb/Interrupt/ exception in every blocked thread.
@@
 
-In order to reduce unnecessary use of Ruby threads, we also find it is simpler to, instead of registering \emph{processes} to the channels, to register \emph{frames}.
-In this way, we can register all inputs and outputs on frame entry, and use standard Ruby exception handling to ensure we properly unregister all inputs and outputs on frame exit.
-This means that different frames in the same process can be redirected to different channels, which makes the \verb/>/ and \verb/</ redirection syntax straightforward to implement---we simply push a new frame with different channels attached.
+In order to reduce unnecessary use of Ruby threads, we also find it is simpler, instead of registering \emph{processes} to the channels, to register \emph{stack frames}.
+In this way, we can register all inputs and outputs on frame entry, and use standard Ruby exception handling to ensure we properly run compensations and deregister inputs and outputs on frame exit.
+This means that different frames in the same process can be connected to different channels, which makes the \verb/>/ and \verb/</ redirection syntax straightforward to implement---we simply push a new frame with different channels attached.
 
@@
 
-\subsubsection{Spawning Order Dependency}\noindent
-It is necessary to take some care with the implementation of the \verb/spawn/ primitive (\verb/&/), that we wait until the spawned process has finished registering its channels before the spawning process continues.
-Consider the example, which outputs 10 numbers, possibly out of order:
+\subsection{Spawning Order Dependency}\noindent
+It is necessary to take some care with the implementation of the spawning primitive (\verb/&/), that we wait until the spawned process has finished registering its channels before the spawning process continues.
+Consider the following example, which outputs 10 numbers, possibly out of order:
 \begin{lstlisting}
@@
 
-For this reason, we have special case handling for the root frame---the channels are registered in the \emph{spawning} process, before the spawned process is allowed to start.
+For this reason, we have special case handling for the root frame of a process---the channels are registered in the \emph{spawning} process, before the spawned process is allowed to start.
 
-\subsubsection{Collectors}\noindent
-In order to implement the return semantics described above, we also implement a write-only channel called a \emph{collector}.
+\subsection{Collectors}\noindent
+In order to implement the return semantics described in Section \ref{return-semantics}, we also implement a write-only channel called a \emph{collector}.
 Collectors cannot be created directly by users, but only appear in the interpreter when we evaluate a \verb/Subst/ node, which represents parentheses in argument position.
@@
 
+\noindent
+In the most common case, there will only be a single thread writing to the collector, so that the channel will be closed by the time the evaluation is finished, making \verb/wait_for_close/ a null operation.
+
 \section{Future Work}\label{future-work}
 
-\subsubsection{Open Checking}\noindent
+\subsection{Open Checking}\noindent
 Because our system implements \emph{lazy interruption}, we must somehow mitigate the problem of producers creating one extra element than is needed, in the case that elements are expensive to produce.
-One possibility is to introduce the concept of \emph{open checking}, so that producing processes can periodically run a command that will interrupt in the case that a channel is closed, without actually writing any data to the channel.
+One possibility is to introduce the concept of \emph{open checking}, so that producer processes can periodically run a command that will interrupt in the case that the output is closed, without actually writing any data to the channel.
 
-\subsubsection{Pattern Matching}\noindent
+\subsection{Pattern Matching}\noindent
 Though the syntax design is not yet complete, we plan to allow pattern matching in all lambda bindings.
@@
 
-\subsubsection{Prototypical Inheritance}\noindent
-One missing feature to support prototypical inheritance is the dynamic assignment of a \verb/self/ variable.
-We plan to implement a system similar to JavaScript, in that this variable is assigned dynamically based on the calling syntax, or overrideable with special-purpose functions.
+\subsection{Prototypical Inheritance}\noindent
+One missing feature to support prototypical inheritance is the dynamic assignment of a special \verb/$self/ variable.
+We plan to implement a system similar to JavaScript, in that the \verb/$self/ variable is assigned dynamically based on the calling syntax, or overrideable with special-purpose functions.
 
-Additionally, we intend to support multiple inheritance by allowing for an environment to point to a list of parents, and resolve names in a depth-first manner.
+Additionally, we plan to support multiple inheritance by allowing for an environment to point to a list of parents, and resolve names in a depth-first manner.
 
-\subsubsection{UNIX Integration}\noindent
+\subsection{UNIX Integration}\noindent
 In order to be viable as a shell, we must integrate seamlessly with a UNIX environment.
@@
 
-\subsubsection{Performance}\noindent
+\subsection{Macros}\noindent
+While we do not currently have a use case for macros, it is likely that there is a use case for syntactic abstraction.
+Our parsing technique involves an intermediate \emph{skeleton tree} representation adapted from Bachrach and Playford\cite{skeleton-trees} that makes macro systems simpler to implement in non-lisp syntaxes.
+
+\subsection{Performance}\noindent
 The current implementation, being a Ruby interpreter, is slow. We plan to bootstrap using a JIT-compiled virtual machine implemented in RPython.
@@
 
-\subsubsection{Iterators}
+\subsection{Iterators}
 \noindent
@@
 
-\subsubsection{General-Purpose Languages}
+\subsection{General-Purpose Languages}
 
 \noindent
-Many general purpose languages, including \verb/Go/ and \verb/Clojure/ contain robust channel implementations, which could be used to implement a wide variety of concurrent algorithms.
+Many general purpose languages, including Go\cite{golang} and Clojure\cite{clojure}, contain robust channel implementations, which could be used to implement a wide variety of concurrent algorithms.
 For example, in an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
@@
 Additionally, Go does not contain any kind of channel closing semantics or automatic process cleanup.
-Finally, such a system would be far too heavyweight for a REPL context, and would not make for a viable shell.
+Finally, such a system would be far too heavyweight for a REPL context, and would not make for a viable interactive shell.
 
-\subsubsection{General-Purpose Language Extensions}
+\subsection{General-Purpose Language Extensions}
 \noindent
-Projects such as \verb/xonsh/\cite{xonsh} and \verb/scheme-shell/(TODO: cite) attempt to extend the syntax of an existing general purpose language to include shell capabilities.
+Projects such as \verb/xonsh/\cite{xonsh} and \verb/scheme-shell/\cite{scheme-shell} attempt to extend the syntax of an existing general purpose language to include shell capabilities.
 Unfortunately, this approach creates a schism between shell concepts and host language concepts---in particular, host language objects cannot be passed through pipes.
 
-\subsubsection{Object Shells}
-\noindent
-Projects such as \verb/powershell/ and \verb/mash/ allow objects (.NET objects and Scala objects, respectively) to be passed through pipelines.
+\subsection{Object Shells}\noindent
+Projects such as \verb/powershell/\cite{powershell} and \verb/mash/\cite{mash} allow objects (.NET objects and Scala objects, respectively) to be passed through pipelines.
 However, this approach still ties the language to one specific platform and makes it difficult to integrate with programs written for other platforms.
 
-\subsubsection{Desktop Scripting Languages}
-Projects such as \verb/Guile Scheme/\cite{guile} and \verb/TCL/(TODO: cite) attempt to create a separate API for integrating desktop programs, eliminating the need for shell entirely.
+\subsection{Desktop Scripting Languages}\noindent
+Projects such as \verb/Guile Scheme/\cite{guile} and \verb/TCL/\cite{tcl} attempt to create a separate API for integrating desktop programs, eliminating the need for shell entirely.
 However since this approach would require an unreasonably large integration effort on the part of application developers, their adoption has been limited in practice.
@@
 
-\subsubsection{RC Shell}
-\noindent
+\subsection{RC Shell}\noindent
 The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash.
@@
 
-\subsubsection{Es Shell}
+\subsection{Es Shell}
 \noindent
@@
 
-\subsubsection{PUSH Shell}
-\noindent
-The PUSH shell\cite{push} extends traditional shells with \emph{fan-out} and \emph{fan-in} operators which separate data into discrete records.
+\subsection{PUSH Shell}\noindent
+The PUSH shell\cite{push} extends a traditional shell with \emph{fan-out} and \emph{fan-in} operators which separate data into discrete records.
 This enables it to be used safely to orchestrate distributed computing tasks.
@@
 
-We believe that Magritte is a promising platform for shell scripting, and that it is also viable for large programs.
+\noindent
+We believe that it should be possible to script, orchestrate, and interact with an OS environment in a language that is also capable of high-level abstraction.
+We believe that Magritte is a promising platform for shell scripting and for interaction, and that it is also viable for large programs.
+
 Our channel implementation allows for robust composition of concurrent algorithms, and can also serve as the fundamental unit of composition for a general-purpose language.

### 2018-12-14 lib/magritte/ast.rb
restructure Block, Subst and Env to contain Group
Group is only responsible for running a group of commands, whereas
Block, Subst, and Env are responsible for creating the correct
environment to run them in.

@@ -95,11 +95,15 @@ module Magritte
     end
 
     class Block < Tree::Node
+      defrec :group
+    end
+
+    class Group < Tree::Node
       deflistrec :elems
     end
 
     class Subst < Tree::Node
-      deflistrec :elems
+      defrec :group
     end
 
     class Vector < Tree::Node

### 2018-12-14 lib/magritte/free_vars.rb

@@ -25,11 +25,7 @@ module Magritte
         out
       end
 
-      def visit_subst(node, bound_vars)
-        visit_block(node, bound_vars)
-      end
-
-      def visit_block(node, bound_vars)
+      def visit_group(node, bound_vars)
         out = Set.new
         so_far = Set.new
         node.elems.each do |elem|

### 2018-12-14 lib/magritte/interpret.rb

@@ -60,16 +60,22 @@ module Magritte
         Status.normal
       end
 
-      def visit_block(node)
+      def visit_group(node)
         out = Status.normal
         node.elems.each { |elem| out = visit_exec(elem) }
         out
       end
 
+      def visit_block(node)
+        Proc.enter_frame(Proc.current.env.extend) do
+          visit_exec(node.group)
+        end
+      end
+
       def visit_subst(node)
-        out = Status.normal
+        out = nil
         s_ do
-          node.elems.each { |elem| out = visit_exec(elem) }
+          out = visit_exec(node.group)
         end.collect.each { |x| yield x }
         out
       end

### 2018-12-14 lib/magritte/parser.rb

@@ -29,7 +29,7 @@ module Magritte
       elems = skel.sub_items.map do |item|
         parse_line(item)
       end
-      AST::Block[elems]
+      AST::Group[elems]
     end
 
     def parse_line(item)
@@ -149,7 +149,7 @@ module Magritte
       bare_command = Skeleton::Item[vec]
 
       bare_command.match(singleton(nested(:lparen,~_))) do |i|
-        return with(redirects, parse_root(i))
+        return with(redirects, AST::Block[parse_root(i)])
       end
 
       vec = parse_terms(bare_command)
@@ -229,10 +229,10 @@ module Magritte
           end
 
           patterns = bindings.map { |i| AST::VectorPattern[i.elems.map { |e| AST::Binder[e.value] }, nil]}
-          return AST::Lambda[name, patterns, bodies.map { |body| AST::Block[body.map { |line| parse_line(line) }]}]
+          return AST::Lambda[name, patterns, bodies.map { |body| AST::Group[body.map { |line| parse_line(line) }]}]
         end
 
-        return AST::Subst[item.sub_items.map { |e| parse_line(e) }]
+        return AST::Subst[AST::Group[item.sub_items.map { |e| parse_line(e) }]]
       end
 
       term.match(nested(:lbrack,~_)) do |item|

### 2018-12-14 spec/parser_spec.rb

@@ -80,25 +80,6 @@ describe Magritte::Parser do
     end
   end
 
-  describe "if-else statement" do
-    let(:input) {
-      """
-      cond && (command) !! cond2 && (command2)
-      """
-    }
-
-    it do
-      assert { ast.elems.size == 1 }
-      assert { ast.elems.first.is_a?(Magritte::AST::Else) }
-      assert { ast.elems.first.lhs.is_a?(Magritte::AST::And) }
-      assert { ast.elems.first.rhs.is_a?(Magritte::AST::And) }
-      assert { ast.elems.first.lhs.lhs.vec.first.value == "cond" }
-      assert { ast.elems.first.lhs.rhs.elems.first.vec.first.value == "command" }
-      assert { ast.elems.first.rhs.lhs.vec.first.value == "cond2" }
-      assert { ast.elems.first.rhs.rhs.elems.first.vec.first.value == "command2" }
-    end
-  end
-
   describe "switch statement" do
     let(:input) {
       """

### 2018-12-14 lib/magritte/code.rb
interrupt properly on missing closure variables

@@ -76,9 +76,14 @@ module Magritte
     end
 
     def collect
+      collection, status = collect_with_status
+      collection
+    end
+
+    def collect_with_status
       c = Collector.new
-      into(c).call
-      c.collection
+      status = into(c).call
+      [c.collection, status]
     ensure
       # make sure not to wait for close until at least the main thread
       # has returned, otherwise we sleep the root thread. in a case with

### 2018-12-14 lib/magritte/interpret.rb

@@ -94,6 +94,9 @@ module Magritte
       def visit_lambda(node)
         free_vars = @free_vars[node]
         yield Value::Function.new(node.name, Proc.current.env.slice(free_vars), node.patterns, node.bodies)
+        Status.normal
+      rescue Env::MissingVariable => e
+        error!(e.to_s)
       end
 
       def visit_assignment(node)

### 2018-12-14 lib/magritte/status.rb

@@ -74,6 +74,10 @@ module Magritte
       property?(:fail) || property?(:crash)
     end
 
+    def crash?
+      property?(:crash)
+    end
+
     def repr
       out = ""
       out << "!" if fail?

### 2018-12-14 spec/interpret_spec.rb

@@ -6,10 +6,13 @@ describe Magritte::Interpret do
   let(:ast) { Magritte::Parser.parse_root(skel) }
   let(:env) { Magritte::Builtins.load(Magritte::Env.empty) }
   let(:results) { ast;
-    Magritte::Spawn.s_ env do
+    collection, @status = Magritte::Spawn.s_ env do
       Magritte::Interpret.interpret(ast)
-    end.collect.map(&:repr)
+    end.collect_with_status
+
+    collection.map(&:repr)
   }
+  let(:status) { results; @status }
   let(:result) { results.join("\n") }
 
   describe "simple input" do
@@ -245,6 +248,20 @@ describe Magritte::Interpret do
 
         it do
           assert { result == "103" }
+        end
+      end
+
+      describe "missing closured variables" do
+        let(:input) {
+          """
+          f = (=> put (=> put %x))
+          put should-have-crashed
+          """
+        }
+
+        it do
+          assert { status.crash? }
+          assert { results.empty? }
         end
       end
 

### 2018-12-20 lib/magritte/builtins.rb
Implement eq builtin

@@ -105,9 +105,6 @@ module Magritte
       Status.normal
     end
 
-    builtin :eq, [:any, :any] do
-    end
-
     builtin :add, [], :Number do |*nums|
       put Value::Number.new(nums.map { |x| x.value.to_f }.inject(0, &:+))
       Status.normal
@@ -165,6 +162,29 @@ module Magritte
 
     builtin :rand, [] do
       put Value::Number.new(rand)
+    end
+
+    builtin :eq, [:any, :any] do |lhs, rhs|
+      (rec_func = lambda { |lhs, rhs|
+        return bool(false) if lhs.class != rhs.class
+
+        case lhs
+        when Value::Number
+          return bool(lhs.value == rhs.value)
+        when Value::String
+          return bool(lhs.value == rhs.value)
+        when Value::Vector
+          return bool(false) if lhs.elems.size != rhs.elems.size
+          lhs.elems.zip(rhs.elems).each do |le, re|
+            return bool(false) if rec_func.call(le, re).fail?
+          end
+          return bool(true)
+        when Value::Channel
+          bool(lhs.channel == rhs.channel)
+        else
+          raise "TODO: Implement equality for #{lhs.inspect}"
+        end
+      }).call(lhs, rhs)
     end
 
     builtin :gt, [:Number, :Number] do |than, is|

### 2018-12-20 lib/magritte/value.rb

@@ -43,7 +43,7 @@ module Magritte
       attr_reader :value
 
       def initialize(value)
-        @value = value
+        @value = value.to_f
       end
 
       def repr

### 2018-12-20 spec/interpret/cond_spec.rb

@@ -44,6 +44,7 @@ interpret_spec "conditionals" do
       c = (make-channel)
       eq $c $c || put fail10
       eq $c (make-channel) && put fail11
+      true
     EOF
 
     results []

### 2018-12-21 lib/magritte/lexer.rb
Allow underscore in barewords

@@ -190,7 +190,7 @@ module Magritte
       elsif scan /'(.*?)'/m
         skip_ws
         return token(:string, group(1))
-      elsif scan /([a-zA-Z\-0-9]+)/
+      elsif scan /([_a-zA-Z\-0-9]+)/
         skip_ws
         return token(:bare, group(1))
       else

### 2018-12-21 lib/magritte/ast.rb
Implement pattern matching in anon lambdas

@@ -21,9 +21,16 @@ module Magritte
       defdata :value
     end
 
+    class StringPattern < Tree::Node
+      defdata :value
+    end
+
     class VectorPattern < Tree::Node
       deflistrec :patterns
       defopt :rest
+    end
+
+    class DefaultPattern < Tree::Node
     end
 
     class Lambda < Tree::Node

### 2018-12-21 lib/magritte/parser.rb

@@ -178,6 +178,30 @@ module Magritte
       return out
     end
 
+    def parse_pattern(pattern)
+      pattern.match(~token(:string)) do |s|
+        return AST::StringPattern[s.value]
+      end
+
+      pattern.match(~token(:bare)) do |b|
+        if b.value == "_"
+          return AST::DefaultPattern[]
+        else
+          return AST::StringPattern[b.value]
+        end
+      end
+
+      pattern.match(~token(:bind)) do |b|
+        return AST::Binder[b.value]
+      end
+
+      pattern.match(nested(:lbrack,~_)) do |vec|
+        return AST::VectorPattern[vec.elems.map { |e| parse_pattern(e) }, nil]
+      end
+
+      error!(term, "unrecognized pattern")
+    end
+
     def parse_term(term)
       term.match(~token(:var)) do |var|
         return AST::Variable[var.value]
@@ -226,11 +250,7 @@ module Magritte
             end
           end
 
-          unless bindings.all? { |i| i.elems.all? { |e| e.token?(:bind) }}
-            error!(term, "TODO: Support patterns")
-          end
-
-          patterns = bindings.map { |i| AST::VectorPattern[i.elems.map { |e| AST::Binder[e.value] }, nil]}
+          patterns = bindings.map { |i| AST::VectorPattern[i.elems.map { |e| parse_pattern(e) }, nil] }
           return AST::Lambda[name, patterns, bodies.map { |body| AST::Group[body.map { |line| parse_line(line) }]}]
         end
 

### 2018-12-21 lib/magritte/pattern.rb

@@ -30,6 +30,15 @@ module Magritte
         @env.let(node.name, val)
       end
 
+      def visit_string_pattern(node, val)
+        fail! unless val.is_a?(Value::String)
+        fail! unless node.value == val.value
+      end
+
+      def visit_default_pattern(node, val)
+        #pass
+      end
+
       def visit_vector_pattern(node, val)
         fail! unless val.is_a?(Value::Vector)
 

### 2018-12-21 spec/interpret/lambda_spec.rb

@@ -120,6 +120,32 @@ interpret_spec "lambda functionality" do
     results %w(10 1 10 2 10 3)
   end
 
+  interpret "vector patterns" do
+    source <<-EOF
+      f = (
+        [one ?x] => put [first %x]
+        [two ?y ?z] => put [second %y %z]
+        [three] => put [third]
+        [four ?q to ?w] => put [fourth %q %w]
+        _ => put default
+      )
+
+      f [one 1]
+      f [two 2 3]
+      f [three]
+      f [four 4 to 5]
+
+      # default cases
+      f [one 1 2]
+      f [two]
+      f [three 4]
+      f [four 5 6]
+    EOF
+
+    results ["[first 1]", "[second 2 3]", "[third]", "[fourth 4 5]",
+             "default", "default", "default", "default"]
+  end
+
   interpret "variable arguments" do
     source <<-EOF
       (?x (?rest) => put %x; put hello; put %rest) 1 2 3

### 2019-01-18 doc/paper.tex
paper updates

@@
-% notes
-\documentclass[english,PRO]{ipsj}
-\usepackage{PROpresentation}
-\PROheadtitle{y-n-(x): Manuscript for presentation at IPSJ-SIGPRO, d m Y.}
+% default is sigpro, define \sigplan to switch
+
+\ifx\sigplan\undefined
+\newcommand{\ifsigplan}[1]{}
+\newcommand{\ifsigpro}[1]{#1}
+\else
+\newcommand{\ifsigplan}[1]{#1}
+\newcommand{\ifsigpro}[1]{}
+\fi
+
+\ifsigpro{ \documentclass[english,PRO]{ipsj} }
+\ifsigplan{ \documentclass[sigplan,9pt]{acmart} }
 
@@
 \usepackage{auto-pst-pdf}
+% \usepackage[symbol]{footmisc}
 
@@
 
+\ifsigpro{
+\usepackage{PROpresentation}
+\PROheadtitle{2018-4-(5): Manuscript for presentation at IPSJ-SIGPRO, 17 Jan 2019.}
+}
+
+\ifsigplan{
+\acmDOI{}
+\copyrightyear{2019}
+\acmISBN{}
+\acmConference[PASS Workshop]{}{April 2019}{Genova, Italy}
+\setcopyright{none}
+\titlenote{This work was presented at the 122nd IPSJ SIGPRO workshop on January 17, 2019.
+The presentation was not refereed, and the same manuscript was distributed only to the participants to the workshop.}
+}
+
 \begin{document}
+
 \title{A Shell-like Model for General Purpose Programming}
 
+\ifsigpro{
 \affiliate{TITech}{%東京工業大学情報理工学院数理・計算科学系 \\
@@
 \author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]
+}
+
+\ifsigplan{
+\author{Jeanine Miller Adkisson}
+\affiliation{%
+  \institution{Tokyo Institute of Technology, Japan}
+}
+\email{jneen@jneen.net}
+
+\author{Johannes Westlund}
+\affiliation{%
+  \institution{Tokyo Institute of Technology, Japan}
+}
+\affiliation{%
+  \institution{KTH Royal Inst. of Technology, Sweden}
+}
+\email{jwestlun@kth.se}
+
+\author{Hidehiko Masuhara}
+\affiliation{
+  \institution{Tokyo Institute of Technology, Japan}
+}
+\email{masuhara@acm.org}
+}
 
@@
 \begin{abstract}
-Shell languages such as bash are designed to integrate with an OS, which mainly involves managing processes with implicit input and output streams.
+Shell scripting languages such as bash are designed to integrate with an OS, which mainly involves managing processes with implicit input and output streams.
 They also attempt to do this in a compact way that could be reasonably typed on a command-line interface.
@@
 By way of a new programming language, \emph{Magritte}, we propose a general-purpose programming language with semantics similar to bash.
-In this paper, we discuss the early design of such a system, in which the primary unit of composition, like bash, is processes with asynchronous inputs and outputs, which can be read from or written to at any time, and which can be chained together via a pipe operator.
+In this paper, we discuss the early design of such a system, in which the primary unit of composition, like bash, is processes with input and output channels, which can be read from or written to at any time, and which can be chained together via a pipe operator.
 We also explore concurrency semantics for such a language.
@@
 
+\settopmatter{printfolios=true}
 \maketitle
@@
 \noindent
-The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes.
-Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no boundaries between values.
+The UNIX shell programming model has played an important role in integrating applications and operating systems by composing programs---spawning independent programs in parallel to communicate over operating-system pipes.
+Beyond the most simple tasks, however, bash and similar tools break down, due to various language deficiencies.
 
-In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.
+In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes which communicate through synchronous channels or pipes in an ad-hoc manner---and in which this facility is the primary method of composing different functions or units.
 
@@
 \subsection*{Outline}\noindent
-In this work, we present a new shell language \emph{Magritte}.
+In this work, we present a new pipe-based language \emph{Magritte}.
 Section 2 describes design requirements for a language in this space.
 Section 3 and 4 describe Magritte's design and implementation.
-Section 5 describes our future goals with this project, and in Section 6 we review various related works and alternative approaches.
+Section 5 describes our future goals with this project, and in Section 6 we review various related work and alternative approaches.
 
@@
 \subsection{Programming With Values}
-\subsubsection{Not Everything Is a String}\noindent
-The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.
 
-Several projects\cite{haahr,duff}, including bash itself, have attempted to add lists and hash maps as standard objects.
-Unfortunately, while they allow these values to be stored in variables, most systems do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
-Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not possible.
+\subsubsection{Value Pipes}\noindent
+It is desirable in a general purpose language to be able to create and use complex data structures, and to use them freely throughout the language.
+In particular, if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through pipes.
 
-Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.
+Several projects\cite{haahr,duff}, including bash itself, have attempted to add arrays and associative arrays as standard objects, but not as first class values: they do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
+%These values are therefore relegated to a second class of value, one for which pipe-based composition is not possible.
 
-\subsubsection{Consistent Channels}\noindent
-Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are used for performance reasons.
-However, it is important if a pipe is to be used for composition that reads correspond exactly to writes, and that data does not become corrupted by interleaving.
+Furthermore, serialization through pipes is generally not possible except in a single-threaded case, since complex data can be corrupted through interleaving.
 
@@
 This might be expressed in bash as:
-\noindent\begin{lstlisting}[morekeywords={do,done}]
+\begin{lstlisting}[morekeywords={do,done}]
 # <@\underline{\color{gray} consume}@> the stream, labeling every line
@@
 
-# process the stream in the background
-# with two threads
+# process the stream with two threads
 split() { label a & label b & ;}
@@
 \end{lstlisting}
+With this code, a user might expect the output to be three lines, each consisting of a letter \verb/a/ or \verb/b/, and a number, for example:
 
-\noindent
-With this code, a user might expect the output to be three lines, each consisting of a letter \verb/a/ or \verb/b/, and a number, for example:
+% \footnotetext{hello world}
 
@@
 \noindent
-Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved, resulting in outputs such as:
+Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved\footnotemark, resulting in outputs such as:
+\footnotetext{We have observed some behavior in the output that cannot be explained simply by interleaving, suggesting there may be some other race conditions in play.}
 
@@
 a4
+
 \end{lstlisting}
@@
 In order to provide a channel implementation that is usable for large programs, we must ensure that it is \emph{consistent}---that is, that every read corresponds to exactly one write.
+Value pipes accomplish this by making a value the smallest atomic unit of communication.
 
-\subsubsection{Return Semantics}\label{return-semantics}\noindent
-Pipeline elements such as these do not generally return values---instead, they output data to their output stream.
-However, for serious programs it is desirable to capture some kind of return value from functions.
-Es Shell\cite{haahr} accomplishes this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.
+% cat $(ls | grep ...)
+\subsubsection{Capture and Substitution}\label{return-semantics}\noindent
+For processes that output values, we need to support a mechanism to capture those values for use in variables, data structures, and function arguments, similar to backticks or \verb/$(...)/ in bash.
+This enables processes to be used like functions, which return data to their caller.
 
-If we allow values to be written to output streams, however, this approach is redundant---a function author must decide whether to return a value or write it to the standard output.
-Thus it is desirable to have an interpolation syntax which \emph{collects} the output of a command synchronously.
+But since in most shells, rich values are not writable to output streams, the output cannot effectively be used as a return path for values.
+Es Shell\cite{haahr} accounts for this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.
 
-A priori, such a collection mechanism would have to return an list, as any function call may output zero or more values.
-However, that approach would require the callers of functions to manually unwrap lists on every function call.
+If we allow values to be written to output streams, however, we can directly capture values written to the output.
 
-\subsection{Language Features}
-\subsubsection{Variable Scope}\noindent
-Like the Es shell, we also wish to provide lambda functions with lexically scoped variables.
-However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables happen to be bound in the function's definition environment.
-Therefore a simple mistype of a variable name will result not in an error, but in the variable silently becoming dynamically bound.
+\subsubsection{Modern Language Features}
+%Suggestion: without using subsubsections, list the features as a list (by using the itemize environment). And add first class functions into the list, and a prototype-based lookup mechanism.
 
-\subsubsection{Objects}\noindent
-For large programs, an object system is desirable as a way to provide user-defined types and abstractions.
+Users will expect a modern programming language to have:
+%Language features required for a modern general purpose language include:
+
+\begin{itemize}
+  \item \emph{Lambda functions with closure}.
+    This requires the introduction of lexically scoped variables.
+  \item \emph{Dynamic variables}.
+     Most shell languages already include these, as OS Environment variables are dynamic by nature.
+  \item \emph{Product structures with support for open recursion}.
+    a prototype-based object system is sufficient for this.
+  \item \emph{Sum structures}. In an untyped language, a method for pattern matching over nestable heterogenous lists is sufficient.
+\end{itemize}
+
+% For large programs, an object system is desirable as a way to provide user-defined types and abstractions.
 The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer.
@@
 
-\subsubsection{Trees and Pattern Matching}\noindent
-For programs that deal with syntax or other kinds of trees, it is convenient to have a compact tree representation, which can be made possible with nestable vectors.
-It is also desirable to have some mechanism for pattern matching against these kinds of values.
-These kinds of structures are impossible using the list features in systems such as RC Shell\cite{duff} and Es Shell, which implicitly flatten all lists.
+% For programs that deal with syntax or other kinds of trees, it is convenient to have compact nestable data structures that support pattern matching, which can be made possible with nestable vectors.
 
 \subsection{Automatic Process Cleanup}
-\subsubsection{Interruption}\noindent
-A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
+\subsubsection{Interruption}
+\noindent
+In shell programming, we tend to compose infinitely running processes together as pipeline elements.
+Thus we require a well-defined semantics of process \emph{interruption}\footnotemark: automatic clean-up of processes that will no longer be used.
+Consider the following Magritte code:
+\footnotetext{We use the term \emph{interruption} in the generic sense---to mean the halting of normal flow in a process, due to some external event. It is unrelated to hardware or OS signalling.}
 \begin{lstlisting}
@@
 \noindent
-In this example, three processes are spawned: (1) a process with an open file that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
+In this example, three processes are spawned concurrently: (1) a process with an open file that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
 A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.
 
-In a pipe-based language, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
 A user will also expect that, after the first 10 lines are processed and the \verb/take/ function returns, the file will be closed, and all three processes exit.
-This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.
+This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is an infinite loop.
 
 In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
-Similarly, the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.
+Similarly, the call to the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.
 
 \subsubsection{Compensation}\noindent
-In order to observe interruption, user code must be able to react to and handle interruption.
-The most direct way to observe interruption is through a system of \emph{compensation actions}, which are expressions which will be evaluated in the case of an interruption.
+In interacting with an operating system, it is necessary to mange side effects, and gracefully recover or restore state in the case of an interruption.
+Such an error-handling system would also be a way for user code to directly observe interruption.
 
-Compensations are the only way in which interruptions can be directly observed by user code---without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or to observe the memory footprint of the program.
+Without compensation of errors, the only way of observing forever-blocked processes would be to inspect the process table, or to observe the memory footprint of the program.
 Thus, we can define the primary task of the interruption system as \emph{running a process's compensation actions at the appropriate time}.
@@
 \subsubsection{Lazy Interruption vs. Eager Interruption}\label{lazy-interruption}\noindent
-In designing an interruption semantic, we must first decide whether processes are eagerly interrupted when their channels close, or only interrupted on channel operations.
+In a system such as this, there is a language design choice that must be made---whether interruption is \emph{eager}, in which processes are interrupted immediately upon closing a channel, or \emph{lazy}, in which processes are only interrupted upon interaction with a closed channel.
+Both approaches have advantages and drawbacks.
+In making this choice, we desire that the behavior of interruption be \emph{predictable}---however, there are two competing viewpoints for predictability.
 
@@
 \noindent
-The first process will write three times, and then continue to do other processing in-thread that does not write to the standard output.
+The left-hand side of the pipe will write three times, and then continue to do other processing in-thread that does not write to the standard output.
 The final \verb/take 3/ operation will return after 3 inputs are read.
-It is important to decide, then, whether the first process should be \emph{interrupted} in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or whether it should be left alone until it attempts to write a value (\emph{lazy interruption}).
+It is important to decide, then, whether the process on the left should be interrupted in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or whether it should be left alone until it attempts to write a value (\emph{lazy interruption}).
 
-An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted---those points which run a \verb/put/ or a \verb/get/.
-Whereas with eager-interruption semantics any point in the code may be interrupted, introducing the need for users to either mark critical uninterruptible sections or be very careful about setting correct compensations in the case of mutative algorithms.
+From the perspective of someone spawning a process, eager interruption can seem more predictable, as they can guarantee a point at which the process has stopped doing work.
+
+However, from the perspective of a function author, lazy interruption is more predictable, because the author can identify precisely which points in the code have the potential to be interrupted---those points which run a \verb/put/ or a \verb/get/.
+Contrast this with eager-interruption semantics, where any point in the code may be interrupted, introducing the need for users to either mark critical sections and be very careful with implementing stateful algorithms.
 
 On the other hand, lazy interruption has the disadvantage that a producing process must do enough work to produce one more value than will be consumed.
-If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in doing a large amount of unnecessary work.
+If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in a large amount of unnecessary work.
 
-\subsubsection{Non-closing of Channels}\label{non-closing}\noindent
+% sensei comments:
+% I see. Let's delete this paragraph, and improve the first sentence.
+
+% The problem was, while Section 2 is supposed to list the requirements, Section 2.3.3 merely points out existence of two viewpoints.
+It made me feel like "well I understand that you see two viewpoints, then which one you require?"
+
+% I think the requirement wrt eager/lazy interruption is that the language should support both interruption semantics and let the programmer use one of them selectively.
+\subsubsection{Closing of Channels}\label{closing}\noindent
 Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel.
@@
 Therefore the appropriate time for it to be interrupted is when it is blocked on a channel that will never receive any more operations.
-In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.
 
-Luckily, in pipe-based languages, the arrangement of pipes is usually specified at process spawn time.
-This enables us to maintain a process register inside of each channel, so that we can decide when all readers or all writers have detached from a channel, at which time the channel closes and interrupts all blocked processes.
+This can be difficult to detect when a reference to a channel can be passed anywhere in the program as a standard value.
+Luckily, the arrangement of pipes and channels are usually specified at process spawn time.
+We can therefore relax our constraint to guarantee that channels will be closed at appropriate times in \emph{common architectures}, and that they never close if there are active readers or writers.
 
-\subsubsection{Closing Permanence}\label{closing-permanence}\noindent
-Even when we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel.
-In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use.
-In UNIX, for example, the named-pipes API allows for persistent channels \cite{linux-pipe} that, being merely files, can be reopened at any time.
-
-However, this can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:
+\subsubsection{Reopening Channels}\label{closing-permanence}\noindent
+Some systems, like UNIX named pipes, allow a channel to be re-used by new processes after it has been closed\cite{linux-pipe}.
+This is, however, not a desirable feature, as it can lead to some unexpected races between a channel closing and a new process spawning.
+Consider the example:
 \begin{lstlisting}
-c = (make-channel)
-& count-forever > $c
-& take 10 < $c
-put 10 > $c
+c = (make-channel)     # create a new channel
+& count-forever > $c   # write infinitely
+& take 10 < $c         # read 10 elements and exit
+put 10 > $c            # write once from a new process
 \end{lstlisting}
@@
 \noindent
-This example, while contrived, represents an unavoidable race with first-class channels---whether the process running \verb/take 10/ closes the channel before or
-after the \verb/put 10/ invocation opens the channel for writing.
+This example represents an unavoidable race with first-class channels: between \verb/take 10/ closing the channel and \verb/put 10/ opening the channel for writing.
 If we allow channel reopening, we will either block forever, or insert the number 10 into the stream, depending on which happens first.
@@
 \subsubsection{Masking Interruptions}\label{masking}\noindent
-A user will often wish to continue after a particular output channel has closed, and in this case will need a facility to mask exceptions for particular channels.
-Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
+We must include a facility to control the extend of interruptions. This is because an interruption semantics can make it difficult to perform final calculations after a channel is closed.
+
+% A user will often wish to perform final calculations after a particular output channel has closed, and interruption semantics can make this difficult.
+Consider the following example, which sums all numbers from the standard input:
+% Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
+% \begin{lstlisting}
+% (sample) = (
+%   hold = (get)
+%   i = 1
+%   each (?v =>
+%     $i = (inc $i)
+%     # mutate with probability 1/$i
+%     prob (recip $i) 1 && ($hold = %v)
+%   )
+% 
+%   put $hold
+% )
+% \end{lstlisting}
+
+% \noindent
+% In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability.
+After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.
 \begin{lstlisting}
-(sample) = (
-  hold = (get)
-  i = 1
-  each (?v =>
-    $i = (inc $i)
-    # mutate the $hold variable
-    # with probability 1/$i
-    prob (reciprocal $i) 1 && ($hold = %v)
+(sum) = (
+  total = 0
+  each (?x => %total = (add %total %x))
+  put %total
 )
-
-  put $hold
-)
-\end{lstlisting}
-
-\noindent
-In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability.
-After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.
+\end{lstlisting}\noindent
+In this example, the function \verb/each/ will consume the input stream and mutate a lexical variable (marked with \verb/%/).
+After the entire input stream is consumed, we wish to output the resulting \verb/%total/ value.
 
 However, with the semantics described above, the \verb/each/ function will loop until it receives an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line.
-In order to avoid this situation, the \verb/each/ function must somehow indicate that it intends to consume the entire stream, and handle the interrupt signal.
 
@@
 \subsubsection{Lexical vs. Dynamic Variables}\noindent
-In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
-Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g.
-\verb/x = 1/), so that we can maintain compatibility with standard environment files.
-Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
-\verb/?x/) to allow for a planned extension for pattern matching.
+In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at variable reference and mutation points, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
+Let-binding is left unadorned (\verb/x = 1/), and will bind a variable both lexically and dynamically.
+
+Let-binding uses this plain syntax to maintain compatibility with standard environment files.
+Variable binders in lambda arguments use \verb/?/ (e.g.
+\verb/?x/) to allow for unambiguous pattern matching, and are also bound both lexically and dynamically.
 
@@
 
+% In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/).
+Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g.
+\verb/x = 1/), so that we can maintain compatibility with standard environment files.
+Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g.
+\verb/?x/) to allow for unambiguous pattern matching.
+
+% For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g.
+\verb/%x = 1/ or \verb/$x = 1/.
+In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
 \subsubsection{Lambda Functions}\noindent
 Lambda functions are specified with parenthesized expressions containing the arrow symbol \verb/=>/, which separates the bindings from the body.
-Multiple matching clauses are possible, and will be extended in the future with more general pattern matching.
-Clauses are separated at the beginning of lines that contain \verb/=>/, which avoids the need for any indentation-based parsing.
+Multiple matching clauses are possible, and can match simple patterns.
+Clauses are separated at the beginning of lines that contain \verb/=>/ (where "lines" are also terminated by \verb/;/, and do not consider nested newlines), which avoids the need for any indentation-based parsing.
 For example:
@@
 \noindent
-Functions are scanned statically for free lexical variables, which are captured at function definition time, and merged into the environment at call time.
+Functions maintain the scope of any free lexical variables in the body, including for mutation.
 
 \subsubsection{Nestable Vectors}\noindent
-We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/. Combined with bareword strings, we can represent trees in a straightforward manner:
+We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/.
+Combined with bareword strings (string literals without quotation), we can represent trees in a straightforward manner:
 \begin{lstlisting}
@@
 \noindent
-We discuss a pattern matching extension in \emph{Future Work}.
-A typical strategy for traversing this kind of structure might be a function that put all leaf node values to its output in a predefined order, to be consumed by another process.
+These can be matched in lambda arguments by patterns such as \verb/[node ?x]/.
+A typical strategy for traversing this kind of structure might be a function that puts all leaf node values to its output in a predefined order, to be consumed by another process.
 
-The builtin function \verb/for/ takes a vector as an argument and outputs each element in order, so that a normal vector can be traversed by the pattern:
+The builtin function \verb/for/ takes a vector as an argument and outputs each element in order, so that a vector can be traversed with:
 \begin{lstlisting}
@@
 \subsubsection{Environments as Objects}\noindent
-We support environment capture with the straightforward \verb/{ }/ syntax.
-The semantics of curly braces are to run the code as normal, then detach the running environment from its parent and return it.
+Variable environments can be captured directly as key-value maps with a parent pointer.
+Environment capture uses \verb/{ }/ syntax to run a block of code, then remove the running environment from its parent list.
+The resulting environment value is substituted in place of the \verb/{ }/ expression.
 Therefore all assignment syntaxes are available, including function definition.
@@
  balance = $balance
-  (deposit ?amt) =
-    (%balance = (add %amt %balance))
-  (withdraw ?amt) = (
-    (%balance = (sub %amt %balance))
+ (deposit ?amt) = (%balance = (add %amt %balance))
+ (withdraw ?amt) = (%balance = (sub %amt %balance))
 }
@@
 \noindent
+A planned extension would allow using special syntax to register additional parents, allowing users to inherit by direct delegation.
+
 Environment lookup uses the \verb/!/ symbol, as in \verb/$env!key/.
-We use this symbol because it is already reserved by bash, unlike both \verb\/\ or \verb/./ which need to be available in bareword syntax to integrate with external programs.
+We use this symbol because it is already reserved by bash, unlike both \verb\/\ or \verb/./ which need to be available in bareword syntax to indicate file paths for external programs.
 Environments can also be mutated using the same access syntax.
@@
 
-\subsubsection{Reading and Writing}\noindent
-Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively. These block until they complete a communication.
+% \subsubsection{Reading and Writing}\noindent
+% Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively. These block until they complete a communication.
 
 \subsubsection{Collection and Substitution}\noindent
-In order to use output values seamlessly as return values without unwrapping arrays at every function call as described in Section \ref{return-semantics}, Magritte uses normal parentheses to collect and \emph{expand} the resulting values into the current command vector---increasing the argument number by the number of values output from the function.
+A priori, a capture mechanism such as described in Section \ref{return-semantics} would have to return a list, as any process may output zero or more values.
+However, as a function calling convention, that would require callers to manually unwrap lists on every function call.
+
+In order to simplify substitution, Magritte uses normal parentheses to collect and \emph{expand} the resulting values into the current command vector---increasing the argument number by the number of values output from the function.
 For example:
@@
 
-# Parentheses collect the three writes and expand
-# 1 2 3 in-place
+# Collect three writes and expand 1 2 3 in-place
 other-fn 0 (count-three) 4
@@
 \noindent
-These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that because we write transactional values to the output, we have the ability to properly separate values without relying on whitespace.
+These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that we have the ability to properly separate values without relying on whitespace.
 
@@
 \begin{lstlisting}
-produce | (process; process; process)
+generate-values | (process; process; process)
 \end{lstlisting}
@@
 In the case that a user might want to use a substitution at the root level---i.e.
-to generate a function and immediately call it, we provide the \verb/exec/ builtin function which executes its arguments as a normal command.
+to generate a function and immediately call it, we provide the \verb/exec/ builtin which executes its arguments as a command.
 
@@
 \subsubsection{Spawning and Redirecting}\noindent
-Spawning uses a syntax similar to bash, with an ampersand (\verb/&/) at the beginning of a command indicating that it should be spawned in a new thread.
+Spawning uses an ampersand (\verb/&/), similar to bash, at the \emph{beginning} of a command indicating that it should be spawned in a new thread.
 
@@
 \subsection{Interruption and Compensation}
+
 \subsubsection{Lazy Interruption}\noindent
-Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for the extra-values problem discussed in Section \ref{lazy-interruption}, and we discuss some features that may mitigate this problem in \emph{Future Work}.
+We have decided that the predictability of lazy interruption is worth the tradeoff for the extra-values problem discussed in Section \ref{lazy-interruption}, and we discuss some ways to mitigate this problem in Section \ref{discussion}.
+
+\subsubsection{Process Registration}\label{process-registration}
+In order to satisfy the constraints of Section \ref{closing}, we maintain a process register inside of each channel, so that we can decide when all readers or all writers have returned or been interrupted, at which time we can guarantee that processes on the other side of the channel cannot be woken up \emph{without spawning new processes}.
+Therefore this covers the architecture of pipelines, in which many processes are spawned together in fixed configurations.
+Other architectures will have to manually manage process shutdown in some cases.
 
 \subsubsection{Compensation and Unconditional Compensation}\noindent
-We implement compensation directly using the \verb/%%/ operator to indicate a compensation action, which is cleared at the end of the current function body:
+ We employ a variant of the compensation mechanism introduced by Inoue et al.\cite{compensation}
+using the \verb/%%/ operator to indicate a compensation action, which is run in case of an interruption in the left hand side or subsequent lines.
+Compensations are cleared at the end of the current function body:
 \begin{lstlisting}
 (my-function) = (
-  # ...
   action %% cleanup-action
-  # ...
+  # in effect until the end of the function
 )
@@
 In this way, they are analogous to \verb/finally/ or \verb/ensure/ sections of standard exception handling.
-
 For example, the function \verb/read-lines/ above could be implemented as:
 \begin{lstlisting}
-(read-lines ?fname) = (
-  f = (open-file $fname) %%! close-file $fname
-  until (=> eof? $f) (=> read-until "\n" $f)
-)
+(read-lines ?fname) =
+ (f = (open-file $fname) %%! close-file $fname
+  until (=> eof? $f) (=> read-until "\n" $f))
 \end{lstlisting}
@@
 \subsection{Channel Registry}\noindent
-Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described in Section \ref{non-closing}.
-When \verb/remove_reader/ or \verb/remove_writer/ result in an empty registered set, they additionally close the channel and raise an internal \verb/Interrupt/ exception in every blocked thread.
+Our channel implementation is a standard implementation of synchronous channels, with the addition of four intrinsic methods, used only internally by the interpreter: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, which register and deregister processes as described in Section \ref{process-registration}.
+When a \verb/remove_*/ method results in an empty set, it will additionally close the channel and raise an internal exception in every blocked thread.
 
-Once the channel is closed, every call to \verb/read/ and \verb/write/ will cause the calling process to be interrupted, as described above in Section \ref{closing-permanence}.
+Once the channel is closed, every call to \verb/read/ and \verb/write/ will interrupt the calling process as described in Section \ref{closing-permanence}.
 
@@
 
-Cascading of interruptions then happens naturally---when a channel closes, a process is interrupted, causing it to unwind its stack and deregister channels, thereby potentially causing other channels to close.
+Interruptions then cascade naturally---when a channel closes, a process is interrupted, causing it to unwind its stack and deregister channels, thereby potentially causing other channels to close.
 
@@
 It is necessary to take some care with the implementation of the spawning primitive (\verb/&/), that we wait until the spawned process has finished registering its channels before the spawning process continues.
+% For this reason the root frame of a process always has its channels registered by the \emph{spawning} process before it is allowed to start.
 Consider the following example, which outputs 10 numbers, possibly out of order:
@@
 However, since they are both spawned in the background, the spawning process will immediately return.
-In general this is not a problem, since the \verb/drain/ processes should be keeping the two pipes open.
+In general, the \verb/drain/ processes should be keeping the two pipes open.
 However, if we do not take care to wait until they have finished registering their channels, there is a risk that the spawning process will return first and close the two pipes.
 
-For this reason, we have special case handling for the root frame of a process---the channels are registered in the \emph{spawning} process, before the spawned process is allowed to start.
-
 \subsection{Collectors}\noindent
-In order to implement the return semantics described in Section \ref{return-semantics}, we also implement a write-only channel called a \emph{collector}.
-Collectors cannot be created directly by users, but only appear in the interpreter when we evaluate a \verb/Subst/ node, which represents parentheses in argument position.
+In order to implement the return semantics described in Section \ref{return-semantics}, we also implement a write-only channel called a \emph{collector}, and an intrinsic that waits for channel closing.
+Collectors cannot be created directly by users, but only appear in the interpreter when we evaluate parentheses in argument position.
 
@@
 Upon closing the channel, all elements of this waiting set are awoken, and flow continues.
-Thus the algorithm for substiution is:
-\begin{lstlisting}
+Thus the algorithm for substituting a block is:
+\begin{lstlisting}[language={}]
 * Create a new collector $c
-* Evaluate the contents with standard output set to the collector
+* Run the parenthesized expressions with standard output set to the collector
 * Run $c.wait_for_close
@@
 
-\section{Future Work}\label{future-work}
+\section{Discussion}\label{discussion}
 
 \subsection{Open Checking}\noindent
-Because our system implements \emph{lazy interruption}, we must somehow mitigate the problem of producers creating one extra element than is needed, in the case that elements are expensive to produce.
-One possibility is to introduce the concept of \emph{open checking}, so that producer processes can periodically run a command that will interrupt in the case that the output is closed, without actually writing any data to the channel.
+Because our system implements \emph{lazy interruption}, we must mitigate the problem of producers creating one extra element than is needed, in the case that elements are expensive to produce.
+One possibility is to introduce the concept of \emph{open checking}, so that producer processes can periodically run a builtin command that will interrupt if the output is closed, without actually writing any data to the channel.
 
-\subsection{Pattern Matching}\noindent
-Though the syntax design is not yet complete, we plan to allow pattern matching in all lambda bindings.
+% \subsection{Pattern Matching}\noindent
+% Though the syntax design is not yet complete, we plan to allow pattern matching in all lambda bindings.
 In order to potentially support glob and regular expression matching, we must be careful which symbols we reserve in this context.
 
-\subsection{Prototypical Inheritance}\noindent
-One missing feature to support prototypical inheritance is the dynamic assignment of a special \verb/$self/ variable.
+% \subsection{Prototypical Inheritance}\noindent
+% One missing feature to support prototypical inheritance is the dynamic assignment of a special \verb/$self/ variable.
 We plan to implement a system similar to JavaScript, in that the \verb/$self/ variable is assigned dynamically based on the calling syntax, or overrideable with special-purpose functions.
 
-Additionally, we plan to support multiple inheritance by allowing for an environment to point to a list of parents, and resolve names in a depth-first manner.
-
-\subsection{UNIX Integration}\noindent
-In order to be viable as a shell, we must integrate seamlessly with a UNIX environment.
+\subsection{OS Integration}\noindent
+In order to be viable as a shell, we must integrate seamlessly with a POSIX-like environment.
 This involves marshalling values to strings and byte streams in order to communicate with external processes, and allowing external processes to interact with Magritte values.
@@
 
+\ifsigpro{
 \subsection{Macros}\noindent
@@
 Our parsing technique involves an intermediate \emph{skeleton tree} representation adapted from Bachrach and Playford\cite{skeleton-trees} that makes macro systems simpler to implement in non-lisp syntaxes.
+}
 
 \subsection{Performance}\noindent
-The current implementation, being a Ruby interpreter, is slow. We plan to bootstrap using a JIT-compiled virtual machine implemented in RPython.
+The current Ruby interpreter is slow. We plan to bootstrap using a JIT-compiled virtual machine implemented in RPython.
+
+\subsection{Desktop Scripting}\noindent
+We believe Magritte offers a promising solution to the \emph{desktop scripting problem}: an interface for end users to integrate and automate independent programs, a problem attempted by projects such as Guile Scheme\cite{guile} and TCL\cite{tcl}.
+What differentiates our approach is that our platform should be able to integrate with programs as they are currently written, without the need for further integration on the part of application developers.
 
@@
 
-\subsection{Iterators}
-\noindent
-Many general purpose languages such as Python implement an iterator API, for code that is intended to produce values in a streamlike fashion.
+% * pipe objects
+% * lightweight syntax
+% * data structures
+
+% \subsection{Iterators}
+% \noindent
+% Many general purpose languages such as Python implement an iterator API, for code that is intended to produce values in a streamlike fashion.
 For example, in a Python Enhancement Proposal (PEP), Smith\cite{pep533} describes an automatic closing system for generators, in which an \verb/__iterclose__/ signal is sent by the consumer back to the producer as soon as a consumer exits early.
 
-But channel-based systems generally allow for multiple concurrent readers and writers---so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
+% But channel-based systems generally allow for multiple concurrent readers and writers---so automatically cleaning up a channel when a single consumer exits would cause processes that are still able to communicate to be terminated early.
 
@@
 \noindent
-Many general purpose languages, including Go\cite{golang} and Clojure\cite{clojure}, contain robust channel implementations, which could be used to implement a wide variety of concurrent algorithms.
-For example, in an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
-\begin{lstlisting}[language=C, morekeywords={fn,channel,nil,go}]
-fn pipe(lhs, rhs, inp, out) {
-  c = channel()
-  go fn () { lhs(inp, c); };
-  rhs(c, out)
-}
+Many general purpose languages, including Go\cite{golang} and Clojure\cite{clojure}, contain robust channel implementations, which can be used to implement a variety of concurrent algorithms.
+In such a system we could implement something similar to our system by adding input and output channels to every function and heavily currying functions.
 
-collect = collector()
-pipe(curry(range, 10), curry(fan, 2, fn(el, inp, out) { write(el + 1, out); }), nil, collect);
-
-return collect.values()
-\end{lstlisting}
-\noindent
+% For example, in an untyped language similar to Go (the following implementation is not possible in general in Go without generic types), we could implement something similar to our system by adding input and output channels to every function and heavily currying functions:
+% \begin{lstlisting}[language=C, morekeywords={fn,channel,nil,go}]
+% fn pipe(lhs, rhs, inp, out) {
+%   c = channel()
+%   go fn () { lhs(inp, c); };
+%   rhs(c, out)
+% }
+% 
+% collect = collector()
+% pipe(curry(range, 10), curry(fan, 2, fn(el, inp, out) { write(el + 1, out); }), nil, collect);
+% 
+% return collect.values()
+% \end{lstlisting}
+% \noindent
 However, such a system would still require manual use of input and output channels when a simple pipeline would suffice---and the user must still translate between returning vs.
 outputting functions and expressions.
-Additionally, Go does not contain any kind of channel closing semantics or automatic process cleanup.
-Finally, such a system would be far too heavyweight for a REPL context, and would not make for a viable interactive shell.
+Additionally, these systems do not contain any kind of channel closing semantics or automatic process cleanup.
+% Finally, such a system would be far too heavyweight for a REPL context, and would not make for a viable interactive shell.
 
-\subsection{General-Purpose Language Extensions}
-\noindent
-Projects such as \verb/xonsh/\cite{xonsh} and \verb/scheme-shell/\cite{scheme-shell} attempt to extend the syntax of an existing general purpose language to include shell capabilities.
+\subsection{Shell Extensions} \noindent
+Many systems, such as \verb/xonsh/\cite{xonsh} and Es Shell\cite{haahr}, attempt to extend traditional shells with object or functional semantics, but do not sufficiently integrate their semantics with pipes, leaving pipes to remain bytes-only.
+
+Systems that extend existing languages with shell semantics, such as \verb/scheme-shell/, also do not integrate fully with pipes.
+
+% Projects such as \verb/xonsh/\cite{xonsh} and \verb/scheme-shell/\cite{scheme-shell} attempt to extend the syntax of an existing general purpose language to include shell capabilities.
 Unfortunately, this approach creates a schism between shell concepts and host language concepts---in particular, host language objects cannot be passed through pipes.
+
+% The \verb/es/ shell \cite{haahr}, which is an extension of the \verb/rc/ shell \cite{duff} from the Plan 9 operating system, attempts to extend the standard shell model with lambda functions with lexical scoping.
+These shell functions are marshalled into strings when passed through pipes (whether to shell functions or external programs), but are still written byte-by-byte and thus are subject to interleaving in concurrent cases.
 
@@
 Projects such as \verb/powershell/\cite{powershell} and \verb/mash/\cite{mash} allow objects (.NET objects and Scala objects, respectively) to be passed through pipelines.
-However, this approach still ties the language to one specific platform and makes it difficult to integrate with programs written for other platforms.
+However, this approach still ties the language to one specific platform or object system and makes it difficult to integrate with programs written for other platforms.
+By \emph{only} relying on the most universal OS concepts of standard input/output, argument vectors, and interrupt signals, we should be able to integrate programs across a wide variety of languages and platforms.
 
-\subsection{Desktop Scripting Languages}\noindent
-Projects such as \verb/Guile Scheme/\cite{guile} and \verb/TCL/\cite{tcl} attempt to create a separate API for integrating desktop programs, eliminating the need for shell entirely.
-However since this approach would require an unreasonably large integration effort on the part of application developers, their adoption has been limited in practice.
-Our goal is to create a system that can interact with programs in their current forms.
-
-\subsection{RC Shell}\noindent
-The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash.
-However its list type is not nestable---nested lists are automatically flattened at creation time.
-Additionally, these lists cannot be passed through channels.
-
-\subsection{Es Shell}
-\noindent
-The \verb/es/ shell \cite{haahr} attempts to extend the standard shell model with lambda functions with lexical scoping.
-These shell functions are marshalled into strings when passed through pipes (whether to shell functions or external programs), but are still written byte-by-byte and thus are subject to interleaving in concurrent cases.
+% Like the Es shell, we also wish to provide lambda functions with lexically scoped variables.
+However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables happen to be bound in the function's definition environment.
+Therefore a simple mistype of a variable name will result not in an error, but in the variable silently becoming dynamically bound.
 
@@
 This enables it to be used safely to orchestrate distributed computing tasks.
-However, the underlying pipe implementation remains byte-based, so that features like Magritte's collection and expansion remain impossible in general.
+However, the underlying pipe implementation remains byte-based, so that features like Magritte's collection and expansion remain impossible.
 
-\section{Conclusion}\label{conclusion}
+\section{Conclusion}\label{conclusion} \noindent
+We proposed Magritte, which is both capable of integrating closely with operating systems, and serving as a general-purpose language for larger applications.
+Our channel implementation allows for robust composition of concurrent algorithms, and can also serve as the fundamental unit of composition for our language, which integrates with modern general-purpose language features like rich data structures and lambda functionsn with lexical scoping.
 
-\noindent
-We believe that it should be possible to script, orchestrate, and interact with an OS environment in a language that is also capable of high-level abstraction.
-We believe that Magritte is a promising platform for shell scripting and for interaction, and that it is also viable for large programs.
+At the same time, the language stays close enough to the common OS spawning model that we believe it should be straightforward to integrate with an operating system.
+Because of this, we believe Magritte is capable of filling a currently unserved niche as a desktop integration language, and also provides an interesting platform for experimenting and building concurrent algorithms.
 
-Our channel implementation allows for robust composition of concurrent algorithms, and can also serve as the fundamental unit of composition for a general-purpose language.
+% hack: acmart apparently requires acks to be verbatim in the file,
+% so we alias acks -> acknowledgement for the ipsj template
+\ifsigpro{
+\newenvironment{acks}{\begin{acknowledgment}}{%
+\end{acknowledgment}\ignorespacesafterend
+}
+}
 
-\begin{acknowledgment}
-Johannes Westlund's contribution to this research was made possible by funding from The Sweden-Japan Foundation and Stockholms Grosshandelssocietet.
-\end{acknowledgment}
+\begin{acks}
+This work was supported by JSPS KAKENHI Grant Number 18H03219.  J. Westlund's contribution to this research was made possible by
+  %funding from
+The Sweden-Japan Foundation and Stockholms Grosshandelssocietet.
+\end{acks}
 
-\bibliographystyle{ipsjsort-e}
+\ifsigpro{\bibliographystyle{ipsjsort-e}}
+\ifsigplan{\bibliographystyle{ACM-Reference-Format}}
 \bibliography{paper}

### 2019-01-18 lib/magritte/builtins.rb
add mod builtin, and even/odd functions

@@ -115,6 +115,11 @@ module Magritte
       Status.normal
     end
 
+    builtin :mod, [:Number, :Number] do |base, n|
+      put Value::Number.new(n.value % base.value)
+      Status.normal
+    end
+
     builtin :exec, [], :any do |*a|
       Value::Vector.new([]).call(a)
       Status.normal

### 2019-01-18 lib/magritte/repl.rb
log infra

@@ -29,6 +29,7 @@ module Magritte
       end
 
       @env = Env.base.extend([@input], [@streamer])
+      @env.let('LOG', Value::Channel.new(@streamer))
     end
 
     def self.run(argv)

### 2019-01-18 lib/magritte/builtins.rb
add (null) builtin to return a null channel

@@ -236,6 +236,11 @@ module Magritte
       end
     end
 
+    # the null channel
+    builtin :null, [] do
+      put Value::Channel.new(Null.new)
+    end
+
     # Initialize environment with functions that can be defined in the language itself
     def self.load_lib(lib_name, source, env)
       # Transform the lib string into an ast

### 2019-01-18 lib/magritte/interpret.rb
better error messages

@@ -18,7 +18,7 @@ module Magritte
       end
 
       def visit_default(node)
-        raise "TODO #{node.inspect}"
+        error! "Cannot interpret as an expression: #{node.inspect}"
       end
 
       def visit_variable(node)

### 2019-01-18 lib/magritte/parser.rb

@@ -199,7 +199,7 @@ module Magritte
         return AST::VectorPattern[vec.elems.map { |e| parse_pattern(e) }, nil]
       end
 
-      error!(term, "unrecognized pattern")
+      error!(pattern, "unrecognized pattern")
     end
 
     def parse_term(term)

### 2019-01-18 lib/magritte/repl.rb
repl: don't add empty lines to history

@@ -60,7 +60,7 @@ module Magritte
     def process_line
       string = @mutex.synchronize { Readline.readline("; ", true) }
       raise IOError if string.nil?
-      return false if string =~ /\A\s*\z/m
+      (Readline::HISTORY.pop; return false) if string =~ /\A\s*\z/m
       status = evaluate(string)
       @mutex.synchronize { puts "% #{status.repr}" }
       !status.fail?
